---
title: "ADaM Concepts"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 85 minutes.**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background: what ADaM adds to SDTM | 25 min | Reading |
| Example 1: The two shapes — ADSL and BDS | 20 min | Worked example |
| Example 2: Baseline is an ADaM decision | 20 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Comprehension Check | 5 min | Self-test |

The first session of Part 3. It has no exercise set — it is the mental
model for [ADSL](adsl.qmd) and [ADAE / ADLB](adae-adlb.qmd), which build
the real analysis datasets. For self-paced study it fits one sitting.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown.
Where you see a "Run It Yourself" box, stop and run the code above it
before reading the explanation that follows. Run R from the project
root, as in earlier sessions.

This session reads the finished SDTM you built in Part 2 —
`data/sdtm/dm.csv` and `data/sdtm/lb.csv` — rather than the raw files.
That hand-off is the point where "what happened" becomes "what the
analysis needs."
:::

## When Do You Use This? {#sec-adam-concepts}

::: {.callout-tip}
A statistician asks for a dataset they can fit a model on with almost no
reshaping. A reviewer wants to see exactly how "change from baseline"
was computed — and to trace each number back to its source. You have
conformant SDTM and now have to turn it into something the analysis can
use directly.

That is the whole job of ADaM, and it raises one question this session
answers: **what does an analysis dataset need that a tabulation dataset
does not?**

Builds on all of Part 2 (the persisted SDTM in `data/sdtm/`), and
especially [Findings](sdtm-findings.qmd), where the authoritative
baseline was deliberately deferred to ADaM. This session is where that
deferral is paid off.
:::

## Learning Objectives

After completing this session you will be able to:

- Explain what "analysis-ready" means and why ADaM exists alongside SDTM
- Describe the two ADaM structures — ADSL and BDS — by what one row is
- Name the core BDS variables (PARAM, PARAMCD, AVAL) and say how a
  parameter differs from an SDTM test
- Explain the two levels of traceability and why ADaM must show its work
- Say why population and baseline flags are re-derived in ADaM rather
  than reused from SDTM

## Background: What ADaM Adds to SDTM

*Estimated time: ~25 minutes (reading)*

::: {.callout-important}
### The one thing to remember

**An ADaM dataset is analysis-ready and self-documenting: shaped so the
statistic is almost no programming away, and built so every derived
value traces back to the SDTM row it came from.**
:::

SDTM answered *what happened in the trial*. ADaM answers *what the
trial shows* — and to do that it does two things SDTM deliberately does
not.

First, it is **analysis-ready**. An ADaM dataset must be structured and
populated so that a statistical analysis runs "with minimal
programming" (ADaMIG v1.3, §2.1). The change-from-baseline column is
already there; the treatment a record belongs to is already there; the
population a subject is in is already flagged. The statistician should
be one procedure away from the result, not a week of data-wrangling
away. (SDTM, by contrast, is a *tabulation* format — faithful to what
was collected, not shaped for any particular analysis.)

Second, it is **traceable**. An ADaM dataset contains both source data
carried from SDTM and values derived within ADaM, and a record can be
observed, derived, or imputed (ADaMIG v1.3, §2.3.2). Because it mixes
the collected and the computed, it must show its work. The ADaMIG
requires traceability at two levels (ADaMIG v1.3, §2.2):

- **Metadata traceability** — a description of how each analysis
  variable was derived from its immediate predecessor. This is
  **required** for ADaM compliance.
- **Datapoint traceability** — direct links in the data to the specific
  predecessor record(s), for example through a `--SEQ` variable.
  Implemented when practical.

This is the other end of a requirement you already met in SDTM: a
conformant submission includes both SDTM and ADaM, and "any data
collected that will be submitted in an analysis (ADaM) dataset must be
traceable to a source in a tabulation (SDTM) dataset" (SDTMIG v3.4,
§2.5). The ADaMIG states the ADaM end (§2.2); the SDTMIG states the
SDTM end (§2.5). The path runs from an analysis value, through its
predecessors, back to SDTM, and ultimately to the data collection
instrument. This is why the persisted `data/sdtm/` from Part 2 is the
input here: ADaM builds *on* SDTM, traceably.

### The two structures

The ADaMIG defines exactly **two** standard data structures (ADaMIG
v1.3, §2.3), and telling them apart is the first practical skill:

- **ADSL — the Subject-Level Analysis Dataset.** One record per
  subject, regardless of trial design; there is exactly one ADSL per
  study (ADaMIG v1.3, §2.3.1). It holds the subject-level facts the
  analysis leans on: population flags, planned and actual treatment,
  demographics, randomization and stratification factors, and important
  dates. It is emphatically **not** the place for endpoints or anything
  that varies over time.
- **BDS — the Basic Data Structure.** One or more records per subject
  per analysis parameter per analysis timepoint (the timepoint is
  conditionally required) (ADaMIG v1.3, §2.3.2). Its two central
  variables are `AVAL` (the value being analysed) and `PARAM` (the
  description of that value). A BDS dataset can be derived from any
  SDTM class — Findings, Events, Interventions, Special-Purpose — from
  other ADaM datasets, or any combination.

A caution the guide makes explicitly (ADaMIG v1.3, §2.3.2): BDS *looks*
like the SDTM Findings class because it tends to be vertical, but it is
not the same thing. A BDS parameter can be a highly derived quantity
assembled from several SDTM domains at once — something no single
Findings record could hold.

There is a third ADaM structure, **OCCDS** (the Occurrence Data
Structure), used for adverse events. It is specified in a *separate*
CDISC document, not the ADaMIG (ADaMIG v1.3, §2.3), so this course
introduces it in [ADAE / ADLB](adae-adlb.qmd) and cites it there.

### What a parameter is

The heart of BDS is the parameter, and it is worth contrasting with the
SDTM test it resembles. In SDTM, `--TESTCD` works *with* qualifier
variables — specimen, position, method — that together pin down what
was measured. In ADaM, `PARAM` stands alone: it "has no accompanying
qualifier variables," and `PARAM` by itself fully describes what is in
`AVAL` (ADaMIG v1.3, §3.3.4.2). Everything that would have been a
qualifier is folded into the parameter's definition.

That is because a parameter is **created to meet an analysis need, not
because something was collected** (ADaMIG v1.3, §3.3.4.2). "HbA1c" is a
test; "HbA1c change from baseline, on-treatment" is a parameter. `PARAM`
(required, on every record, up to 200 characters) carries the full
description; `PARAMCD` (required, on every record) is its short code —
and `PARAMCD` has the same shape rule you already know from SDTM's
`--TESTCD`: at most 8 characters, letters/numbers/underscore only.
Slightly stricter, in fact: it must **start with a letter** (not an
underscore), because `PARAMCD` values double as valid ADaM variable
names (ADaMIG v1.3, §3.3.4.1). The value itself lives in `AVAL`
(numeric) or `AVALC` (character); at least one is present, usually only
one (ADaMIG v1.3, §3.3.4.2).

Note the consequence, and hold onto it: **ADaM parameters are not
controlled terminology.** `PARAMCD`/`PARAM` are defined by the sponsor
for the analysis, not drawn from a published codelist. There is no
`is_term()` check for a parameter, because a parameter is a design
decision, not a standard value. This is the opposite of the SDTM
discipline, and deliberately so.

### The change-from-baseline machinery, and re-derived flags

Analysis-ready means the change-from-baseline is already computed, and
BDS has standard variables for it (ADaMIG v1.3, §3.3.4, §3.3.8):
`ABLFL` flags the baseline record; `BASE` is the `AVAL` copied from the
record where `ABLFL = "Y"`; `CHG` is `AVAL − BASE`; `PCHG` is the
percent change. `ABLFL` is *conditionally* required — required whenever
`BASE` is present in the dataset (ADaMIG v1.3, §3.3.8).

And here is where a thread from Part 2 closes. Whenever population or
baseline flags are needed for analysis, **the ADaM variables must be
used** — ADaM re-derives them; it does not reuse SDTM's (ADaMIG v1.3,
§3.5). A baseline record identified in SDTM may not be the baseline
record identified in ADaM. This is the precise reason the [Findings
session](sdtm-findings.qmd) deferred the "authoritative baseline" from
SDTM's `LBLOBXFL` to ADaM: which value counts as baseline for the
analysis is an ADaM decision, and Example 2 shows the two flags landing
on different rows for the same subject.

## Example 1: The Two Shapes — ADSL and BDS

*Estimated time: ~20 minutes (worked example)*

Read the finished SDTM and shape a small preview of each ADaM structure
— not the full builds (those are [ADSL](adsl.qmd) and [ADAE /
ADLB](adae-adlb.qmd)), just enough to *see* the difference in what one
row is.

First, an ADSL-shaped preview: one row per subject, with a planned
treatment and a population flag:

```{r adsl-shape}
#| eval: true

dm <- read_csv("data/sdtm/dm.csv", show_col_types = FALSE,
               col_types = cols(.default = col_character()))

adsl_preview <- dm |>
  transmute(
    USUBJID,
    TRT01P = ARM,   # planned treatment, period 1
    SAFFL  = "Y"    # safety population: every randomised GLPX subject was dosed
  )

c(rows = nrow(adsl_preview), subjects = n_distinct(adsl_preview$USUBJID))

head(adsl_preview, 3)
```

Now a BDS-shaped preview for one parameter, HbA1c:

```{r bds-shape}
#| eval: true

lb <- read_csv("data/sdtm/lb.csv", show_col_types = FALSE,
               col_types = cols(.default = col_character()))

adlb_hba1c <- lb |>
  filter(LBTESTCD == "HBA1C") |>
  transmute(
    USUBJID,
    PARAMCD = "HBA1C",
    PARAM   = "Hemoglobin A1C (%)",
    AVISIT  = VISIT,
    AVAL    = as.numeric(LBSTRESN),
    ADT     = LBDTC,
    ADY     = as.integer(LBDY)
  )

adlb_hba1c |> filter(USUBJID == "GLPX001-101-026")
```

::: {.callout-tip}
### Run It Yourself

Run both chunks. The ADSL preview has **400 rows for 400 subjects** —
one each. The BDS preview returns **8 rows for the one subject** shown —
one per visit. That is the entire structural distinction: ADSL is one
row per subject; BDS is one row per subject per parameter per timepoint.
:::

::: {.callout-note}
### Reading the Output Line by Line

| What you see | What it means |
|---|---|
| ADSL: 400 rows = 400 subjects | Subject-level. `TRT01P` (planned treatment) and `SAFFL` (safety flag) are facts *about the subject*, constant across the study. |
| BDS: 8 rows, one subject | Parameter-and-timepoint level. The same subject appears eight times because HbA1c was measured at eight visits. |
| `PARAMCD` / `PARAM` | The parameter, standing alone — no qualifier columns. `PARAM` is the human-readable description; `PARAMCD` the 8-character code. |
| `AVAL` | The analysis value — here copied from the standardised SDTM result `LBSTRESN`. In a real parameter it might be highly derived. |
| `ADT` / `ADY` / `AVISIT` | Analysis date, analysis day, analysis visit — the ADaM timing variables. `AVISIT` is *derived*; it need not match SDTM `VISIT` one-to-one (ADaMIG v1.3, §3.3.3). |

These are previews: `SAFFL` here is a simplification (every GLPX
subject was dosed, so all are `"Y"`), and the real derivations — from
EX, from the population definitions — are the work of the [ADSL](adsl.qmd)
and [ADLB](adae-adlb.qmd) sessions. The point now is only the *shape*.
:::

## Example 2: Baseline Is an ADaM Decision

*Estimated time: ~20 minutes (worked example)*

In [Findings](sdtm-findings.qmd) you derived `LBLOBXFL` — SDTM's
operational baseline flag, the last value before first dose. The
analysis, though, may define baseline differently, and the ADaMIG is
explicit that the ADaM baseline can differ from the SDTM one (ADaMIG
v1.3, §3.5). Here the analysis defines baseline as the **Baseline
visit** value. Build `ABLFL`, `BASE` and `CHG`, and carry the SDTM flag
alongside so the two can be compared:

```{r baseline}
#| eval: true

adlb <- lb |>
  filter(LBTESTCD == "HBA1C") |>
  transmute(USUBJID, AVISIT = VISIT, AVAL = as.numeric(LBSTRESN), LBLOBXFL) |>
  mutate(ABLFL = if_else(AVISIT == "BASELINE", "Y", NA_character_)) |>
  group_by(USUBJID) |>
  mutate(
    BASE = AVAL[match("Y", ABLFL)],   # AVAL on the ABLFL = "Y" record
    CHG  = round(AVAL - BASE, 1)
  ) |>
  ungroup()

adlb |>
  filter(USUBJID == "GLPX001-101-026") |>
  select(AVISIT, AVAL, LBLOBXFL, ABLFL, BASE, CHG)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk. Look at the `LBLOBXFL` and `ABLFL` columns side by side
for this subject: the SDTM flag sits on **screening**, the ADaM flag on
**baseline** — two different rows. `BASE` is the baseline-visit value
(8.9), and `CHG` is each visit's value minus that. The two flags
disagreeing is not an error; it is the ADaMIG's point made concrete.
:::

::: {.callout-note}
### Reading the Output Line by Line

| Column | What it means |
|---|---|
| `LBLOBXFL = "Y"` on SCREENING | The SDTM baseline flag: the last value *before first dose*, computed in the Findings session. |
| `ABLFL = "Y"` on BASELINE | The ADaM analysis baseline: the value the *analysis* treats as baseline. A different row. |
| `BASE` (8.9 throughout) | The `AVAL` from the `ABLFL = "Y"` record, copied onto every row so change can be computed (ADaMIG v1.3, §3.3.4). |
| `CHG` | `AVAL − BASE`, the analysis-ready change from baseline (ADaMIG v1.3, §3.3.4). No further computation needed downstream. |

Two flags, two rows, both correct — because they answer different
questions. `LBLOBXFL` asks "what was the last pre-treatment value?";
`ABLFL` asks "what does *this analysis* call baseline?" Whenever a
baseline flag is needed for analysis, the ADaM variable must be used
(ADaMIG v1.3, §3.5), and that is why the flag is re-derived here rather
than copied from SDTM.

One more, for later: our baseline is an *observed* value (the
Baseline-visit measurement), so no derivation type is needed. If the
analysis baseline were itself *derived* — say, an average of two
pre-dose values — then `DTYPE` would have to be populated to record how
(ADaMIG v1.3, §3.3.8). A derived baseline is still an ADaM decision,
just a more elaborate one; the traceability requirement is the same.
:::

## What Can Go Wrong

*Estimated time: ~10 minutes (reading)*

**Treating ADaM as "SDTM with extra columns"**

ADaM is a different shape for a different purpose: analysis-ready, both
source and derived, and traceable. Bolting analysis columns onto an
SDTM domain gives you neither the structure nor the traceability the
standard requires. The two structures (ADSL, BDS) exist precisely so
the analysis and its provenance are predictable.

**Reusing SDTM's flags for analysis**

The SDTM population and baseline flags are not the analysis flags.
Whenever a flag is needed for analysis, the ADaM variable must be used
and re-derived (ADaMIG v1.3, §3.5). Using `LBLOBXFL` as if it were the
analysis baseline skips the decision ADaM exists to record — and, as
Example 2 showed, may point at the wrong row.

**Putting time-varying data in ADSL**

ADSL is one row per subject; endpoints and anything that changes over
time belong in a BDS dataset (ADaMIG v1.3, §2.3.1). A week-26 HbA1c is
not an ADSL variable.

**Expecting PARAM to have qualifiers**

`PARAM` stands alone — no specimen, position or method columns beside it
(ADaMIG v1.3, §3.3.4.2). If you find yourself wanting a qualifier
variable, the distinction it captures belongs *inside* the parameter
definition (a new `PARAM`), not in a separate column.

**Breaking traceability**

Dropping the SDTM inputs a derivation used, or deriving a value without
recording how, breaks the traceability that is required for compliance
(ADaMIG v1.3, §2.2). Every derived value must be explainable back to its
predecessor.

::: {.callout-warning}
### Common misinterpretations

**"ADaM is just a reformat of SDTM"**
ADaM adds derived analysis values, a purpose-built structure, and
required traceability. A reformat has none of those. The change-from-
baseline in Example 2 exists in ADaM and nowhere in SDTM.

**"PARAMCD is controlled terminology like LBTESTCD"**
No — `PARAMCD`/`PARAM` are sponsor-defined for the analysis (ADaMIG
v1.3, §3.3.4). They share `--TESTCD`'s 8-character *shape*, but there is
no codelist to check them against, because a parameter is a design
decision.

**"The SDTM baseline flag and the ADaM baseline flag should match"**
They may, but they need not (ADaMIG v1.3, §3.5). They answer different
questions, and Example 2 shows them on different rows for the same
subject — legitimately.

**"Analysis-ready means the data is finished"**
It means the dataset is *shaped* so analysis needs minimal programming
(ADaMIG v1.3, §2.1). The analysis decisions — baseline, populations,
parameters — are still yours to make and to document; ADaM is where you
record them, not a substitute for making them.
:::

## Comprehension Check

*Estimated time: ~5 minutes*

1. In one sentence each, what does "analysis-ready" mean, and what does
   "traceable" mean, for an ADaM dataset?
2. A dataset has one row per subject per lab test per visit, with `AVAL`
   and `PARAM`. Which ADaM structure is it, and how do you know?
3. Why does `PARAM` need no qualifier variables, when SDTM's `--TESTCD`
   works alongside several?
4. For subject `101-026`, `LBLOBXFL` flags the screening record and
   `ABLFL` flags the baseline record. Is this a bug? Explain.
5. Why is there no `is_term()` check for a `PARAMCD` value, when there is
   one for an `LBTESTCD` value?

::: {.callout-note collapse="true"}
### Answers

1. **Analysis-ready**: the dataset is structured and populated so a
   statistical analysis runs with minimal programming — the analysis
   values (e.g. change from baseline) and selection variables are
   already present (ADaMIG v1.3, §2.1). **Traceable**: every analysis
   value can be followed back through its derivation to its SDTM source
   and ultimately the collection instrument, with the derivation
   described in metadata (ADaMIG v1.3, §2.2).

2. BDS — the Basic Data Structure. One or more records per subject per
   analysis parameter per analysis timepoint, built around `AVAL` and
   `PARAM` (ADaMIG v1.3, §2.3.2). ADSL, by contrast, is one row per
   subject (ADaMIG v1.3, §2.3.1), so a per-test-per-visit structure
   cannot be ADSL.

3. Because a `PARAM` is created to fully describe its `AVAL` for a
   specific analysis need, folding every distinguishing detail into the
   parameter definition itself; it "has no accompanying qualifier
   variables" (ADaMIG v1.3, §3.3.4.2). SDTM `--TESTCD` describes a
   *collected* result and relies on qualifiers to pin down the
   conditions of collection.

4. Not a bug. `LBLOBXFL` is SDTM's operational flag (last value before
   first dose); `ABLFL` is the analysis baseline (here the Baseline
   visit). The ADaMIG explicitly allows the ADaM baseline record to
   differ from the SDTM one, and requires the ADaM variable to be used
   for analysis (ADaMIG v1.3, §3.5). They answer different questions and
   correctly land on different rows.

5. Because `PARAMCD`/`PARAM` are **sponsor-defined for the analysis**,
   not drawn from controlled terminology (ADaMIG v1.3, §3.3.4). A
   parameter is a design decision, so there is no codelist to validate
   it against — unlike `LBTESTCD`, which must be a published CT value.
   `PARAMCD` still obeys a *shape* rule (≤8 characters, starts with a
   letter), but shape is not membership.
:::
