---
title: "ADLB"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(sdtm.terminology)
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 105 minutes.**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background: one row per subject, parameter, timepoint | 25 min | Reading |
| Example 1: Parameters, values and analysis timing | 25 min | Worked example |
| Example 2: Baseline, change, and the flag that matters | 25 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Discussion: Which Baseline Would You Defend? | 10 min | Discussion |
| Exercises + Comprehension Check | 5 min | Practice / Self-test |

The last session of Part 3, and the one the baseline thread has been
building towards since Part 2. For self-paced study, split after
Example 1.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown. Run R
from the project root, as in earlier sessions.

This session reads `data/sdtm/lb.csv` and `data/adam/adsl.csv`.
:::

## When Do You Use This? {#sec-adlb}

::: {.callout-tip}
The primary endpoint of GLPX-1 is a *change* from baseline in HbA1c.
Before anyone can compute it, something has to decide which value
baseline is. A safety review needs every lab result classified against
its reference range at every visit. A reviewer needs to see the measured
value and the analysis value side by side, and to know which record the
change was measured from.

All three need the same thing: a dataset where each row is one
measurement, described well enough to stand on its own, with the
baseline it is compared against carried alongside it. That raises the
question this session answers: **when the whole trial is measured as
change from baseline, who decides which value baseline is?**

Builds on [ADaM Concepts](adam-concepts.qmd), [ADSL](adsl.qmd), and
[Findings](sdtm-findings.qmd), where `LBLOBXFL` was deliberately left
unfinished.
:::

## Learning Objectives

After completing this session you will be able to:

- Build a BDS dataset with one row per subject, per parameter, per
  timepoint, and say why that is the structure
- Define `PARAM` and `PARAMCD` correctly, and explain why neither has a
  codelist to check against
- Derive `AVAL`, `ADT`, `ADY` and `AVISIT`, and know which of them the
  analysis is free to define
- Derive `ABLFL` from a stated baseline definition, and `BASE`, `CHG`
  and `PCHG` from it
- Explain why ADaM's baseline flag and SDTM's may disagree, and measure
  what that disagreement costs

## Background: One Row per Subject, Parameter, Timepoint

**Reading task, about 25 minutes.** BDS is the structure most analysis
datasets use. Read it properly; the rest of Course B assumes it.

*Estimated time: ~25 minutes (reading)*

::: {.callout-important}
### The one thing to remember

**ADLB is one row per subject, per parameter, per timepoint: `PARAM`
says what was measured, `AVAL` says what it was, and `ABLFL` nominates
the one row the analysis calls baseline, so that every change in the
trial is measured from a point the SAP chose rather than one the lab
happened to flag.**
:::

The Basic Data Structure is the workhorse of ADaM, and its shape is
stated precisely: a BDS dataset contains "1 or more records per subject,
per analysis parameter, per analysis timepoint" (ADaMIG v1.3, §2.3.2).
Analysis timepoint is conditionally required; where an analysis has no
timepoint, the structure is one or more records per subject per
parameter.

That is a vertical shape, and it invites a comparison the guide
explicitly warns against:

> ADaM dataset structures do not have counterparts in SDTM. Because the
> BDS tends toward a vertical design, some might perceive it as similar
> to the SDTM Findings class. However, BDS datasets may be derived from
> Findings, Events, Interventions and Special-purpose SDTM domains,
> other ADaM datasets, or any combination thereof (ADaMIG v1.3, §2.3.2).

Our ADLB does happen to come from a Findings domain, which makes the
resemblance easy to over-read. Resist it. BDS is defined by what it is
*for*, not by which SDTM class it happened to come from.

One more caution before the variables, because it is easy to get wrong
in the opposite direction. ADSL is a source of subject-level variables
for BDS, but "this does not mean that every ADSL variable should be
included in the BDS dataset" (ADaMIG v1.3, §2.3.2). Carry what the
analysis needs. A BDS dataset with all 28 ADSL columns bolted on is not
more traceable, only wider.

### Parameters: what was measured

In SDTM, a lab result needs `LBTESTCD`, `LBTEST` and `LBSTRESU` before
you know what it is. In BDS, one variable does that work. `PARAM` is
Required, and it "must include all descriptive and qualifying
information relevant to the analysis purpose of the parameter" and
"should be sufficient to describe unambiguously the contents of `AVAL`
and/or `AVALC`" (ADaMIG v1.3, §3.3.4.1). Units are named as a candidate
qualifier, which is why our parameters read `Hemoglobin A1C (%)` rather
than `Hemoglobin A1C` with the unit parked in a neighbouring column.

The guide is explicit about the contrast: "in contrast to SDTM
`--TEST`, no additional variable is needed to further qualify `PARAM`"
(ADaMIG v1.3, §3.3.4.1). `PARAM` is limited to 200 characters, and the
guide suggests keeping it to 40 if it will be used as a variable label
in a transposed dataset.

`PARAMCD` is its short form, also Required, and it carries constraints
that look fussy until you know their purpose:

> The values of `PARAMCD` must be no more than 8 characters in length,
> start with a letter (not underscore), and be comprised only of letters
> (A-Z), underscore (\_), and numerals (0-9) (ADaMIG v1.3, §3.3.4.1).

The reason is given in the next sentence: those are exactly the
constraints that let a BDS dataset be transposed so that values of
`PARAMCD` become valid variable names. There must be a one-to-one
relationship between `PARAM` and `PARAMCD` within a dataset, and both
must be present and populated on every record.

::: {.callout-warning}
### There is no codelist for `PARAMCD`

You met this with `AGEGR1` in [ADSL](adsl.qmd) and `ASEV` in
[ADAE](adae.qmd), and it is worth stating a third time because it is the
single most common misunderstanding about ADaM: **analysis parameters
are sponsor-defined.** There is no CDISC codelist of `PARAMCD` values to
validate against. What governs them is the character constraint above
and the requirement that `PARAM` describe `AVAL` unambiguously.

Our `PARAMCD` values happen to read `HBA1C`, `GLUC`, `ALT` and `CREAT`,
which are also `LBTESTCD` submission values. That is a deliberate
convenience for readability, not a requirement, and it is worth not
being misled by it: nothing obliges an analysis parameter to be named
after the SDTM test it came from, and a highly derived parameter has no
SDTM test to be named after at all.
:::

`AVAL` (numeric) and `AVALC` (character) hold the value itself. Both are
Conditional, and the condition is mutual: "either `AVAL` or `AVALC` must
be present in the dataset", each being required if the other is absent
(ADaMIG v1.3, §3.3.4.1). Where both are populated within a parameter
there must be a one-to-one relationship between them, and `AVALC`
"should not be used to categorize the values of `AVAL`". Categorising is
what `AVALCATy` is for.

### Timing: the anchor moves again

`ADT` is "the date associated with `AVAL` and/or `AVALC`", and `ADY` is
its relative day. If you read [ADAE](adae.qmd), the definition of `ADY`
will look familiar:

> The number of days from an anchor date (**not necessarily
> `DM.RFSTDTC`**) to `ADT` (ADaMIG v1.3, §3.3.3).

That is the same parenthesis OCCDS used for `ASTDY`, and it makes the
same point. The study-day arithmetic is fixed and has no day zero. The
day it counts *from* is the analysis's choice, and must be documented.
We anchor to `ADSL.TRTSDT`, first dose.

There is also a structural requirement hiding in that table: "if a
dataset contains more than one record per parameter per subject, then an
SDTM or ADaM relative timing variable must be present (`ADY` would meet
this requirement)" (ADaMIG v1.3, §3.3.3). ADLB has eight records per
parameter per subject, so `ADY` is not optional decoration.

`AVISIT` is Conditional, "required if an analysis is done by nominal,
assigned or analysis visit", and it comes with three precisions that
matter more than they look:

- "`AVISIT` is a derived field and **does not have to map to `VISIT`
  from the SDTM**"
- "`AVISIT` represents the analysis visit of the record, but **it does
  not mean that the record was analyzed**"
- "Values of `AVISIT` are producer-defined" (all ADaMIG v1.3, §3.3.3)

GLPX-1 has a clean nominal visit schedule, so our `AVISIT` copies
`VISIT` unchanged. That is a defensible choice for this trial, not a
general rule. In studies where subjects attend outside the protocol
window, `AVISIT` is where windowing decisions get recorded, and it can
differ from `VISIT` on most rows.

### Ranges

The reference range you built in [Findings](sdtm-findings.qmd) has ADaM
counterparts: `ANRLO` and `ANRHI` for the limits, `ANRIND` for the
indicator, and `BNRIND` for the indicator on the baseline record. All
four are Permissible (ADaMIG v1.3, §3.3.7).

Note how they are described. `ANRIND` "indicates where `AVAL` or `AVALC`
falls with respect to the normal reference range for analysis; **may be
based on SDTM `--NRIND` or an imputed or assigned value**", and
`BNRIND` is "`ANRIND` of the baseline record identified by `ABLFL`"
(ADaMIG v1.3, §3.3.7). The pattern is the one you have now seen for
population flags, for baseline, and for treatment emergence: SDTM's
version is available and often correct, and the analysis variable is
still the analysis's own to derive.

### Baseline, and the decision the SAP has to make

Now the centre of the session.

`ABLFL` "identifies the baseline record for each subject, parameter, and
baseline type (`BASETYPE`) combination", and its conditionality is
precise: **"`ABLFL` is required if `BASE` is present in the dataset"**
(ADaMIG v1.3, §3.3.8). `BASE` in turn "contains the value of `AVAL`
copied from a record within the parameter on which `ABLFL = "Y"`", and
the two are locked together by an integrity rule: if `BASE` is populated
for a parameter and is non-null for a subject, "then there must be a
record flagged by `ABLFL` for that subject and parameter" (ADaMIG v1.3,
§3.3.4.1).

From there the derived columns follow mechanically. `CHG` is "equal to
`AVAL-BASE`" and `PCHG` is "equal to `((AVAL-BASE)/BASE)*100`", both
Permissible, and both "should be populated for all post-baseline records
of that `PARAM` regardless of whether that record is used for analysis"
(ADaMIG v1.3, §3.3.4.1).

What the guide does *not* do is tell you what baseline is. It gives you
the machinery to record a decision, and leaves the decision to the
statistical analysis plan. Two further variables mark the edges of that
freedom:

- **`BASETYPE`** is "producer-defined text describing the definition of
  baseline relevant to the value of `BASE` on the current record", and is
  "required when there are multiple ways that baseline is defined"
  (ADaMIG v1.3, §3.3.4.1). GLPX-1 has one definition, so `BASETYPE` is
  not required here and is not built. A study with a screening baseline,
  a double-blind baseline and an open-label-extension baseline needs it
  on every affected row (see ADaMIG v1.3, §4.2.1.6, Rule 6).
- **`DTYPE`** "is used to denote, and must be populated, when the value
  of `AVAL` or `AVALC` has been imputed or derived differently than the
  other analysis values within the parameter" (ADaMIG v1.3, §3.3.5).
  A baseline record "may be derived (e.g., it may be an average), in
  which case `DTYPE` must also be populated" (ADaMIG v1.3, §3.3.8).
  Every `AVAL` in our ADLB is an observed lab result, so `DTYPE` is null
  throughout. Null `DTYPE` is a statement, not an omission: it says this
  row is what was measured.

::: {.callout-important}
### GLPX-1's baseline definition, and why it is not "the BASELINE visit"

The obvious definition is "the record at the visit named BASELINE." For
this trial that would be wrong, and the data says so.

GLPX-1's baseline visit is scheduled in a window that, for many
subjects, falls *after* the first dose. Flagging the BASELINE visit
record would make **958 of 1600** subject-parameter baselines post-dose:
values measured after the subject had already started study drug, used
as the anchor for measuring the drug's effect.

So the SAP for this course defines baseline the way real protocols
usually word it: **the last non-missing value on or before the first
dose of study drug.** Example 1 checks that this produces zero post-dose
baselines, and Example 2 shows what it costs and what it buys.

The general lesson is the one worth carrying: a baseline definition is a
claim about which measurement is untreated. Check it against the dosing
dates rather than trusting the visit label.
:::

## Example 1: Parameters, Values and Analysis Timing

**Worked example, about 25 minutes.** Two stops. This is where one
column stops meaning one thing.

*Estimated time: ~25 minutes (worked example)*

Both inputs, as in ADAE: the measurements from SDTM, the subject-level
context from ADSL.

```{r read-inputs}
#| eval: true

lb <- read_csv("data/sdtm/lb.csv", show_col_types = FALSE,
               col_types = cols(.default = col_character()))

adsl <- read_csv("data/adam/adsl.csv", show_col_types = FALSE)

c(lb_rows = nrow(lb), subjects = n_distinct(lb$USUBJID),
  tests = n_distinct(lb$LBTESTCD))
```

Parameters first, because in BDS the parameter is the organising idea.
Define them in one place, with the units inside `PARAM`:

```{r params}
#| eval: true

adlb_params <- tribble(
  ~LBTESTCD, ~PARAMCD, ~PARAM,                           ~PARAMN,
  "HBA1C",   "HBA1C",  "Hemoglobin A1C (%)",                   1L,
  "GLUC",    "GLUC",   "Glucose (mmol/L)",                     2L,
  "ALT",     "ALT",    "Alanine Aminotransferase (U/L)",       3L,
  "CREAT",   "CREAT",  "Creatinine (umol/L)",                  4L
)

# PARAMCD constraints (ADaMIG v1.3, §3.3.4.1): <= 8 chars, starts with a
# letter, letters/underscore/digits only.
adlb_params |>
  mutate(chars = nchar(PARAMCD),
         ok    = chars <= 8 & grepl("^[A-Za-z][A-Za-z0-9_]*$", PARAMCD))
```

Now the spine: values, timing, ranges, treatment and traceability.

```{r adlb-spine}
#| eval: true

adlb_spine <- lb |>
  inner_join(adlb_params, by = "LBTESTCD") |>
  left_join(
    adsl |> select(USUBJID, SUBJID, SITEID, TRT01P, TRT01A, TRTSDT, SAFFL),
    by = "USUBJID"
  ) |>
  mutate(
    LBSEQ = as.integer(LBSEQ),
    AVAL  = as.numeric(LBSTRESN),

    AVISIT  = VISIT,                       # producer-defined; VISIT is clean here
    AVISITN = as.integer(VISITNUM),

    ADT = as.Date(LBDTC),
    ADY = if_else(ADT >= TRTSDT,           # no day 0, anchored to ADSL.TRTSDT
                  as.integer(ADT - TRTSDT) + 1L,
                  as.integer(ADT - TRTSDT)),

    ANRLO  = as.numeric(LBSTNRLO),         # analysis versions of the range
    ANRHI  = as.numeric(LBSTNRHI),
    ANRIND = LBNRIND,

    TRTP = TRT01P,                         # record-level treatment
    TRTA = TRT01A,

    SRCDOM = "LB",                         # datapoint traceability
    SRCVAR = "LBSTRESN",
    SRCSEQ = LBSEQ,

    DTYPE = NA_character_                  # every AVAL is observed
  )

adlb_spine |>
  filter(USUBJID == "GLPX1-101-010", PARAMCD == "HBA1C") |>
  select(PARAM, AVISIT, ADT, ADY, AVAL, ANRLO, ANRHI, ANRIND)
```

::: {.callout-tip}
### Run It Yourself

**Your turn, about 5 minutes.** Before running, predict how many rows
one subject and one parameter should produce, and what distinguishes
them.

Run both chunks. Eight rows for one subject and one parameter, which is
the BDS shape in miniature: same subject, same `PARAM`, one row per
timepoint.

Look at `ADY`. Screening is negative at -16, and the sequence steps up
to 181 by week 26 without ever passing through zero. That is the study-day rule
from [The DM Domain](sdtm-dm.qmd), now anchored to first dose rather
than to `RFSTDTC`.
:::

### Traceability, and a contrast with ADAE

`SRCDOM`, `SRCVAR` and `SRCSEQ` record where each analysis value came
from (ADaMIG v1.3, §3.3.9). It is worth pausing on `SRCVAR`, because
[ADAE](adae.qmd) explicitly did not have it: OCCDS excludes `SRCVAR`
because it names the variable that derives `AVAL` or `AVALC`, and OCCDS
has no `AVAL` (OCCDS v1.1, §3.2.2).

Here there is an `AVAL`, so naming its source variable is meaningful:
every row records that its value came from `LB.LBSTRESN`, at the
sequence number in `SRCSEQ`. Any number in a table built from this
dataset can be walked back to one row of one SDTM domain.

### Check the shape

```{r shape-check}
#| eval: true

c(rows            = nrow(adlb_spine),
  subjects        = n_distinct(adlb_spine$USUBJID),
  parameters      = n_distinct(adlb_spine$PARAMCD),
  one_row_per_key = nrow(distinct(adlb_spine, USUBJID, PARAMCD, AVISITN)) ==
                    nrow(adlb_spine))
```

12,388 rows, one per subject per parameter per visit, with no
duplicates. That last check matters for a reason worth naming.

::: {.callout-note}
### Why `ANL01FL` is absent

`ANLzzFL` is a conditionally required flag, "to be used in addition to
other selection variables when the other selection variables in
combination are insufficient to identify the exact set of records used
for one or more analyses" (ADaMIG v1.3, §3.3.8).

The check above is exactly that condition, tested. `USUBJID` plus
`PARAMCD` plus `AVISIT` already identify one record, so nothing is
ambiguous and no analysis flag is needed. Added to a dataset like this,
`ANL01FL` would be noise.

The condition fails as soon as a subject has two records for the same
parameter and visit, which happens routinely: an unscheduled repeat, a
record windowed into the same visit, a derived row sitting beside the
observed one. Then something has to say which record the analysis used,
and `ANLzzFL` is that something.
:::

### Reference ranges and controlled terminology

```{r ct-check}
#| eval: true

nrind_code <- (ct("list") |> filter(term == "NRIND"))$clst_code[1]

adlb_spine |>
  count(PARAMCD, ANRIND) |>
  pivot_wider(names_from = ANRIND, values_from = n, values_fill = 0L)

c(codelist  = nrind_code,
  all_valid = all(is_term(unique(adlb_spine$ANRIND), clst_code = nrind_code)))
```

::: {.callout-tip}
### Run It Yourself

**Your turn, about 4 minutes.** A passing controlled-terminology check.
Read which codelist it passed against and satisfy yourself that is the
right one.

`ANRIND` validates against `NRIND` (`C78736`), whose members are
`ABNORMAL`, `HIGH`, `LOW` and `NORMAL`.

Then read the counts, because they are a reminder of what trial this is.
HbA1c is `HIGH` on 3,080 of 3,097 records and `NORMAL` on 17; glucose is
`HIGH` on 3,084. In a type 2 diabetes trial that is not a data problem,
it is the disease. ALT and creatinine, which are safety labs rather than
efficacy ones, sit mostly `NORMAL`, as you would hope.

A reference-range indicator answers "is this value inside the normal
range." It does not answer "is this value concerning," and in a
population selected for an abnormal lab it will say `HIGH` almost
everywhere.
:::

## Example 2: Baseline, Change, and the Flag That Matters

**Worked example, about 25 minutes.** The most consequential example in
the course. Two defensible baseline definitions give two different
headline results, and you have to choose one.

*Estimated time: ~25 minutes (worked example)*

Now derive baseline from the SAP's definition: the last non-missing
value on or before first dose. Under the no-day-zero rule, `ADY == 1` is
the dosing day itself, so "on or before" is `ADY <= 1`.

```{r baseline}
#| eval: true

adlb <- adlb_spine |>
  arrange(USUBJID, PARAMN, AVISITN) |>
  group_by(USUBJID, PARAMCD) |>
  mutate(
    bl_ady = if (any(ADY <= 1L)) max(ADY[ADY <= 1L]) else NA_integer_,
    ABLFL  = if_else(!is.na(bl_ady) & ADY == bl_ady, "Y", NA_character_),
    BASE   = AVAL[match("Y", ABLFL)],
    BNRIND = ANRIND[match("Y", ABLFL)],
    CHG    = if_else(ADY >= bl_ady, AVAL - BASE, NA_real_),
    PCHG   = if_else(ADY >= bl_ady & BASE != 0,
                     100 * (AVAL - BASE) / BASE, NA_real_)
  ) |>
  ungroup() |>
  select(-bl_ady)

# Which visit does the baseline land on, and is any of it post-dose?
adlb |> filter(ABLFL == "Y") |> count(AVISIT, post_dose = ADY > 1)
```

::: {.callout-tip}
### Run It Yourself

**Your turn, about 6 minutes.** Two things to compare, and they
disagree. Work out which record each rule is picking before you read
the explanation.

Two things to read here.

First, **no baseline is post-dose**. That was the point of defining
baseline by dosing date rather than by visit label, and the check
confirms the definition did what it claimed.

Second, the baseline is not at the same visit for everyone: 642
subject-parameter pairs take it from the BASELINE visit and 958 from
SCREENING. That is not an inconsistency. It is one rule, applied to
subjects whose visits happened on different days relative to their own
first dose, which is what a real schedule looks like.
:::

Now the conformance rules the ADaMIG states for these variables:

```{r baseline-conformance}
#| eval: true

bl <- adlb |> filter(ABLFL == "Y")

tibble(
  check = c("exactly one ABLFL per subject and parameter (3.3.8)",
            "BASE non-null implies an ABLFL record exists (3.3.4.1)",
            "CHG is 0 on the baseline record (sponsor convention)",
            "DTYPE null: every AVAL observed, none derived (3.3.5)"),
  pass = c(
    all(count(bl, USUBJID, PARAMCD)$n == 1),
    all(!is.na(adlb$BASE)),
    all(bl$CHG == 0),
    all(is.na(adlb$DTYPE))
  )
)
```

### The flag that matters

Here is the payoff of a thread that started in [The DM
Domain](sdtm-dm.qmd) and was deliberately left open in
[Findings](sdtm-findings.qmd). SDTM has its own baseline flag,
`LBLOBXFL`. ADaM has `ABLFL`. Compare them on real data:

```{r flags-compare}
#| eval: true

sdtm_baseline <- lb |>
  filter(LBLOBXFL == "Y") |>
  transmute(USUBJID, PARAMCD = LBTESTCD, sdtm_visit = VISIT,
            BASE_sdtm = as.numeric(LBSTRESN))

comparison <- bl |>
  select(USUBJID, PARAMCD, adam_visit = AVISIT, adam_ady = ADY, BASE) |>
  inner_join(sdtm_baseline, by = c("USUBJID", "PARAMCD"))

count(comparison, adam_visit, sdtm_visit)

comparison |>
  filter(adam_visit != sdtm_visit) |>
  count(adam_ady, name = "disagreements")
```

::: {.callout-important}
### The two flags disagree on 342 of 1,600 pairs, for exactly one reason

Read the second table. **Every single disagreement is at `ADY == 1`**,
and that is the whole story.

SDTM defines `--LOBXFL` as the last non-missing value **strictly prior
to** `RFXSTDTC` (SDTMIG v3.4, §4.5.9). The SAP defines baseline as the
last value **on or before** the first dose. Those differ on exactly one
kind of record: a lab drawn on the dosing day itself, before the dose
was given.

Clinically that draw is unambiguously a baseline measurement, since the
subject had not yet received study drug. But it is not *prior to* first
exposure by the date comparison SDTM uses, so `LBLOBXFL` skips it and
reaches back to screening instead.

Neither flag is wrong. They answer different questions. `LBLOBXFL` is
operationally derived and describes the data as collected; `ABLFL`
records an analysis decision. Which is why the ADaMIG is blunt that only
the ADaM variable `ABLFL` is used to signify baseline for analysis.
:::

Seeing it on one subject makes the mechanism concrete. Subject
`101-010`, whom you met in [Findings](sdtm-findings.qmd) as the
duplicate-record case:

```{r one-subject}
#| eval: true

adlb |>
  filter(USUBJID == "GLPX1-101-010", PARAMCD == "HBA1C") |>
  select(AVISIT, ADY, AVAL, ABLFL, BASE, CHG, PCHG) |>
  head(4)

lb |>
  filter(USUBJID == "GLPX1-101-010", LBTESTCD == "HBA1C",
         VISIT %in% c("SCREENING", "BASELINE")) |>
  select(VISIT, LBDTC, LBSTRESN, LBLOBXFL)
```

::: {.callout-note}
### Reading the Output Line by Line

| What you see | What it means |
|---|---|
| BASELINE at `ADY = 1`, `AVAL = 8.6`, `ABLFL = "Y"` | The lab was drawn on the day of first dose, before dosing. The SAP counts it, so it is this subject's baseline. |
| SCREENING at `ADY = -16`, `AVAL = 8.2`, no flag | Sixteen days before dosing. A real measurement, but not the most recent one on or before first dose. |
| In SDTM, `LBLOBXFL = "Y"` on **SCREENING** | Strictly prior to first exposure, so SDTM reaches back past the dosing-day draw. |
| `BASE = 8.6`, not 8.2 | The analysis uses 8.6. Had it used SDTM's flag, every `CHG` for this subject would be 0.4 larger. |
| `CHG = 0` and `PCHG = 0` on the baseline row | A sponsor convention. The ADaMIG leaves the baseline and pre-baseline values of `CHG` to producer choice (§3.3.4.1), so this is documented, not derived from a rule. |
| `CHG` null at SCREENING | A change from baseline is not defined before baseline. |
:::

### What the choice costs

A 0.4 difference for one subject is easy to wave away. Measure it on the
primary endpoint instead:

```{r headline-impact}
#| eval: true

week26 <- adlb |>
  filter(PARAMCD == "HBA1C", AVISIT == "WEEK 26") |>
  left_join(sdtm_baseline |> select(USUBJID, PARAMCD, BASE_sdtm),
            by = c("USUBJID", "PARAMCD")) |>
  mutate(CHG_sdtm = AVAL - BASE_sdtm)

result <- week26 |>
  group_by(TRTA) |>
  summarise(n = n(),
            mean_CHG_adam = round(mean(CHG), 3),
            mean_CHG_sdtm = round(mean(CHG_sdtm), 3),
            .groups = "drop")

result

c(difference_adam = round(result$mean_CHG_adam[1] - result$mean_CHG_adam[2], 3),
  difference_sdtm = round(result$mean_CHG_sdtm[1] - result$mean_CHG_sdtm[2], 3))
```

::: {.callout-tip}
### Run It Yourself

**Your turn, about 6 minutes.** Two baseline definitions, two answers.
Decide which you would put in a study report, and be ready to say why.
This is the discussion below.

The week-26 treatment difference in HbA1c is **-1.116** using the SAP's
baseline and **-1.138** using SDTM's operational flag. A gap of 0.022
percentage points.

Be careful how you read that. It is **small**, and this course is not
going to claim a trial would turn on it. It is also **real**, it is
**not the number the SAP defined**, and nothing in the code would have
told you it was wrong. It came from using an operational flag to answer
an analysis question, and 75 of 400 subjects had a different baseline
HbA1c as a result.

Errors of this shape do not announce themselves. They produce plausible
numbers, slightly wrong, from code that runs cleanly. That is why the
standard puts the decision in a variable with a name, and why "which
flag did you use" is a reasonable question to be asked at a review.

One note on the arithmetic, because it is the kind of thing worth
checking. The code above rounds each mean to three decimals and then
subtracts, so the difference you see equals the difference of the two
means you see. Subtracting first and rounding afterwards gives -1.117
rather than -1.116, because the unrounded difference is -1.1165. Both
are defensible; a table whose displayed difference does not match its
displayed means invites a reader to think you cannot subtract, which is
why this one rounds first.
:::

### One subject where the baseline choice really matters

`101-052` is the subject from [Findings](sdtm-findings.qmd) whose
screening HbA1c came back at 13.9%, tripped an edit check, was
investigated, and was confirmed as a real value:

```{r d5}
#| eval: true

adlb |>
  filter(USUBJID == "GLPX1-101-052", PARAMCD == "HBA1C") |>
  select(AVISIT, ADY, AVAL, ABLFL, BASE, CHG)
```

::: {.callout-note}
### One real value, anchoring everything after it

That 13.9% is this subject's most recent result on or before first dose,
so it is their baseline. Every `CHG` in the column below it is measured
from 13.9, and the week-26 change is **-6.5** against a study mean of
about -1.4.

Nothing here is wrong. The value is real, the definition was applied
correctly, and the subject genuinely improved a great deal. But it shows
how much weight a single baseline record carries: one measurement sets
the reference point for a subject's entire change profile, and an
extreme baseline produces an extreme change.

This is why baseline definitions are argued over in statistical analysis
plans rather than chosen casually, and why some studies define baseline
as an average of two pre-dose measurements. That is a legitimate
alternative, and the ADaMIG accommodates it directly: the derived
average becomes the baseline record, and `DTYPE` must be populated to
say so (ADaMIG v1.3, §3.3.8).
:::

### The finished dataset

```{r final}
#| eval: true

adlb_final <- adlb |>
  group_by(USUBJID) |>
  mutate(ASEQ = row_number()) |>
  ungroup() |>
  select(STUDYID, USUBJID, SUBJID, SITEID, ASEQ,
         TRTP, TRTA, SAFFL, TRTSDT,
         PARAM, PARAMCD, PARAMN,
         AVISIT, AVISITN, ADT, ADY,
         AVAL, BASE, CHG, PCHG, DTYPE,
         ABLFL, ANRLO, ANRHI, ANRIND, BNRIND,
         SRCDOM, SRCVAR, SRCSEQ)

adlb_final |>
  filter(PARAMCD == "HBA1C", AVISIT %in% c("BASELINE", "WEEK 26")) |>
  group_by(AVISIT, TRTA) |>
  summarise(n = n(), mean_AVAL = round(mean(AVAL), 2),
            mean_CHG = round(mean(CHG), 3), .groups = "drop")
```

This is written to `data/adam/adlb.csv` by `R/build_adam.R`, and it is
what the TLF sessions read. With ADSL, ADAE and ADLB in place, Part 3 is
complete: a subject-level dataset, an occurrence dataset, and a basic
data structure, each built from persisted SDTM and each traceable back
to it.

## What Can Go Wrong

**Reading task, about 10 minutes.** No code. Mostly about baseline
choices that look like technicalities and are not.

*Estimated time: ~10 minutes (reading)*

**Using the SDTM baseline flag for analysis.** `--LOBXFL` is
operationally derived; `ABLFL` records the SAP's decision, and only
`ABLFL` signifies baseline for analysis (ADaMIG v1.3, §3.5; SDTMIG v3.4,
§4.5.9). In this dataset the two disagree on 21% of subject-parameter
pairs.

**Trusting the visit label.** Defining baseline as "the BASELINE visit"
would have made 958 of 1,600 baselines post-dose here. Check a baseline
definition against the dosing dates, not the visit name.

**`BASE` populated with no `ABLFL` record.** Explicitly prohibited
(ADaMIG v1.3, §3.3.4.1), and easy to introduce by joining a baseline
value in from elsewhere rather than copying it from a flagged row.

**A derived baseline with null `DTYPE`.** If the baseline is an average
or an imputed value, `DTYPE` must be populated on that record (ADaMIG
v1.3, §3.3.8). A derived value that looks observed is a traceability
failure.

**A second baseline definition without `BASETYPE`.** Required as soon as
there is more than one, and required on *every* record of the affected
parameter where `BASE` is non-null (ADaMIG v1.3, §3.3.4.1).

**`PARAMCD` that will not transpose.** Over 8 characters, or starting
with a digit or an underscore, breaks the property the constraint exists
to protect (ADaMIG v1.3, §3.3.4.1).

**Units in a separate column.** `PARAM` must describe `AVAL`
unambiguously and needs no further qualifying variable (ADaMIG v1.3,
§3.3.4.1). A `PARAM` of `Glucose` with the unit elsewhere is not
analysis-ready.

**Copying all of ADSL into BDS.** ADSL is a source of subject-level
variables, which "does not mean that every ADSL variable should be
included" (ADaMIG v1.3, §2.3.2).

### Common misinterpretations

**"BDS is the ADaM version of the Findings class."** BDS datasets may be
derived from Findings, Events, Interventions, Special-purpose domains,
other ADaM datasets, or any combination (ADaMIG v1.3, §2.3.2). Ours
happens to come from LB, which makes the resemblance a coincidence
rather than a rule.

**"`AVISIT` must match `VISIT`."** `AVISIT` is derived and
producer-defined and "does not have to map to `VISIT` from the SDTM"
(ADaMIG v1.3, §3.3.3). It matches here because this trial's visits are
clean.

**"A row in ADLB is a row that was analysed."** `AVISIT` "does not mean
that the record was analyzed" (ADaMIG v1.3, §3.3.3). Selection for a
particular analysis is what `ANLzzFL` and the population flags are for.

**"There must be a codelist for `PARAMCD` somewhere."** There is not.
Analysis parameters are sponsor-defined, like `AGEGR1` and `ASEV` before
them.

## Discussion: Which Baseline Would You Defend?

**Discussion, about 10 minutes.** Alone, write your answer down; in a
session, this is the argument.

The SAP's baseline and SDTM's `LBLOBXFL` disagree on **342 of 1,600**
subject-parameter pairs. The week-26 treatment difference in HbA1c is
**-1.116** one way and **-1.138** the other.

1. **One minute, on your own.** Which number goes in the clinical study
   report, and what is your one-sentence justification?
2. **Five minutes, together.** The gap is 0.022. Somebody will say it is
   too small to matter. Decide, as a group, whether that is an argument
   or an excuse.
3. **The rest.** Suppose the gap had been large enough to change the
   conclusion. Does that change which definition is correct, and if not,
   what does it change?

The point is not that one number is right. It is that the choice was
made before anyone saw either number, and that is what makes it
legitimate.

## Exercises

**Exercise, about 15 minutes.** Do these before the next session.

Practice: [Exercise 6: ADLB](../exercises/adlb.qmd).

## Comprehension Check

**Self-test, about 5 minutes.** Answer before expanding.

1. GLPX-1's `ABLFL` lands on SCREENING for 958 subject-parameter pairs
   and on BASELINE for 642. Is that an inconsistency? Explain in one
   sentence.
2. Every disagreement between `ABLFL` and `LBLOBXFL` in this dataset is
   at `ADY == 1`. What is it about `ADY == 1` that causes this, and
   which definition does each flag use?
3. `DTYPE` is null on all 12,388 records of our ADLB. Under what change
   to the baseline definition would it stop being null, and on which
   rows?
4. Your ADLB has `BASE` populated but you removed `ABLFL` because "the
   baseline value is already there." Which rule does that break, and
   what can a reviewer no longer do?
5. A colleague proposes the `PARAMCD` values `HBA1C_PCT` and `2HRGLUC`.
   What, if anything, is wrong with each?

::: {.callout-note collapse="true"}
### Answers

1. **No.** It is one rule, "the last non-missing value on or before
   first dose", applied to subjects whose visits fell on different days
   relative to their own first dose. Where the BASELINE visit happened
   on or before dosing it supplies the baseline; where it happened after
   dosing, the most recent qualifying value is the screening draw. A
   consistent definition does not imply a consistent visit.
2. `ADY == 1` is the day of first dose, under the no-day-zero rule. A
   lab drawn that morning, before dosing, is **on or before** first
   dose, so the SAP's definition counts it, but it is not **strictly
   prior to** first exposure, which is how SDTM defines `--LOBXFL`
   (SDTMIG v3.4, §4.5.9). So `ABLFL` flags it and `LBLOBXFL` reaches
   back to screening.
3. If baseline were defined as, say, the average of the screening and
   pre-dose values, that average would be a new derived record, and
   `DTYPE` "must also be populated" on the baseline record it creates
   (ADaMIG v1.3, §3.3.8). It would be populated on the derived baseline
   rows only, not across the dataset.
4. "If `BASE` is populated for a parameter, and `BASE` is non-null for a
   subject for that parameter, then there must be a record flagged by
   `ABLFL` for that subject and parameter" (ADaMIG v1.3, §3.3.4.1). With
   the flag gone, `BASE` is a number with no visible provenance: a
   reviewer can no longer see *which* record it came from, so they
   cannot check that the baseline definition was applied correctly.
   Traceability is broken even if every value is right.
5. **Both are invalid, for different reasons** (ADaMIG v1.3, §3.3.4.1).
   `HBA1C_PCT` is 9 characters, one over the 8-character limit.
   `2HRGLUC` is 7 characters and so passes that test, but it starts with
   a digit rather than a letter. Both constraints exist so the values
   can serve as variable names in a transposed dataset, which is why
   checking beats eyeballing: the second one looks fine until you read
   the rule.
:::
