---
title: "The Analysis Data Reviewer's Guide"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(xml2)
library(knitr)
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 80 minutes.**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background: the document that says why | 20 min | Reading |
| Example 1: The inventory sections | 15 min | Worked example |
| Example 2: Treatment and imputation | 15 min | Worked example |
| Example 3: The conformance summary | 10 min | Worked example |
| The half that cannot be generated | 10 min | Reading |
| Exercises + Comprehension Check | 15 min | Practice / Self-test |

Reads after [Define-XML](define-xml.qmd) and
[Define-XML from a Specification Workbook](define-from-specs.qmd),
because the ADRG's whole reason for existing is what define.xml cannot
do. Split after Example 2 for self-paced study.
:::

::: {.callout-important}
### Scope of This Session

This session teaches **what the ADRG is, what it must contain, and how
to generate its data-driven components from R.**

It does not teach submission strategy, agency interaction, or review
expectations. Those require experience of running a submission, which
the author does not have, and a course that fakes them is worse than a
course that omits them. Where a section of the ADRG requires that kind
of judgement, this session says so and shows you what a real study
would put there, rather than inventing plausible prose.

Every regulatory statement below is cited to a published document you
can read yourself: the FDA **Study Data Technical Conformance Guide
v6.0 (March 2025)** and the PhUSE **ADRG Completion Guidelines v1.2
(2019-07-12)**.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Run R from the project root. The tables shown here are generated by
`R/build_adrg_components.R` and committed under `data/adrg/`, so you can
read them without rebuilding.

The PhUSE ADRG Package is free and does not require membership. It is
reference material, not redistributable by this course:
<https://advance.hub.phuse.global/wiki/spaces/WEL/pages/26804660/Analysis+Data+Reviewer+s+Guide+ADRG+Package>
:::

## When Do You Use This?

::: {.callout-tip}
Your ADaM datasets are built and your define.xml is generated. Then:

- Four subjects have no `AGE` because their birth dates are partial.
  define.xml can say `AGE` is derived. Where do you say *why four
  subjects do not have one*?
- Your SAP defines baseline as the last value on or before first dose.
  `ABLFL` records the result. Where do you record the *choice*, and the
  fact that a different defensible choice would have moved the headline
  number?
- A reviewer opens your submission with no memory of your study. What
  single document orients them?

**Where do the decisions live, once the data has recorded only their
consequences?**
:::

## Learning Objectives

After completing this session you will be able to:

- State what the ADRG is, what file it must be, and where the FDA
  guidance says it goes
- List the seven required sections and two optional sections of the
  PhUSE ADRG template
- Generate the ADRG's data-driven tables from analysis datasets and
  define.xml, rather than maintaining them by hand
- Distinguish the sections that can be generated from those that
  require a human, and explain why that boundary is where it is
- Produce a conformance issues summary from real checks, including when
  the findings are your own

## Background: The Document That Says Why

*Estimated time: ~20 minutes (reading)*

[Define-XML](define-xml.qmd) built a document that records, in
machine-readable form, what every variable is and where it came from.
That is a great deal of information. It is also, in a specific and
important way, silent.

`def:Origin Type="Derived"` tells a reviewer that `AGE` was computed. A
`MethodDef` can even tell them the algorithm. Neither tells them that
four subjects in GLPX-1 have no age at all, that this is because
their birth dates arrived partial, and that the study chose to leave
them null rather than impute. That is a decision, with a rationale, and
XML has no element for a rationale.

The FDA guidance is unusually direct about the ADRG's role, including
the part that sounds like a flaw:

> The ADRG **purposefully duplicates** limited information found in
> other submission documentation (e.g., the protocol, statistical
> analysis plan (SAP), clinical study report, define.xml) in order to
> provide FDA reviewers with a single point of orientation to the
> analysis datasets (Study Data Technical Conformance Guide v6.0,
> §2.3)

Duplication is normally a defect. Here it is the design. And the
guidance immediately closes the obvious misreading:

> It should be noted that the submission of an ADRG does not eliminate
> the requirement to submit a complete and informative define.xml file
> corresponding to the analysis datasets (TCG v6.0, §2.3)

So the ADRG is not a substitute for define.xml, and define.xml is not a
substitute for the ADRG. That gives the intuition to carry through the
session:

> **define.xml records what you decided. The ADRG exists to say why.**
> A machine-readable record of a decision is not an explanation of it,
> and the half of the ADRG that cannot be generated is precisely the
> half worth reading.

### What it is, concretely

The TCG is specific about the artifact, and this much is pure fact
rather than expectation:

- Its preparation is **"recommended as an important part of a
  standards-compliant analysis data submission for clinical trials"**
  (§2.3)
- It must be a PDF called **`adrg.pdf`** (§2.3)
- It goes with the analysis data in **Module 5 of the eCTD**, file-tagged
  `analysis-data-reviewers-guide` in eCTD v3.2.2, or with document type
  keyword `analysis data reviewers guide` in eCTD v4.0 (§2.3)

One detail is worth noticing because it is easy to overstate the
ADRG's status. For the *Study* Data Reviewer's Guide, the TCG's
footnotes record that the PhUSE templates "have been reviewed by FDA
and published in the Federal Register" and that "FDA prefers but does
not require" their use. **The ADRG footnote carries no equivalent
language.** It says only:

> A specific template for an Analysis Data Reviewer's Guide is not
> specified. However, an example can be found at
> https://phuse.global/Deliverables (TCG v6.0, §2.3, footnote 16)

No template is mandated. The PhUSE template is what the industry
converged on, not what the agency imposed.

### The seven required sections

The PhUSE ADRG Completion Guidelines are explicit about structure:

> The ADRG has seven required sections: Introduction; Protocol
> Description; Analysis Considerations Related to Multiple Analysis
> Datasets; Analysis Data Creation and Processing Issues; Analysis
> Dataset Descriptions; Data Conformance Summary; Submissions of
> Programs (ADRG Completion Guidelines v1.2, ADRG Overview)

Plus two optional ones: an **Appendix**, and a **Legacy Data Conversion
Plan & Report Appendix** for sponsors who converted non-SDTM data to
ADaM.

::: {.callout-warning}
### Mind the Version

This session uses **ADRG Completion Guidelines v1.2 (2019-07-12)**, the
current version. It matters, because v1.2 **renumbered section 3**:

| Topic | Earlier template | v1.2 |
|---|---|---|
| SDTM/ADaM content comparison | 3.1 | **5.1** |
| Core Variables | 3.2 | **3.1** |
| Treatment Variables | 3.3 | **3.2** |
| Visit windowing | 3.5 | **3.4** |
| Imputation/Derivation | 3.6 | **3.5** |
| Variable Conventions | 4.4 | folded into **3.4/3.5** |

An ADRG citing "section 3.2, Core Variables" is working from a
superseded template. This is the same lesson
[Define-XML](define-xml.qmd) taught about enumerations: check the
current publication, do not trust a section number you remember.
:::

## Example 1: The Inventory Sections

*Estimated time: ~15 minutes (worked example)*

Sections 1.3 and 3.1 are tables. Neither should ever be typed.

### 1.3 Study Data Standards and Dictionary Inventory

The guidelines ask for the ADaM, SDTM and Define versions used. Those
facts already exist in the define.xml, so read them from there, if the
ADRG and the define.xml disagree about which version of Define-XML the
study used, that is a finding, and it should be impossible rather than
merely unlikely.

```{r ex1-standards}
read_csv("data/adrg/01_standards_inventory.csv", show_col_types = FALSE) |>
  kable()
```

`R/build_adrg_components.R` pulls the Define-XML version out of the
`MetaDataVersion` element rather than restating it.

### 3.1 Core Variables

> Core variables are those that are represented across all/most
> analysis datasets (ADRG Completion Guidelines v1.2, §3.1)

"Represented across all analysis datasets" is a computation, not an
opinion. It is the intersection of the variable names:

```{r ex1-core}
read_csv("data/adrg/02_core_variables.csv", show_col_types = FALSE) |>
  kable()
```

### Run It Yourself

```{r ex1-run}
ads <- list(
  ADSL = read_csv("data/adam/adsl.csv", show_col_types = FALSE),
  ADAE = read_csv("data/adam/adae.csv", show_col_types = FALSE),
  ADLB = read_csv("data/adam/adlb.csv", show_col_types = FALSE)
)

reduce(map(ads, names), intersect)
```

### Reading the Output Line by Line

Six variables appear in all three analysis datasets. Two of them,
`STUDYID` and `USUBJID`, are there because ADaM requires them: the
guidelines note that as the minimum this table would contain. The other
four are this study's own shape: `SUBJID` and `SITEID` carried through
for traceability, `TRTSDT` because every dataset needs the treatment
start date to compute study day, and `SAFFL` because every analysis in
this study is run on the safety population.

That last one is the kind of thing worth a reviewer's attention, and it
is visible only because the table was computed rather than copied from
the last study's ADRG.

## Example 2: Treatment and Imputation

*Estimated time: ~15 minutes (worked example)*

### 3.2 Treatment Variables

The guidelines ask a yes/no question: are the values of planned
treatment equivalent in meaning to actual treatment? A cross-tabulation
answers it and shows the working:

```{r ex2-trt}
read_csv("data/adrg/03_treatment_agreement.csv", show_col_types = FALSE) |>
  kable()
```

Zero discrepancies across 400 subjects. In GLPX-1 nobody received a
treatment other than the one they were randomised to, which is a
simulation convenience, and an ADRG for a real study would very
probably have a non-empty "No" row and a paragraph explaining it.

### 3.5 Imputation/Derivation Methods

This is where the course's deliberate data defects finally get written
down where a reviewer would see them:

```{r ex2-imp}
read_csv("data/adrg/04_imputation_summary.csv", show_col_types = FALSE) |>
  kable()
```

Read the `Convention` column, because each entry is a decision that
[ADSL](adsl.qmd), [ADAE](adae.qmd) and [ADLB](adlb.qmd) argued through
at length:

- **Four subjects with no `AGE`**, partial birth dates, the D1 defect
  from [The DM Domain](sdtm-dm.qmd). Left null, not imputed. define.xml
  can say `AGE` is derived; only this table says four people have none.
- **One adverse event with no `ASTDT`**, and the convention matters
  more than the count. `TRTEMFL` is left **null, never `"N"`**, because
  "we do not know whether this event was treatment-emergent" is a
  different claim from "this event was not treatment-emergent". That
  distinction is invisible in the dataset and invisible in define.xml.
- **Zero records with no `BASE`**: a real number, not an absence of
  effort. It is worth stating precisely because a reader would
  otherwise wonder.

### 5.2 Analysis Datasets

```{r ex2-inv}
read_csv("data/adrg/05_dataset_inventory.csv", show_col_types = FALSE) |>
  kable()
```

The `Class` column is the same vocabulary
[Define-XML](define-xml.qmd) found in the `ItemGroupClass` enumeration
- including `OCCURRENCE DATA STRUCTURE` for ADAE, the decision
[ADAE](adae.qmd) made and defended.

## Example 3: The Conformance Summary

*Estimated time: ~10 minutes (worked example)*

Section 6 is required, and section 6.2 asks for an issues summary.
[Define-XML](define-xml.qmd) built the machinery: two
Specification-sourced conformance rules that no schema check will
catch. Run them against this study's own define.xml and report what
comes back.

```{r ex3-conf}
read_csv("data/adrg/06_conformance_issues.csv", show_col_types = FALSE) |>
  select(Rule, `Diagnostic Message`, Records) |>
  kable()
```

Rule 65 is clean. Rule 73 is not, and the finding is ours:

```{r ex3-detail}
read_csv("data/adrg/06_conformance_issues.csv", show_col_types = FALSE) |>
  filter(Records > 0) |>
  pull(Explanation) |>
  str_wrap(70) |>
  cat()
```

### Reading the Output Line by Line

Seventeen variables in GLPX-1's define.xml have `Origin Type="Derived"`
and no `MethodOID`: nothing states how they were derived. The cause is
upstream, in the specification workbook: `COMPUTATION_METHOD` defines
five methods, but twenty-three variables are marked derived.

**This finding is left in deliberately, and reported rather than
quietly fixed.** Two reasons, and both are the point of section 6:

1. It is what a real conformance summary looks like. A submission with
   zero findings is rarer than people expect, and the section exists
   because findings are normal. What matters is that each one is
   *known, stated, and explained*, not that the list is empty.
2. Fixing it belongs in the workbook, not in the XML and not in the
   ADRG, exactly as
   [the previous session](define-from-specs.qmd) argued. The ADRG
   reports the state of the submission; it does not launder it.

### The Same Rule, Two XPaths, Two Answers

Before trusting that seventeen, look at how it was counted. Rule 73
says a `Derived` origin needs a `MethodOID` on the `ItemRef` that
points at it. So you find the derived `ItemDef`s and check their
`ItemRef`s. The obvious way to write that is wrong.

```{r rule73-scope}
define <- read_xml("data/define/define.sdtm.xml")

derived <- xml_attr(
  xml_find_all(define, "//d1:ItemDef[def:Origin/@Type='Derived']"), "OID")

count_missing <- function(path) {
  sum(map_lgl(derived, function(oid) {
    ref <- xml_find_first(define, sprintf("%s[@ItemOID='%s']", path, oid))
    !is.na(ref) && is.na(xml_attr(ref, "MethodOID"))
  }))
}

c(narrow = count_missing("//d1:ItemGroupDef/d1:ItemRef"),
  wide   = count_missing("//d1:ItemRef"))
```

Same rule, same document, same day. Four findings appear or disappear
depending on one path fragment.

The narrow version looks reasonable: variables live in datasets, so
their `ItemRef`s live in `ItemGroupDef`. But **value-level definitions
are referenced from `def:ValueListDef`**, not from `ItemGroupDef`. For
those four, `xml_find_first()` returns a missing node, `!is.na(ref)`
is `FALSE`, and they are dropped.

Not skipped with a warning. Dropped.

```{r rule73-missed}
setdiff(
  keep(derived, function(oid) {
    ref <- xml_find_first(define, sprintf("//d1:ItemRef[@ItemOID='%s']", oid))
    !is.na(ref) && is.na(xml_attr(ref, "MethodOID"))
  }),
  keep(derived, function(oid) {
    ref <- xml_find_first(
      define, sprintf("//d1:ItemGroupDef/d1:ItemRef[@ItemOID='%s']", oid))
    !is.na(ref) && is.na(xml_attr(ref, "MethodOID"))
  })
)
```

The four value-level `LBSTRESN` definitions, the ones
[Define-XML](define-xml.qmd) built to give each lab parameter its own
units and ranges. They are derived, they have no method, and the narrow
check never looked at them.

::: {.callout-important}
### A check that reports zero is saying one of two things

*There are no findings*, or *I did not look there*. These are not the
same claim and the output is identical.

You cannot tell them apart by reading the number. You can only tell them
apart by reading the check, or by running it against a document you have
broken on purpose and confirming it complains.

This matters more than it sounds. Conformance tooling is how a sponsor
convinces itself a submission is sound. A check with an unadvertised
blind spot does not fail loudly, it reassures quietly, which is the one
behaviour you cannot afford in an instrument you are using to decide
whether you are done.
:::

A real ADRG's issues summary would carry, for each finding, an
explanation of why it is acceptable or what was done about it. Writing
that explanation is a judgement call, which brings us to the rest of
the document.

## The Half That Cannot Be Generated

*Estimated time: ~10 minutes (reading)*

Everything above was a table computed from data. Now count what is
left. Of the seven required sections, these cannot be generated at all:

| Section | What it needs | Why no script can write it |
|---|---|---|
| 2.1 Protocol Number and Title | The protocol | Not in the data |
| 2.2 Protocol Design in Relation to ADaM Concepts | Study design judgement | Relates the *design* to the variables; requires knowing the intent |
| 3.3 Subject Issues Requiring Special Analysis Rules | Study conduct knowledge | Which subjects, and why they are special, is a clinical matter |
| 3.4 Visit Windowing and Record Selection | The SAP | The rule is a choice, not a property of the data |
| 4.1–4.3 Split Datasets, Dependencies, Intermediate Datasets | Pipeline knowledge | Describes how the programming was organised |
| 5.1 Overview | SAP objectives | Asks whether the datasets support every protocol objective |
| 7 Submission of Programs | The submission package | Itemises what is being shipped |

::: {.callout-note}
### VERIFY, sections this course does not write

Sections 2.2, 3.3, 4.1–4.3, 5.1 and 7 require submission experience and
study conduct knowledge that this course does not have and will not
fake.

For GLPX-1 they would say roughly this, and a real study would say it
with authority this course cannot claim:

- **2.2** would relate `TRT01P`/`TRT01A` and the analysis periods to
  the trial's two-arm parallel design with eight visits over 26 weeks.
- **3.4** would state the visit-windowing rule from the SAP. GLPX-1
  uses nominal visits with no windowing, which is unusually simple and
  would itself be worth stating.
- **4.2** would describe that ADSL is built first and ADAE and ADLB
  both read it: the dependency `R/build_adam.R` encodes.
- **5.1** would confirm whether the datasets support every objective in
  the protocol and SAP. GLPX-1 has no formal SAP, so this cannot be
  answered honestly here at all.
- **7** would itemise the programs shipped. This course ships
  `R/build_sdtm.R`, `R/build_adam.R`, `R/build_define.R` and
  `R/build_adrg_components.R`, but a submission's program inventory has
  requirements this course has not verified.

If you are writing a real ADRG, read the Completion Guidelines section
by section. They are 33 pages and they answer these questions properly.
:::

That table is the session's real content. **Roughly half the ADRG is a
report and half is an explanation**, and the boundary is not arbitrary:
everything that is a *property of the data* can be generated, and
everything that is a *reason for a choice* cannot. A script knows that
`TRTEMFL` is null on one record. Only a person knows that null was
chosen over `"N"` because the two mean different things.

### A note on `ts.xpt`

Three earlier sessions promised you would meet `ts.xpt` again here.
[Why Standards](why-standards.qmd) taught the chain by which the FDA's
technical rejection criteria check a submission's study tagging file
before any human opens it.

The connection to this session is narrow and worth stating plainly:
`ts.xpt` is where the *machine* learns which standards a study used;
ADRG section 1.3 is where a *person* does. The same facts, twice, for
two different readers, which is the "purposefully duplicates"
principle again. GLPX-1 has no TS domain, so this course cannot build
one, and how `ts.xpt` is populated for a real submission is outside
what this session teaches.

## What Can Go Wrong

- **Hand-maintaining the generated tables.** The core-variables table
  copied from the last study's ADRG will be wrong in a way nobody
  notices, because it looks right. Compute it.
- **An empty issues summary.** If section 6.2 is empty, the likeliest
  explanation is that nothing was checked, not that nothing was found.
- **Citing a superseded section number.** v1.2 renumbered section 3.
  "See ADRG section 3.2 for core variables" is a v1.1 sentence.
- **Treating the ADRG as a substitute for define.xml.** The TCG closes
  this off explicitly (§2.3), and it is worth knowing the sentence
  exists.
- **Writing the ADRG last.** Every table above was derivable the moment
  the datasets existed. The sections that need a human are the ones
  whose reasoning is freshest while the work is being done, and
  hardest to reconstruct six months later.

## Exercises

[Exercise 10: The Analysis Data Reviewer's Guide](../exercises/adrg.qmd)
asks you to close the rule-73 finding at its source, and to draft the
one narrative section this course can honestly ask you to write.

## Comprehension Check

::: {.callout-note collapse="true"}
### Questions

1. The TCG says the ADRG "purposefully duplicates" information found in
   the protocol, SAP and define.xml. Why is that not a design flaw?
2. GLPX-1's define.xml records that `AGE` has `Origin Type="Derived"`.
   What can it not record, and which ADRG section carries it?
3. Your ADRG's section 6.2 lists seventeen rule-73 findings. Where do
   you fix them, and what do you do about the ADRG in the meantime?
4. Which of the seven required sections could be regenerated
   automatically after a data refresh, and which would need rewriting
   by hand?
5. A colleague's ADRG has an empty Data Conformance Summary. What is
   the most likely explanation?
:::

::: {.callout-note collapse="true"}
### Answers

1. **Because the reviewer needs a single point of orientation.** The
   information exists elsewhere, but scattered across documents in
   different formats written for different purposes. The ADRG's job is
   to be the one place someone with no memory of the study can start,
   and that job requires repeating things (TCG v6.0, §2.3).

2. **It cannot record that four subjects have no `AGE`, why, or what
   was decided about it.** `Origin` describes the variable; it says
   nothing about the records where the derivation could not be
   performed. That belongs in **section 3.5,
   Imputation/Derivation Methods**, where GLPX-1 states the partial
   birth dates, the count, and the decision to leave the values null
   rather than impute.

3. **Fix them in the specification workbook**, add the missing
   `COMPUTATION_METHOD` entries and the `COMPUTATIONMETHODOID`
   references, then regenerate. Not in the XML, which would be
   overwritten, and not in the ADRG, which reports the state of the
   submission rather than improving it. In the meantime the ADRG
   reports the findings honestly, with an explanation, because that is
   what section 6.2 is for.

4. **Regenerable: 1.3, 3.1, 3.2, 3.5, 5.2, and the findings half of 6.**
   **By hand: 2.1, 2.2, 3.3, 3.4, 4.1–4.3, 5.1, 7, and the
   *explanation* half of 6.2.** The boundary is properties of the data
   versus reasons for a choice.

5. **That nothing was checked.** Section 6.1 asks which conformance
   software was used and against which version of the checks; an empty
   6.2 with a filled 6.1 would be a strong claim, and an empty 6.2 with
   an empty 6.1 is not a claim at all. Findings are normal; the section
   exists because they are.
:::
