---
title: "ADSL"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(sdtm.terminology)
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 90 minutes.**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background: the roster every analysis starts from | 20 min | Reading |
| Example 1: The ADSL spine, subjects, treatment, dates | 25 min | Worked example |
| Example 2: Population flags and disposition | 20 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Exercises + Comprehension Check | 10 min | Practice / Self-test |

Builds the course's first real ADaM dataset (one row per subject) from the persisted SDTM. For self-paced study, split after Example 1.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown. Run R
from the project root, as in earlier sessions.

This session reads `data/sdtm/dm.csv` and `data/sdtm/vs.csv`: the
finished SDTM from Part 2, and writes the first dataset in
`data/adam/`. The whole chain is now visible: raw → SDTM → ADaM.
:::

## When Do You Use This? {#sec-adsl}

::: {.callout-tip}
Every table in a clinical study report needs a denominator: *how many
subjects, in which population?* Every by-treatment summary needs to know
each subject's treatment and when they were on it. Every safety listing
needs to know who was dosed. A reviewer opening the submission needs one
authoritative answer to "who was in this trial?"

Ask each of those questions of SDTM and you get a different, partial
answer from a different domain. That is the gap this session closes, and
it raises one question: **what is the one dataset that every other
analysis dataset and every table depends on?**

Builds on [ADaM Concepts](adam-concepts.qmd), where you saw the ADSL
*shape*, and on the persisted SDTM DM from [The DM Domain](sdtm-dm.qmd).
:::

## Learning Objectives

After completing this session you will be able to:

- Build an ADSL with exactly one row per subject from SDTM DM
- Derive the treatment variables (`TRT01P`, `TRT01A`) and treatment
  dates (`TRTSDT`, `TRTEDT`, `TRTDURD`) and say where each comes from
- Derive subject-level population flags and explain why they are ADaM's
  rather than SDTM's
- Distinguish a variable *carried* from SDTM from one *derived* for
  analysis, and know which rule governs each
- Recognise where a dataset's derivations rest on assumptions that must
  be documented rather than assumed away

## Background: The Roster Every Analysis Starts From

**Reading task, about 20 minutes.** ADSL is the dataset every other
analysis dataset borrows from. Read the whole thing before building.

*Estimated time: ~20 minutes (reading)*

::: {.callout-important}
### The one thing to remember

**ADSL is the trial's roster: exactly one row per subject that every
analysis starts from (which population, which treatment, from when to
when) each value carried or derived from SDTM and traceable to it.**
:::

An ADaM-compliant ADSL is **required in a CDISC-based submission even if
no other ADaM datasets are submitted**, and its structure is one record
per subject regardless of trial design (ADaMIG v1.3, §3.2). That is a
strong claim, and it follows from what ADSL is for. `USUBJID` is the key
variable that links ADSL to every other dataset, SDTM and ADaM alike
(ADaMIG v1.3, §3.2). Filter a BDS dataset to a population, split a table
by treatment, count a denominator: all of it comes back to ADSL.

The corollary matters as much: ADSL holds subject-level facts only.
Anything that varies over time belongs in a BDS dataset (ADaMIG v1.3,
§2.3.1). A week-26 HbA1c is not an ADSL variable, however convenient it
would be to put it there.

### Two kinds of variable, two different rules

Almost every ADSL variable is one of two things, and confusing them is
the most common way to get ADSL wrong.

Some variables are **carried from SDTM**. `STUDYID`, `USUBJID`,
`SUBJID` and `SITEID` are all Required in ADSL and sourced directly from
DM; note that `SUBJID` and `SITEID` are *required in ADSL* though merely
permissible in other datasets (ADaMIG v1.3, §3.2). The demographics
follow the same pattern: `AGE`, `AGEU`, `SEX` and `RACE` are Required
and sourced from `DM.AGE`, `DM.AGEU`, `DM.SEX`, `DM.RACE` (ADaMIG v1.3,
§3.2).

When you carry a variable across, one rule governs it absolutely:

> Any variable in an ADaM dataset whose name is the same as an SDTM
> variable must be a copy of the SDTM variable, and its label, meaning,
> and values must not be modified — a principle of harmonization known
> as **"same name, same meaning, same values"** (ADaMIG v1.3, §3.1.1,
> item 3).

So if `SEX` appears in ADSL it *is* `DM.SEX`, untouched. If you need a
modified version, it is mandatory to do that in an ADaM variable with a
different name; changing the value under an SDTM name is prohibited
(ADaMIG v1.3, §3.1.1, item 3).

This rule also explains something the ADSL variable tables do not: our
ADSL carries `ETHNIC` and `COUNTRY`, but **neither appears anywhere in
the ADaMIG's ADSL variable tables**. They are not ADaM standard
variables. They are legitimate all the same, as SDTM variables copied
in under the same-name rule, which is exactly what §2.2 asks for when it
says all SDTM variables used in ADaM derivations should be included when
practical, for traceability. The lesson generalises: *not everything in
a conformant ADSL is an ADaM standard variable*, and the standard tells
you how to include the rest rather than forbidding it.

Other variables are **derived for the analysis**. `AGEGR1`: an age
grouping such as `<65` / `>=65`, is the clearest case. The ADaMIG
defines `AGEGRy` as "a character description of a grouping or pooling of
the subject's age for analysis purposes" and offers `"<18"`, `"18-65"`,
`">65"` only as *examples* (ADaMIG v1.3, §3.2). The cut point is a
sponsor decision recorded in the statistical analysis plan, not a
standard value. As with `PARAMCD` in the last session, **there is no
codelist to check `AGEGR1` against**, because it is a design decision.

Grouping variables carry one structural rule worth internalising: a
numeric counterpart such as `AGEGR1N` must have a one-to-one
relationship with `AGEGR1`, cannot be present unless `AGEGR1` is, and on
any given row **either both are populated or both are null** (ADaMIG
v1.3, §3.2).

### Treatment: planned, actual, and why the distinction is conditional

`ARM` is Required in ADSL and sourced from `DM.ARM`; `ACTARM` is
Permissible, from `DM.ACTARM` (ADaMIG v1.3, §3.2). Alongside them sit
the ADaM treatment variables, which are period-aware:

- **`TRTxxP`**, planned treatment for period `xx`. In a one-period
  randomized trial `TRT01P` is the treatment the subject was randomized
  to, and it may be derived from `DM.ARM`. **At least `TRT01P` is
  required** (ADaMIG v1.3, §3.2).
- **`TRTxxA`**, actual treatment for period `xx`. This one is
  *conditionally* required: required when actual treatment does not
  match planned **and** there is an analysis of the data as treated
  (ADaMIG v1.3, §3.2).

GLPX-1 is single-period, and no subject received the wrong treatment,
so `TRT01A` is not strictly required here. We include it anyway: a
safety analysis is conventionally "as treated", and carrying the
variable makes that explicit rather than implied. Knowing it is
*conditional*, and why, is the point.

### Dates and duration

`TRTSDT` (date of first exposure) and `TRTEDT` (date of last exposure)
are each conditionally required, required if there is an
investigational product (ADaMIG v1.3, §3.2). They come from SDTM DM's
`RFXSTDTC` and `RFXENDTC`, which SDTMIG v3.4 §5.2 defines as the
earliest and latest exposure datetimes. That is a clean, documented
SDTM→ADaM derivation.

`TRTDURD`, total treatment duration in days, is Permissible (ADaMIG
v1.3, §3.2). Note what the guide does *not* do: it defines the variable
as "total treatment duration, as measured in days" but does not fix the
formula. Whether you count `TRTEDT − TRTSDT` or `TRTEDT − TRTSDT + 1` is
the sponsor's convention, and it must be documented. We use the
inclusive form (both endpoints count), so a subject dosed on one single
day has a duration of 1, not 0.

### Population flags belong to ADaM

At least one population flag is required by ADaM, and a minimum of one
subject-level population flag variable is required in ADSL (ADaMIG v1.3,
§3.2, Table 3.2.3). The flags are character `Y`/`N`, Conditional, and
named to a convention: character subject-level flags end in `FL`, their
numeric counterparts in `FN` (ADaMIG v1.3, §3.1.4, item 3). The standard
list includes `SAFFL`, `ITTFL`, `FASFL`, `PPROTFL`, `COMPLFL`, `RANDFL`
and `ENRLFL`, and is explicitly *not* meant to be all-inclusive (ADaMIG
v1.3, §3.2, Table 3.2.3).

Which of them you include is not a matter of taste. A character
indicator variable is required for **every population defined in the
statistical analysis plan**, and all applicable subject-level population
flags must be present in ADSL (ADaMIG v1.3, §3.1.4, item 2). The SAP
drives the variable list.

Two precise rules make these flags unlike almost everything else in
ADaM. First, on Core attributes generally: "unless otherwise specified,
all ADaM variables are populated as appropriate, meaning **nulls are
allowed**" (ADaMIG v1.3, §3). Population flags are the "otherwise
specified", for subject-level character population flags, `N` = not in
the population, `Y` = in it, and **null values are not allowed** (ADaMIG
v1.3, §3.1.4, item 4). A subject-level population flag must commit,
every time.

Second, and closing the thread opened in [Findings](sdtm-findings.qmd)
and [ADaM Concepts](adam-concepts.qmd): whenever population or baseline
flags are needed for analyses, **the ADaM variables must be used**
(ADaMIG v1.3, §3.5). SDTM carries some conceptually-parallel flags as
Supplemental Qualifiers - `SAFETY` maps to `SAFFL`, `ITT` to `ITTFL`,
`FULLSET` to `FASFL`, `PPROT` to `PPROTFL`, `COMPLT` to `COMPLFL`, but
the guide is blunt that they might not match:

> It is possible that ADaM subject-level population flags might not
> match their conceptual counterparts in the SDTM… It is not incumbent
> on those creating ADaM datasets to go back and "fix" the SDTM
> population supplemental qualifiers (ADaMIG v1.3, §3.5).

So the flags are re-derived in ADaM from stated conditions. They are
never copied across, even when they would come out identical.

### A scope limit worth naming

Two ADSL variables in this course have **no SDTM source**, because
GLPX-1's SDTM stops at DM, AE, LB and VS and never builds **DS
(Disposition)**: the domain where a real trial records randomization
and end-of-study.

- **`RANDDT`** (date of randomization) is conditionally required:
  "Required in randomized trials" (ADaMIG v1.3, §3.2). In a real
  submission it comes from DS. Here we set it equal to `TRTSDT`.
- **`EOSSTT`** (end-of-study status) and **`EOSDT`** (end-of-study date)
  are both Permissible (ADaMIG v1.3, §3.2). In a real submission they
  come from DS too. Here we derive them from visit attendance.

Example 1 checks the first of these against the data rather than
assuming it, and both are labelled as what they are. Say this plainly
now, because it is the kind of thing a course can quietly teach wrong:
**`RANDDT = TRTSDT` is not a general rule.** Randomization and first
dose are distinct events, and in most trials they differ by days. Treat
both derivations here as scaffolding that a real study would replace
with DS.

## Example 1: The ADSL Spine

**Worked example, about 25 minutes.** Three stops. Two of them are
checks that pass, and the useful question is what they are checking.

*Estimated time: ~25 minutes (worked example)*

Start from the persisted SDTM DM: one row per subject already, because
that is what DM is:

```{r read-dm}
#| eval: true

dm <- read_csv("data/sdtm/dm.csv", show_col_types = FALSE,
               col_types = cols(.default = col_character()))

c(rows = nrow(dm), subjects = n_distinct(dm$USUBJID))
```

Before deriving `RANDDT` from `TRTSDT`, check whether the two actually
coincide in this trial. The raw EDC file recorded a randomization date;
it never reached SDTM, because DM has no randomization-date variable,
but it is still there to test against:

```{r randdt-check}
#| eval: true

dm_raw <- read_csv("data/raw/dm_raw.csv", show_col_types = FALSE,
                   col_types = cols(.default = col_character()))

randdt_check <- dm_raw |>
  arrange(SUBJID, RANDDT) |>
  distinct(SUBJID, .keep_all = TRUE) |>
  select(SUBJID, RANDDT) |>
  inner_join(dm |> select(SUBJID, RFXSTDTC), by = "SUBJID") |>
  mutate(diff_days = as.integer(as.Date(RFXSTDTC) - as.Date(RANDDT)))

count(randdt_check, diff_days)
```

::: {.callout-tip}
### Run It Yourself

**Your turn, about 4 minutes.** Every subject returns zero. Ask yourself
whether that is reassuring or suspicious before you read the answer.

Run the chunk. Every one of the 400 subjects has `diff_days = 0`: first
dose fell on the randomization date, for all of them. That makes
`RANDDT = TRTSDT` **true in this dataset**, which is exactly why it
needs saying out loud that it is an artefact of how the trial was
simulated, not a property of randomized trials. A real study would take
`RANDDT` from DS and would find the two dates differing routinely.
:::

Now the spine. Identifiers and demographics are carried; treatment
variables and dates are derived:

```{r adsl-spine}
#| eval: true

adsl_spine <- dm |>
  mutate(
    TRTSDT  = as.Date(RFXSTDTC),                       # DM.RFXSTDTC
    TRTEDT  = as.Date(RFXENDTC),                       # DM.RFXENDTC
    TRTDURD = as.integer(TRTEDT - TRTSDT) + 1L,        # inclusive convention
    RANDDT  = TRTSDT,                                  # scope limit: real source is DS

    AGE     = as.numeric(AGE),
    AGEGR1  = case_when(AGE <  65 ~ "<65",             # sponsor-defined grouping
                        AGE >= 65 ~ ">=65"),
    AGEGR1N = case_when(AGEGR1 == "<65"  ~ 1,
                        AGEGR1 == ">=65" ~ 2),

    TRT01P  = ARM,                                     # planned, period 1
    TRT01PN = if_else(ARMCD == "GLPX10", 1, 2),
    TRT01A  = ACTARM                                   # actual, period 1
  ) |>
  select(STUDYID, USUBJID, SUBJID, SITEID,
         AGE, AGEU, AGEGR1, AGEGR1N, SEX, RACE, ETHNIC, COUNTRY,
         ARM, ARMCD, ACTARM, ACTARMCD, TRT01P, TRT01PN, TRT01A,
         RANDDT, TRTSDT, TRTEDT, TRTDURD)

c(rows = nrow(adsl_spine), subjects = n_distinct(adsl_spine$USUBJID))

adsl_spine |>
  select(USUBJID, SITEID, AGE, AGEGR1, SEX, TRT01P, TRTSDT, TRTEDT, TRTDURD) |>
  head(3)
```

The age grouping is worth a second look, because GLPX-1's deliberate
data defects reach it:

```{r agegr1}
#| eval: true

count(adsl_spine, AGEGR1, AGEGR1N)

adsl_spine |> filter(is.na(AGE)) |> select(USUBJID, AGE, AGEU, AGEGR1, AGEGR1N)
```

::: {.callout-tip}
### Run It Yourself

**Your turn, about 5 minutes.** Four subjects fall into no age group.
Decide what you would do about them before reading what the SAP would
have to say.

Run both chunks. 303 subjects are `<65`, 93 are `>=65`, and **four are
null**. Those are the four subjects from [The DM Domain](sdtm-dm.qmd)
whose birth dates were incomplete, so `AGE` could not be derived without
imputing. That null propagates: no age, no age group.

This is legal. Nulls are allowed in ADaM variables unless otherwise
specified (ADaMIG v1.3, §3), and `AGE` being Required means the
*variable* must be present, not that every value is populated. But legal
is not the same as decided. See the note below.
:::

::: {.callout-note}
### Reading the Output Line by Line

| What you see | What it means |
|---|---|
| 400 rows, 400 subjects | One row per subject, the defining ADSL structure (ADaMIG v1.3, §3.2). DM was already subject-level, so no collapsing was needed. |
| `SITEID`, `SUBJID` carried | Required *in ADSL*, permissible elsewhere (ADaMIG v1.3, §3.2). Carried unmodified under the same-name rule. |
| `AGEGR1` / `AGEGR1N` | Sponsor-defined grouping and its numeric counterpart. One-to-one, and null together on the same four rows: the rule from §3.2 holds. |
| `TRT01P` from `ARM` | Planned treatment for period 1, derived from `DM.ARM` (ADaMIG v1.3, §3.2). `TRT01P` is required; `ARM` is carried alongside it. |
| `TRTSDT` / `TRTEDT` | First and last exposure, from `DM.RFXSTDTC` / `DM.RFXENDTC` (SDTMIG v3.4, §5.2). Both conditionally required because there is an investigational product. |
| `TRTDURD = 183` | The planned 26-week treatment period, counted inclusively. Discontinuing subjects have shorter durations. You will see them in Example 2. |
| Four null `AGEGR1` | Missing data flowing through a derivation, honestly. |

**The decision the nulls force.** Four subjects have no age group, so
every age-subgroup table will have a fourth category or a dropped
denominator. Which of those happens is a decision, and it belongs in the
statistical analysis plan *before* the analysis runs: an explicit
"Missing" category, or documented exclusion from age-subgroup summaries.
Leaving it to whatever `count()` happens to do is how a table quietly
loses four subjects. ADaM's job here is not to hide the gap but to carry
it forward visibly.
:::

The carried columns are still governed by SDTM controlled terminology:
the same-name rule means they must be the SDTM values, so they must
still be valid ones. Check rather than assume:

```{r ct-check}
#| eval: true

ct_check <- tibble(
  variable = c("SEX", "RACE", "ETHNIC", "AGEU"),
  codelist = c("C66731", "C74457", "C66790", "C66781"),
  values   = list(unique(adsl_spine$SEX),
                  unique(adsl_spine$RACE),
                  unique(adsl_spine$ETHNIC),
                  unique(na.omit(adsl_spine$AGEU)))
) |>
  mutate(n_values  = lengths(values),
         all_valid = map2_lgl(values, codelist,
                              \(v, cl) all(is_term(v, clst_code = cl))))

ct_check |> select(variable, codelist, n_values, all_valid)
```

::: {.callout-tip}
### Run It Yourself

**Your turn, about 4 minutes.** Four passing checks. Read what each one
asserts; a check you do not understand is not evidence.

All four come back `TRUE`. Nothing new was introduced, that is the
point. Because these values were carried unmodified from SDTM, they
remain valid CT; had we "tidied" `RACE` into nicer labels under the same
variable name, we would have broken both the codelist and §3.1.1's
same-name rule in one move.
:::

## Example 2: Population Flags and Disposition

**Worked example, about 20 minutes.** One stop. Flags look trivial
until you need to explain who is excluded and why.

*Estimated time: ~20 minutes (worked example)*

Population flags are derived from *conditions*, never copied and never
hard-coded. GLPX-1 makes this feel almost too easy (every randomized
subject was dosed, so every flag comes out `"Y"`) but writing the
condition is what makes the dataset defensible, and it is what would
catch the subject who was randomized and never dosed:

```{r flags}
#| eval: true

adsl_flags <- adsl_spine |>
  mutate(
    RANDFL = if_else(!is.na(RANDDT), "Y", "N"),   # randomized
    ITTFL  = if_else(!is.na(RANDDT), "Y", "N"),   # ITT: analyse as randomized
    SAFFL  = if_else(!is.na(TRTSDT), "Y", "N")    # safety: received >= 1 dose
  )

count(adsl_flags, RANDFL, ITTFL, SAFFL)

# subject-level population flags may never be null (ADaMIG v1.3, §3.1.4)
adsl_flags |>
  summarise(across(c(RANDFL, ITTFL, SAFFL), ~ sum(is.na(.x))))
```

Now disposition. SDTM DM does not record whether a subject finished the
trial, so derive it from what the trial observed: the last visit each
subject actually attended:

```{r disposition}
#| eval: true

vs <- read_csv("data/sdtm/vs.csv", show_col_types = FALSE,
               col_types = cols(.default = col_character()))

eos <- vs |>
  group_by(USUBJID) |>
  summarise(
    last_visitnum = max(as.integer(VISITNUM)),
    EOSDT         = as.Date(max(VSDTC)),
    .groups       = "drop"
  ) |>
  mutate(EOSSTT = if_else(last_visitnum == 8L, "COMPLETED", "DISCONTINUED"))

adsl <- adsl_flags |>
  left_join(eos |> select(USUBJID, EOSSTT, EOSDT), by = "USUBJID")

count(adsl, EOSSTT)

adsl |>
  group_by(EOSSTT) |>
  summarise(n = n(), min_days = min(TRTDURD), median_days = median(TRTDURD),
            max_days = max(TRTDURD), .groups = "drop")
```

::: {.callout-tip}
### Run It Yourself

**Your turn, about 5 minutes.** Everyone is in every population. Work
out why that is true here and why it almost never is in a real study.

Run both chunks. All 400 subjects are `"Y"` on all three flags, with no
nulls. Then 367 subjects completed and 33 discontinued, and the
duration summary shows why that distinction matters: completers all sit
at 183 days, while discontinuing subjects range from 41 to 169. Those 33
are exactly the dropouts you first counted back in [The GLPX-1
Trial](the-trial-glpx001.qmd). The number has survived the whole
pipeline intact.
:::

One more check, because it exposes a distinction that is easy to
collapse:

```{r eosdt-vs-trtedt}
#| eval: true

adsl |>
  summarise(n_differ = sum(EOSDT != TRTEDT),
            n_same   = sum(EOSDT == TRTEDT))

adsl |>
  filter(EOSDT != TRTEDT) |>
  select(USUBJID, TRTEDT, EOSDT, EOSSTT, TRTDURD) |>
  head(4)
```

::: {.callout-note}
### Reading the Output Line by Line

| What you see | What it means |
|---|---|
| All flags `"Y"`, no nulls | The derivation is conditional, not hard-coded. Null is not an option for subject-level population flags (ADaMIG v1.3, §3.1.4, item 4). |
| `RANDFL` and `ITTFL` identical | In this trial the ITT population *is* the randomized population. That is a design fact worth stating in the SAP, not a redundancy to delete: the two flags answer different questions and could diverge in another trial. |
| 367 / 33 | Completers and discontinuers, reconciling with the trial overview from Part 1. |
| `EOSDT != TRTEDT` for 30 subjects | End of *study* is not end of *treatment*. These subjects stopped attending visits before their exposure record ended. Two variables, two questions. |

**On the disposition derivation.** `EOSSTT` and `EOSDT` here are derived
from visit attendance because this course's SDTM has no DS domain. In a
real trial, disposition is *collected*: the site records why and when a
subject came off study, and ADSL reads it from DS. Attendance is a
reasonable proxy and it is right for this dataset, but it is a proxy,
and it is documented as one. If a subject skipped the final visit yet
formally completed the study, this derivation would misclassify them; DS
would not.

Note also that `EOSSTT` is Permissible, with the codelist `SBJTSTAT` and
the ADaMIG's own examples being `COMPLETED` and `DISCONTINUED` (ADaMIG
v1.3, §3.2), which is where our two values come from.
:::

::: {.callout-warning}
### Why there is no `is_term()` check here

`SBJTSTAT` is an **ADaM** codelist, and the `sdtm.terminology` package
this course uses carries **SDTM** controlled terminology. Search it and
`SBJTSTAT` is absent; the term `DISCONTINUED` does not appear in any
SDTM codelist at all. So the honest position is: our two values are
taken from the ADaMIG's own stated examples for `EOSSTT`, and verifying
them against the published `SBJTSTAT` codelist needs the ADaM CT
package, which is a separate release from NCI EVS.

Worth knowing in its own right - **ADaM has its own controlled
terminology, distinct from SDTM's**, and a tool that validates one will
silently tell you nothing about the other. "The codelist wasn't found"
and "the value is invalid" are different findings, and conflating them
is how a validation report becomes decoration.
:::

### Persisting ADSL

As with SDTM in Part 2, the derivations above are reified in a script so
later sessions do not rebuild the chain. `R/build_adam.R` reads
`data/sdtm/` and writes `data/adam/adsl.csv`:

```{r build-adam}
#| eval: false

source("R/build_adam.R")
main()
#> adsl: 400 rows x 28 cols
```

It is deterministic: its inputs are the committed SDTM CSVs, which are
themselves fixed-seed output of `simulate_trial.R` and `build_sdtm.R`.
The whole chain reproduces from three scripts, and `data/adam/adsl.csv`
is committed so the [TLF](tlf.qmd) session can read a finished ADSL
directly. That is the same reproducibility argument that justified
committing `data/sdtm/`, one link further along.

## What Can Go Wrong

**Reading task, about 10 minutes.** No code. Mostly about copying SDTM
values into ADSL without checking whether they still mean the same
thing.

*Estimated time: ~10 minutes (reading)*

**More than one row per subject**

ADSL is one record per subject, full stop (ADaMIG v1.3, §3.2). The usual
culprit is a join to a multi-record source (a subject with two enrolment
records, two exposure periods, two disposition events) that silently
fans the dataset out. Every downstream denominator inherits the error.
Check `nrow()` against `n_distinct(USUBJID)` after every join, and
remember it was a real risk here: subject `101-012` has two raw
enrolment records, resolved back in [The DM Domain](sdtm-dm.qmd).

**Copying SDTM's population flags**

The SDTM Supplemental Qualifier flags are not the analysis flags.
Whenever a flag is needed for analysis, the ADaM variable must be used
and re-derived (ADaMIG v1.3, §3.5), and the guide explicitly allows the
two to disagree. Copying `SAFETY` into `SAFFL` skips the derivation ADSL
exists to record.

**Hard-coding a flag because "they're all Y"**

Writing `SAFFL = "Y"` produces the right column for this trial and no
defensible answer to "how was it derived?" It also breaks silently the
moment one subject is randomized but never dosed. Derive from the
condition; let the data produce the uniformity.

**Leaving a population flag null**

Subject-level population flags do not permit nulls (ADaMIG v1.3, §3.1.4,
item 4). A subject is either in the population or not. This is a genuine
exception to ADaM's general "nulls are allowed" (ADaMIG v1.3, §3), and it
is deliberate: a null denominator is not a denominator.

**Modifying a value under an SDTM variable name**

Relabelling `RACE`, recoding `SEX`, tidying `COUNTRY`: all prohibited
under the same-name rule (ADaMIG v1.3, §3.1.1, item 3). Whenever values
are modified it is mandatory to do so in an ADaM variable with its own
name, such as `RACEGR1`.

**Deriving `TRTSDT` from the wrong date**

`TRTSDT` is first *exposure*, from `RFXSTDTC`, not consent, not
randomization, not the first visit. Using the wrong reference date shifts
every study day and every treatment-emergent classification built on it
downstream.

**Putting time-varying data in ADSL**

Anything that changes over time belongs in BDS (ADaMIG v1.3, §2.3.1). The
pressure to add "just the week-26 value" to ADSL is real; resist it.

::: {.callout-warning}
### Common misinterpretations

**"Every variable in ADSL is an ADaM standard variable"**
No. Our `ETHNIC` and `COUNTRY` appear in no ADaMIG ADSL table. They are
conformant as SDTM variables carried in under the same-name rule (ADaMIG
v1.3, §3.1.1, item 3), which §2.2 encourages for traceability. A
variable's absence from the tables is not a prohibition.

**"`Required` means the value can never be missing"**
`Req` means the *variable* must be included in the dataset (ADaMIG v1.3,
§3). Unless otherwise specified, nulls are allowed. `AGE` is Required and
is null for four GLPX subjects, legitimately, because their birth dates
were incomplete and nothing was imputed.

**"`RANDDT` equals `TRTSDT`"**
It does in this simulated trial, for all 400 subjects, and that is a
property of the simulator, not of randomized trials. Randomization and
first dose are distinct events, normally days apart, and `RANDDT`'s real
source is the DS domain.

**"`AGEGR1` is controlled terminology"**
It is sponsor-defined for the analysis (ADaMIG v1.3, §3.2). The ADaMIG's
`"<18"` / `"18-65"` / `">65"` are examples, not a codelist. Same lesson
as `PARAMCD` in [ADaM Concepts](adam-concepts.qmd): there is no
`is_term()` check for a design decision.

**"If a value passes no validation check, it must be fine"**
`EOSSTT` in this session passes no CT check, not because it is valid,
but because the tool covers SDTM terminology and `SBJTSTAT` is an ADaM
codelist. An unchecked value is unchecked, not verified.
:::

## Exercises

**Exercise, about 10 minutes.** Do these before the next session.

Work through [Exercise 4: ADSL](../exercises/adsl.qmd), which extends
this ADSL with a second age grouping and an analysis-age case, and asks
you to construct a population-flag scenario where the flags legitimately
disagree.

## Comprehension Check

**Self-test, about 10 minutes.** Answer before expanding.

*Estimated time: ~5 minutes*

1. ADSL has 400 rows and 400 subjects. After joining an exposure dataset
   it has 412 rows. What has gone wrong, and which ADaMIG requirement
   does the result now violate?
2. `AGE` is Required in ADSL, yet four GLPX subjects have a null `AGE`.
   Is the dataset non-conformant? Explain.
3. Your SDTM has a `SAFETY` Supplemental Qualifier for every subject. Can
   you use it to populate `SAFFL`? Why or why not?
4. Why does `ETHNIC` appear in this ADSL even though it is not listed in
   any ADaMIG ADSL variable table, and what rule constrains its values?
5. `EOSSTT` was not checked with `is_term()`. Does that mean its values
   are unverified, invalid, or standard-compliant? Explain the
   distinction.

::: {.callout-note collapse="true"}
### Answers

1. A one-to-many join fanned the dataset out: most likely to a source
   with more than one record for some subjects (multiple exposure
   periods, a re-enrolment). ADSL must have exactly one record per
   subject (ADaMIG v1.3, §3.2), so 412 rows for 400 subjects is
   structurally non-conformant, and every denominator computed from it
   would be wrong.

2. It is conformant. `Req` means the *variable* must be included in the
   dataset (ADaMIG v1.3, §3); it does not require every value to be
   populated, since "unless otherwise specified, all ADaM variables are
   populated as appropriate, meaning nulls are allowed." The four nulls
   follow from incomplete birth dates that were deliberately not imputed.
   What *is* required is that the SAP says how those subjects are handled
   in age-subgroup analyses.

3. No, not by copying it. Whenever population flags are needed for
   analysis, the ADaM variables must be used (ADaMIG v1.3, §3.5), and the
   guide explicitly allows the SDTM Supplemental Qualifier and the ADaM
   flag to disagree, without obliging you to "fix" the SDTM. Derive
   `SAFFL` in ADaM from its stated condition; the SDTM qualifier may
   inform the definition, but it is not the source of the value.

4. Because a variable that is not an ADaM standard variable may still be
   carried from SDTM - §2.2 encourages including the SDTM variables used
   in ADaM for traceability. Its values are constrained by the same-name
   rule: an ADaM variable whose name matches an SDTM variable must be a
   copy, with label, meaning and values unmodified (ADaMIG v1.3, §3.1.1,
   item 3). So `ETHNIC` must remain exactly `DM.ETHNIC`, valid against
   the SDTM `ETHNIC` codelist.

5. **Unverified.** `EOSSTT` uses the `SBJTSTAT` codelist, which belongs
   to ADaM controlled terminology; the `sdtm.terminology` package carries
   SDTM terminology only, so the codelist is not there to check against.
   The values are taken from the ADaMIG's own examples for `EOSSTT`
   (ADaMIG v1.3, §3.2), which is good evidence but not a codelist check.
   "No check was possible" is a different finding from "the check
   passed", and reporting it as the latter is how validation becomes
   theatre.
:::
