---
title: "Dataset-JSON"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(jsonlite)
library(haven)
library(datasetjson)
library(knitr)
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 110 minutes.**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background: the format the standards bent around | 20 min | Reading |
| Example 1: The same domain, twice | 20 min | Worked example |
| Example 2: What transport cannot say | 15 min | Worked example |
| Example 3: The size question, honestly | 15 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Discussion: Does a Better Format Make It Submittable? | 10 min | Discussion |
| Exercises + Comprehension Check | 15 min | Practice / Self-test |

The last session in the course, and the only one about how data
*travels* rather than how it is shaped. It closes a thread opened in
Part 2: every session has referred to `dm.xpt` and `lb.xpt` without ever
producing one. Here we finally do, and then produce the format meant to
replace it.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Run R from the project root. Three new packages: **`datasetjson`** for
reading and writing Dataset-JSON, **`xportr`** for writing SAS Transport
v5, and **`haven`** for reading it back. All pinned in `renv.lock`.

Most of what follows runs against artifacts committed in this
repository: `data/xpt/` and `data/json/`, produced by
`R/build_dataset_json.R`.

**Example 3 also compares our files against CDISC's own.** Those are
**public on GitHub**, with no CDISC account needed, unlike Define-XML.
Clone the repository into `refs/` before you start:

```bash
git clone --depth 1 https://github.com/cdisc-org/DataExchange-DatasetJson.git \
  refs/DatasetJSON_v1.1/repo
```

That gives you the specification (as Markdown, in `doc/`), the JSON
Schema (`schema/`), and CDISC's example datasets in every format
(`examples/`). `refs/` is gitignored, so nothing you download is
committed.

Skip the clone and everything except the two comparison chunks in
Example 3 still runs.
:::

## When Do You Use This?

::: {.callout-tip}
Your datasets are built, your define.xml describes them and your ADRG
explains them. Then:

- The submission has to physically arrive somewhere. In what file
  format, and who decided?
- Your `ADLB` has a variable called `LBSTRESN` rather than
  `LABORATORY_RESULT_NUMERIC`. Why is every CDISC variable name eight
  characters or fewer? Is that a style rule?
- [ADAE](adae.qmd) insisted that a null `TRTEMFL` and a value of `"N"`
  mean different things. Does the file format you submit in preserve
  that distinction?

**What does the container do to the data you put in it?**
:::

## Learning Objectives

After completing this session you will be able to:

- Describe the constraints of SAS Transport v5 and how they shaped
  CDISC's own standards
- Write a dataset as both `.xpt` and Dataset-JSON v1.1 from R
- Read the required and optional attributes of a Dataset-JSON file, and
  explain what `itemGroupOID` is doing there
- Demonstrate a data distinction that transport format destroys and
  Dataset-JSON preserves
- Interpret file-size comparisons between the two formats without
  overclaiming

## Background: The Format the Standards Bent Around

**Reading task, about 20 minutes.** The last background section in the
course, and it reframes a constraint you have been living with since
session one without being told about it.

*Estimated time: ~20 minutes (reading)*

Every session in Part 2 referred to files this course never made.
`def:ArchiveLocationID="dm.xpt"`. `<def:leaf xlink:href="lb.xpt">`.
GLPX-1's data has lived in CSV throughout, because CSV is convenient for
teaching. A real submission does not use CSV.

It uses **SAS Transport Format Version 5**. Not SAS's current format:
a specific, frozen one that "dates from 1989 and was first available as
part of SAS version 5" (Transport for the Next Generation v1.0). The FDA
is unambiguous about its standing:

> The Transport Format (XPORT) Version 5 is the file format for the
> submission of all electronic datasets (Study Data Technical
> Conformance Guide v6.0, §3.1.1)

PhUSE's 2017 paper *Transport for the Next Generation*, which is the
document that started the effort to replace it, lists what it costs:

> Only supports US ASCII Character encoding. No multibyte characters are
> possible; this requires translation and/or transcription from the
> source data.
>
> Field names are restricted in terms of width and format. Field names
> must be alphanumeric, Variable names are limited to a maximum length
> of **8 characters**, Variable labels are restricted to a maximum
> length of **40 characters**.
>
> Character field widths are limited to **200 characters**.
>
> Does not make efficient use of storage space. There is often empty
> space for columns allocated, but not used by data and this can lead
> up to **70% wasted space**.
> (Transport for the Next Generation v1.0, "SAS Transport v5")

Read the second item again, then look at any variable name in this
course. `USUBJID`. `LBSTRESN`. `TRTEMFL`. `AOCCIFL`.

### The constraint you never noticed

```{r bg-namecheck}
datasets <- c("data/sdtm/dm.csv", "data/sdtm/lb.csv",
              "data/adam/adsl.csv", "data/adam/adae.csv", "data/adam/adlb.csv")

all_names <- datasets |>
  map(~ names(read_csv(.x, n_max = 0, show_col_types = FALSE))) |>
  unlist()

c(variables = length(all_names),
  over_8_characters = sum(nchar(all_names) > 8))
```

**Not one variable in 131 exceeds eight characters.** That is not
restraint on the part of whoever named them, and it is not coincidence.
`LBSTRESN` is called `LBSTRESN` rather than something readable because a
1989 file format could not have stored a longer name.

This is the thing worth taking from the session, and it is larger than
one file format:

> **The container shaped the contents.** SDTM and ADaM variable names
> are what they are because of the format they had to travel in. A
> constraint you cannot see, because everything you have been taught
> already accommodates it.

Dataset-JSON is the attempt to remove the constraint:

> Dataset-JSON is a data exchange standard for sharing tabular data
> using JSON. It is designed to meet a wide range of data exchange
> scenarios, including regulatory submissions and API-based data
> exchange. Each Dataset-JSON dataset can **optionally** reference a
> Define-XML document containing more complete metadata for the dataset
> (Dataset-JSON v1.1, Introduction)

Two rules from the same section, both worth knowing before you open a
file:

> Dataset-JSON must contain only **1 dataset per file**. Dataset-JSON
> uses the file extension .json (Dataset-JSON v1.1, Introduction)

## Example 1: The Same Domain, Twice

**Worked example, about 20 minutes.** The same data written two ways.
Compare the files, not just the code.

*Estimated time: ~20 minutes (worked example)*

`R/build_dataset_json.R` writes DM and LB in both formats. Variable
labels come from `data/spec/SDTM_METADATA.xlsx`: the same workbook
[the previous session](define-from-specs.qmd) used to generate
define.xml, so the transport file, the metadata document and the JSON
all tell one story.

```{r ex1-files}
tibble(directory = c("data/xpt", "data/json")) |>
  mutate(files = map_chr(directory, ~ paste(list.files(.x), collapse = ", "))) |>
  kable()
```

### What is in the JSON

```{r ex1-header}
dm <- fromJSON("data/json/dm.json", simplifyVector = FALSE)

tibble(attribute = setdiff(names(dm), c("columns", "rows"))) |>
  mutate(value = map_chr(attribute, ~ paste(unlist(dm[[.x]]), collapse = " / "))) |>
  kable()
```

Compare that against what the specification requires. Of the
seventeen top-level attributes, only seven are **Required**:
`datasetJSONCreationDateTime`, `datasetJSONVersion`, `itemGroupOID`,
`records`, `name`, `label` and `columns`.

One pairing repays attention:

| Attribute | Usage |
|---|---|
| `itemGroupOID` | **Required** |
| `metaDataRef` | Optional |
| `metaDataVersionOID` | Optional |

`itemGroupOID` is `IG.DM`. That is not a Dataset-JSON invention. It is
the OID from [Define-XML](define-xml.qmd), the identifier the
specification insisted has "no intrinsic meaning." Here is what it is
for. Every column carries one too:

```{r ex1-columns}
dm$columns |> head(4) |> map_dfr(as_tibble) |> kable()
```

`IT.DM.STUDYID`, `IT.DM.USUBJID`: the same OIDs the hand-built
define.xml declared. So a Dataset-JSON file **must** carry the keys that
join it to its metadata, but is **not** required to say which document
those keys belong to. It hands you a foreign key and leaves the table
optional.

### Row data

```{r ex1-rows}
dm$rows[1:2] |> map_chr(~ paste(map_chr(.x, ~ format(.x %||% "null")),
                                collapse = " | ")) |>
  str_trunc(78)
```

> **rows** is an array of records with variables values. Each record
> itself is also represented as an array of variables values
> (Dataset-JSON v1.1, Row Data)

An **array**, not an object. There are no keys on the rows at all: the
column names are stated once, in `columns`, and never repeated. A naive
JSON export would write `"USUBJID": "GLPX1-101-010"` on all 12,388 rows
of LB. This writes it zero times.

### Does it conform?

```{r ex1-validate}
#| eval: false
validate_dataset_json("data/json/dm.json")
```

```
File is valid per the Dataset JSON v1.1.0 schema
```

Checked against the JSON Schema CDISC publishes in the same repository
as the specification, exactly as [Define-XML](define-xml.qmd) checked
against the XSD. Both of GLPX-1's files pass.

## Example 2: What Transport Cannot Say

**Worked example, about 15 minutes.** One stop, and it closes a thread
running since ADAE: the difference between a missing value and an empty
one.

*Estimated time: ~15 minutes (worked example)*

[ADAE](adae.qmd) made a point at some length: subject `GLPX1-103-199`
has an adverse event with no start date, so `TRTEMFL` is **null**, not
`"N"`. "We do not know whether this event was treatment-emergent" is a
different claim from "this event was not treatment-emergent," and the
dataset must not conflate them.

The specification is explicit that Dataset-JSON preserves the
distinction:

> Missing values are represented by **null**. Empty strings are
> represented by **""** (Dataset-JSON v1.1, Row Data)

Which raises a question nobody asks: does the format we actually submit
in preserve it too?

### Run It Yourself

**Your turn, about 6 minutes.** Predict what each of the three values
looks like after a round trip through each format. One of them does not
survive, and which one matters.

Three values: a string, a genuine missing, and an empty string:

```{r ex2-setup}
probe <- tibble(ID = c("A", "B", "C"), VAL = c("x", NA_character_, ""))
is.na(probe$VAL)
```

Round-trip through SAS Transport v5:

```{r ex2-xpt}
xpt_path <- file.path(tempdir(), "probe.xpt")
write_xpt(probe, xpt_path, version = 5)
back_xpt <- read_xpt(xpt_path)

is.na(back_xpt$VAL)
```

Round-trip through Dataset-JSON:

```{r ex2-json}
cols <- tibble(itemOID = paste0("IT.PROBE.", names(probe)), name = names(probe),
               label = names(probe), dataType = "string", length = 8L)

ds <- dataset_json(probe, item_oid = "IG.PROBE", name = "PROBE",
                   dataset_label = "Missingness probe", columns = cols)

json_path <- file.path(tempdir(), "probe.json")
write_dataset_json(ds, json_path)
back_json <- read_dataset_json(json_path)

is.na(back_json$VAL)
```

### Reading the Output Line by Line

The original had one missing value: `FALSE TRUE FALSE`.

After the transport round-trip: **`FALSE FALSE FALSE`**. The missing
value is gone. Not lost as in corrupted, silently converted into an
empty string, indistinguishable from the value in row C that was
deliberately empty.

After the Dataset-JSON round-trip: `FALSE TRUE FALSE`. Unchanged.

Look at what is actually written to the file:

```{r ex2-raw}
readLines(json_path, warn = FALSE) |>
  paste(collapse = "") |>
  str_extract('"rows".*')
```

`null` for the missing value, `""` for the empty string. Two different
tokens for two different facts.

**This is the strongest practical argument in the session.** The course
has spent Part 3 insisting that null and a value are different claims -
`TRTEMFL` null rather than `"N"`, `AGE` null rather than imputed,
`AEDECOD` null because MedDRA was never applied. Every one of those
distinctions survives in the CSV and in the define.xml. Submitted as
transport format, the character ones quietly stop existing.

## Example 3: The Size Question, Honestly

**Worked example, about 15 minutes.** Our numbers point the opposite way
to CDISC's. Read why before deciding who is right.

*Estimated time: ~15 minutes (worked example)*

Dataset-JSON's stated aims include "the efficient use of storage space,"
against transport format's "up to 70% wasted space." So the JSON should
be smaller. Ours is not:

```{r ex3-sizes}
read_csv("data/json/size_comparison.csv", show_col_types = FALSE) |> kable()
```

**Uncompressed Dataset-JSON is about 1.3 times larger than the `.xpt`
for both domains.** That is the opposite of the advertised result, and
it is not a mistake in the build.

Now the same comparison on CDISC's own published example files, shipped
in the specification repository:

```{r ex3-cdisc}
ex <- "refs/DatasetJSON_v1.1/repo/examples/adam"

tibble(dataset = c("adae", "adlbc")) |>
  mutate(xpt  = file.size(file.path(ex, paste0(dataset, ".xpt"))),
         json = file.size(file.path(ex, paste0(dataset, ".json"))),
         dsjc = file.size(file.path(ex, paste0(dataset, ".dsjc"))),
         json_vs_xpt = round(json / xpt, 3),
         dsjc_vs_xpt = round(dsjc / xpt, 3)) |>
  kable()
```

Theirs go the other way: JSON is **smaller** than the transport file.
Same two formats, opposite conclusions. One of these is misleading and
it is worth working out which.

### Why the answer differs

The difference is not the formats. It is the **declared lengths**.

```{r ex3-lengths}
our_len   <- fromJSON("data/json/lb.json", simplifyVector = FALSE)$columns |>
  map_int(~ .x$length %||% NA_integer_)
cdisc_len <- fromJSON(file.path("refs/DatasetJSON_v1.1/repo/examples/sdtm/lb.json"),
                      simplifyVector = FALSE)$columns |>
  map_int(~ .x$length %||% NA_integer_)

list(GLPX1 = our_len, CDISC = cdisc_len) |>
  map(~ c(n = length(.x), max = max(.x, na.rm = TRUE),
          at_200 = sum(.x == 200, na.rm = TRUE)))
```

CDISC's LB declares three fields at the transport maximum of 200
characters. Ours declares nothing above 24, because
`R/author_spec_workbook.R` computed every length from the data actually
present.

A transport file allocates the declared width on **every row**, whether
the data fills it or not. Three 200-character fields across thousands of
rows is exactly the "empty space for columns allocated, but not used"
the PhUSE paper describes. Dataset-JSON writes only what is there, so
the padding is what it wins on, and where there is no padding, it
loses, because JSON pays for quotes, commas and brackets.

Which of the two is realistic? Theirs. The Define-XML specification says
so directly: variable lengths are **planned** lengths, not observed
ones. A real study declares `LBORRES` at 200 because some future result
might need it. This course computed lengths from 12,388 rows that
already exist, which no live study can do.

So the honest statement of the central lesson:

> **The storage-efficiency claim depends on how generously you declared
> your lengths, not on the format.** Dataset-JSON does not compress your
> data. It stops transport format from padding it. If you had nothing to
> pad, you get nothing back.

### The compressed form

One number in both tables is not ambiguous:

`.dsjc`, compressed Dataset-JSON, is **0.10 to 0.14** of the
transport file in our data and **0.07** in CDISC's. Seven to ten times
smaller, in every case, regardless of how the lengths were declared.
That is the size argument, and it does not depend on anyone's
declarations.

## What Can Go Wrong

**Reading task, about 10 minutes.** No code. Mostly about assuming a new
format removes an old obligation.

*Estimated time: ~10 minutes (reading)*

- **Assuming the format preserves your distinctions.** Example 2 is the
  case in point. A pipeline that is careful about null-versus-`"N"` all
  the way through ADaM, and then submits transport format, has thrown
  the distinction away at the last step for every character variable.
- **Quoting the storage claim without the caveat.** "Dataset-JSON is
  smaller than XPT" is true of realistic data and false of ours. If you
  cite it, cite what it depends on.
- **Expecting one file to hold a study.** "Dataset-JSON must contain
  only 1 dataset per file." A submission is a directory, not a document.
- **Treating `itemGroupOID` as decoration.** It is Required. If it does
  not match an `ItemGroupDef` in your define.xml, the two artifacts
  disagree and nothing in either file will tell you.
- **Assuming eight-character names are a CDISC style choice.** They are
  a 1989 file-format constraint. Worth knowing, because it explains a
  great deal of what otherwise looks like arbitrary ugliness.

### Common misinterpretations

- **"JSON means human-readable."** A 12,388-row LB file is two megabytes
  of nested arrays. It is *parseable*, which is different.
- **"Dataset-JSON replaces define.xml."** It optionally references one.
  `columns` carries name, label, type and length; define.xml carries
  origin, derivations, codelists and value-level metadata.
- **"The rows are objects."** They are arrays, which is precisely why
  the format is not far larger than transport.

## Discussion: Does a Better Format Make It Submittable?

**Discussion, about 10 minutes.** Alone, write your answer down; in a
session, this is the last argument of the course.

Dataset-JSON will happily store a variable called
`LAB_RESULT_NUMERIC`. Transport v5 will not.

1. **One minute, on your own.** Is a dataset with that variable name
   submittable as SDTM? Yes or no, and name the document that settles
   it.
2. **Five minutes, together.** Somebody will argue the eight-character
   limit was only ever a transport limitation and should now go away.
   Argue it out. It is a better argument than it first appears.
3. **The rest.** Which other conventions you have learned in this course
   exist only because of the file format, and which would you keep even
   if the format disappeared tomorrow?

The course ends where it started: a format can carry anything, and
agreeing what the contents mean is a separate job that never goes away.

## Exercises

**Exercise, about 15 minutes.** The last exercise set in the course.

[Exercise 11, Dataset-JSON](../exercises/dataset-json.qmd) asks you to
break the eight-character rule and find out what happens, and to settle
the size question on data where the lengths are planned rather than
observed.

## Comprehension Check

**Self-test, about 10 minutes.** Answer before expanding.

::: {.callout-note collapse="true"}
### Questions

1. Why is the ADaM variable for treatment-emergent flag called
   `TRTEMFL` rather than `TREATMENT_EMERGENT_FLAG`?
2. `GLPX1-103-199` has a null `TRTEMFL`. What happens to that null in a
   `.xpt` submission, and why does it matter?
3. A colleague says Dataset-JSON files are smaller than XPT. Under what
   condition is that false?
4. `metaDataRef` is optional but `itemGroupOID` is required. What does
   that combination let a file do, and not do?
5. Why are `rows` arrays rather than objects?
:::

::: {.callout-note collapse="true"}
### Answers

1. **Because SAS Transport v5 limits variable names to eight
   characters.** Every CDISC variable name in this course obeys it,
   131 of 131. It is a file-format constraint that predates SDTM, not a
   naming convention someone chose.

2. **It becomes an empty string, indistinguishable from a deliberate
   blank.** Demonstrated in Example 2: `is.na()` returns `TRUE` before
   the round-trip and `FALSE` after. It matters because
   [ADAE](adae.qmd) established that null means "we do not know" while
   `"N"` means "no", and the transport file cannot carry the
   difference.

3. **When the declared lengths are close to the data.** GLPX-1's
   Dataset-JSON is 1.3× *larger* than its `.xpt`, because the
   specification workbook computed lengths from observed data. CDISC's
   examples declare fields at 200 characters, so their transport files
   carry padding that the JSON does not, and the JSON wins. Compressed
   `.dsjc` is smaller in both cases.

4. **It can be joined to its metadata, but it need not name it.**
   `itemGroupOID` is the foreign key to an `ItemGroupDef` in a
   define.xml; `metaDataRef` is the URI of that document. Required key,
   optional pointer, so a file always *can* be matched to metadata,
   but a recipient may have to be told which define.xml to match it
   against.

5. **To avoid repeating the column names on every record.** Names are
   stated once in `columns`. On LB that saves writing 22 keys across
   12,388 rows. It is the single design decision that keeps
   Dataset-JSON competitive with a binary format on size.
:::
