---
title: "Define-XML from a Specification Workbook"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(openxlsx)
library(xml2)
library(defineR)
library(metacore)

dx <- "refs/DefineV2111_0"
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 105 minutes.**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background: the workbook is the deliverable | 15 min | Reading |
| Example 1: Reading a real specification | 20 min | Worked example |
| Example 2: Generating define.xml from it | 15 min | Worked example |
| Example 3: What the tool could not say | 15 min | Worked example |
| Example 4: Reading the metadata back | 10 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Exercises + Comprehension Check | 15 min | Practice / Self-test |

The other half of [Define-XML](define-xml.qmd). That session built the
document by hand to show its structure; this one builds it the way a
sponsor does, from a specification workbook, and examines what the
production route costs you. Read them in order. This session assumes
you know what an `ItemGroupDef` is.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Run R from the project root, as always. This session uses three
packages you have not seen before: **`defineR`**, **`openxlsx`** and
**`metacore`**, all pinned in `renv.lock`.

Everything here runs against artifacts committed in this repository -
`data/spec/SDTM_METADATA.xlsx` and `data/define/`. You do not need the
CDISC Define-XML package for the main path; one comparison in Example 3
uses it, and is marked.
:::

::: {.callout-warning}
### A Note on `defineR`

`defineR` is at version **0.0.6**. That leading zero is not decoration:
it is a pre-1.0 package, and its interface may change. It is used here
because it is the most direct open-source path in R from a
specification workbook to a define.xml, and because the shape of what
it does is what matters, not because it is settled infrastructure you
should build a submission pipeline on without further thought.
:::

## When Do You Use This?

::: {.callout-tip}
[Define-XML](define-xml.qmd) built a document with three variables. A
real submission has thousands. Then:

- Nobody is going to hand-write `xml_add_child()` calls for 2,000
  variables across 30 domains. What actually produces the file?
- Your standards group maintains dataset metadata in a spreadsheet, and
  has done for years, before anyone mentioned XML. How does that become
  a submission deliverable?
- The tool you use was written against an older version of the
  standard. What exactly does that cost you, and how would you find
  out?

**How is define.xml actually produced, and what does the production
route quietly decide on your behalf?**
:::

## Learning Objectives

After completing this session you will be able to:

- Describe what a specification workbook contains and why it, not the
  XML, is the sponsor's source of truth
- Generate a Define-XML document from a workbook with `defineR`
- Identify what a Define-XML 2.0 tool cannot express, by comparing its
  output against the 2.1 enumerations
- Read published metadata back into R with `metacore` and explain why
  that round-trip is worth doing
- Explain why "our tool produces a valid define.xml" is not the same
  claim as "our define.xml is right"

## Background: The Workbook Is the Deliverable

**Reading task, about 15 minutes.** This session changes who you are.
In the last one you wrote the XML; here somebody hands you a
spreadsheet and the XML is generated. Read before running.

*Estimated time: ~15 minutes (reading)*

Here is the thing that surprises people coming to clinical programming
from other data work: **in most sponsors, nobody writes define.xml.**

The document that people write, argue about, review, version and sign
off is a **specification workbook**: an Excel file, one sheet per kind
of metadata, maintained by a standards or data-management group. It
lists every dataset, every variable, its type, its length, its label,
its origin, its codelist and its derivation. It is a controlled
document. It exists before any of the datasets do, because it is what
programming is written *against*.

define.xml is generated from it, usually near the end, often
automatically, and frequently by someone who did not write a line of
the metadata it contains.

That arrangement has a consequence worth stating plainly, because it
governs everything else in this session:

> **If the define.xml is wrong, the workbook is wrong. Fixing the XML
> fixes one submission; fixing the workbook fixes the study.**

This repository mirrors that arrangement deliberately. `data/spec/SDTM_METADATA.xlsx`
is a committed artifact: a binary file that git cannot show you diffs
of, which is a genuine cost and also genuinely how this works. `R/build_define.R`
**reads** that workbook and never writes it. There is a second script,
`R/author_spec_workbook.R`, which created the workbook once and stands
in for the standards group this course does not have.

## Example 1: Reading a Real Specification

**Worked example, about 20 minutes.** Look at the workbook properly.
The sheet names and column names are the vocabulary for the whole
session.

*Estimated time: ~20 minutes (worked example)*

GLPX-1's workbook covers two domains, DM and LB. Start with its
shape:

```{r ex1-sheets}
spec <- "data/spec/SDTM_METADATA.xlsx"

tibble(sheet = getSheetNames(spec)) |>
  mutate(rows = map_int(sheet, ~ nrow(read.xlsx(spec, sheet = .x))))
```

Nine sheets. Read them as a list of definitions rather than a
spreadsheet and the structure of [Define-XML](define-xml.qmd) is
already visible: `TOC_METADATA` becomes `ItemGroupDef`,
`VARIABLE_METADATA` becomes `ItemDef`, `CODELISTS` becomes `CodeList`,
`COMPUTATION_METHOD` becomes `MethodDef`, `WHERE_CLAUSES` becomes
`def:WhereClauseDef`. The workbook is the same web of definitions,
flattened into sheets.

### The variable sheet

This is where almost all the content lives:

```{r ex1-vars}
vars <- read.xlsx(spec, sheet = "VARIABLE_METADATA") |> as_tibble()

vars |>
  filter(DOMAIN == "LB") |>
  select(VARNUM, VARIABLE, TYPE, LENGTH, LABEL, ORIGIN, CODELISTNAME) |>
  head(10)
```

### Run It Yourself

Count what the specification actually commits to:

```{r ex1-counts}
vars |> count(DOMAIN, name = "variables")
vars |> count(ORIGIN, sort = TRUE)
```

### Reading the Output Line by Line

48 variables across two domains, and every one of them carries an
origin. Look at the origin counts closely, because Example 3 comes back
to them: `Derived`, `Assigned`, `CRF`, `eDT`.

Two of those four are about to become a problem.

### Where the labels came from

One detail matters more than it looks. The labels in that sheet were
not typed from memory. `R/author_spec_workbook.R` takes 36 of them from
the CDISC-published example define.xml shipped in the Define-XML v2.1
package, and the remaining 12 from the SDTMIG v3.4 specification tables
(§5.2 for DM, §6.3.5.6 for LB). The script **stops with an error** if
any variable is left unlabelled rather than proceeding with a blank.

That is not fussiness. A wrong variable label is one of the easiest
things in a submission to get wrong and one of the hardest for anyone
downstream to notice, because it looks like content rather than
metadata. It is worth a guard.

## Example 2: Generating define.xml From It

**Worked example, about 15 minutes.** One command does the work. The
interesting part is what it validates against.

*Estimated time: ~15 minutes (worked example)*

`defineR::write_define()` reads the workbook and writes the document.
That is the whole production step:

```{r ex2-build}
#| eval: false
write_define(
  path  = "data/define/SDTM_METADATA.xlsx",
  dir   = "data/define",
  type  = "sdtm",
  check = TRUE,
  html  = TRUE,
  view  = FALSE
)
```

Note the `path`. `write_define()` writes its output **beside the
workbook you hand it**, so `R/build_define.R` copies
`data/spec/SDTM_METADATA.xlsx` into `data/define/` first, runs the
generation, then deletes the copy. That is why the path above points at
a file you will not find on disk: it exists only while the build runs.
Hand it the workbook where it lives and your define.xml lands in
`data/spec/` beside it.

`R/build_define.R` runs exactly that, and the outputs are committed, so
you can inspect them without rebuilding:

```{r ex2-outputs}
list.files("data/define")
```

Three files from one workbook:

| File | What it is |
|---|---|
| `define.sdtm.xml` | The submission deliverable |
| `define.sdtm.html` | The browsable rendering a reviewer opens |
| `check.sdtm.pdf` | `defineR`'s own check report |

The HTML is worth dwelling on. A reviewer does not read raw XML. They
open define.xml in a browser, where a stylesheet turns it into a
navigable document. That rendering is not a courtesy the tool added; it
is produced by an XSL stylesheet CDISC publishes, and the reference to
it is a line inside the XML itself.

### Run It Yourself

```{r ex2-inspect}
doc <- read_xml("data/define/define.sdtm.xml")

c(datasets  = length(xml_find_all(doc, "//d1:ItemGroupDef")),
  variables = length(xml_find_all(doc, "//d1:ItemDef")),
  codelists = length(xml_find_all(doc, "//d1:CodeList")))
```

### Reading the Output Line by Line

Two datasets and six codelists match the workbook exactly. **52
`ItemDef`s from 48 variables** does not, and the extra four are not a
bug.

They are the value-level metadata. `VALUELEVEL_METADATA` has four rows,
one per lab analyte, because `LBSTRESN` means something different
depending on `LBTESTCD`: a percentage for HbA1c, mmol/L for glucose,
U/L for ALT, µmol/L for creatinine. One variable, four meanings, and
Define-XML expresses that by defining four additional items and
attaching a where-clause to each.

Four rows in a spreadsheet became four `ItemDef`s, four `ItemRef`s and
four `def:WhereClauseDef`s. That is the leverage the workbook route
buys you.

## Example 3: What the Tool Could Not Say

**Worked example, about 15 minutes.** The heart of the session. Read
every output here rather than skimming to the conclusion.

*Estimated time: ~15 minutes (worked example)*

Now the part that matters.

```{r ex3-validate}
schema_20 <- read_xml(system.file(
  "extdata/2.0.0/cdisc-define-2.0/define2-0-0.xsd", package = "defineR"))

as.logical(xml_validate(doc, schema_20))
```

Valid. And against the current standard:

```{r ex3-validate21}
schema_21 <- read_xml(file.path(dx, "schema/cdisc-define-2.1/define2-1-0.xsd"))

as.logical(xml_validate(doc, schema_21))
```

`defineR` emits **Define-XML 2.0**, whose specification carries the
production date 2013-03-05 in its own revision history. The current
standard is 2.1. It is tempting to file that as a version lag: an old
tool that will catch up. Look at what actually differs before accepting
that reading.

### Origin

```{r ex3-origin}
xml_find_all(doc, "//def:Origin") |>
  xml_attr("Type") |>
  table()
```

Note first that `Derived` is 23 here against 19 in the workbook: the
four value-level items carry origins of their own. The two tables
reconcile.

Two things are wrong here by 2.1's account, and only one of them is the
obvious one.

The obvious one: **there is no `Source` attribute.** [Define-XML](define-xml.qmd)
established that SDTM requires both `Type` and `Source`, and that a
central lab result is `Collected` + `Vendor`. The workbook has a single
`ORIGIN` column. There is nowhere to put "Vendor".

The less obvious one is worse. Compare those emitted types against what
2.1 actually permits:

```{r ex3-enum}
enum <- read_xml(file.path(dx, "schema/cdisc-define-2.1/define-enumerations.xsd"))
ns   <- c(xs = "http://www.w3.org/2001/XMLSchema")

permitted <- xml_attr(xml_find_all(
  enum, "//xs:simpleType[@name='OriginType']//xs:enumeration", ns), "value")

emitted <- unique(xml_attr(xml_find_all(doc, "//def:Origin"), "Type"))

setdiff(emitted, permitted)
```

**`CRF` and `eDT` are not values that 2.1 permits.** They are not
missing a `Source`. They do not exist. 2.1 did not extend origin; it
reorganised it. The old types were re-expressed as pairs:

| Define-XML 2.0 | Define-XML 2.1 |
|---|---|
| `CRF` | `Collected` + `Investigator` |
| `eDT` | `Collected` + `Vendor` |

That second row is exactly the pairing the previous session derived for
GLPX-1's central lab results from §4.3.2.1. The two sessions meet
here.

So the central intuition of this session is not about version numbers:

> **The tool is not describing the same thing the standard now asks
> for.** A 2.0 document is not a 2.1 document missing an attribute. It
> is a document whose vocabulary was retired.

### Class

The same pattern, in a second place, and the difference here is
structural rather than a matter of spelling. Look for `def:Class` as an
element, the way [Define-XML](define-xml.qmd) built it:

```{r ex3-class-elem}
length(xml_find_all(doc, "//def:Class"))
```

None. It is not missing, in 2.0 the dataset class is an **attribute**
on `ItemGroupDef`:

```{r ex3-class-attr}
xml_attr(xml_find_all(doc, "//d1:ItemGroupDef"), "Class")
```

Whereas in the CDISC-published 2.1 example, the same information is a
**child element** on all eleven of its datasets:

```{r ex3-class-21}
ex21 <- read_xml(file.path(dx, "examples/DefineXML-2-1-SDTM/defineV21-SDTM.xml"))

c(as_element   = length(xml_find_all(ex21, "//def:Class")),
  as_attribute = sum(!is.na(xml_attr(xml_find_all(ex21, "//d1:ItemGroupDef"),
                                     "Class"))))
```

So the class moved from an attribute to an element between versions,
and the values changed casing too, from `Findings` to the `FINDINGS`
that 2.1's `ItemGroupClass` enumeration requires. The previous session
demonstrated deliberately that a lowercased class name fails schema
validation with a `[facet 'enumeration']` error. Here the same failure
arrives from a tool rather than a typo, and nothing in the workbook
would have warned you.

### Namespace

```{r ex3-ns}
xml_ns(doc)
```

The declared extension namespace ends `v2.0`. This is why the 2.1
validation returned `FALSE` immediately: the document is not claiming
to be a 2.1 document at all. Everything above (the retired origin
types, the relocated class) follows from that one declaration.

## Example 4: Reading the Metadata Back

**Worked example, about 10 minutes.** Short round trip, to show the
metadata is machine-readable in both directions.

*Estimated time: ~10 minutes (worked example)*

A define.xml is machine-readable, which means the trip runs both ways.
`metacore` parses a published define.xml into an R object:

```{r ex4-roundtrip}
mc <- define_to_metacore("data/define/define.sdtm.xml", verbose = "silent")

c(datasets    = nrow(mc$ds_spec),
  variables   = nrow(mc$ds_vars),
  codelists   = nrow(mc$codelist),
  derivations = nrow(mc$derivations))
```

Everything the workbook committed comes back out:

```{r ex4-ds}
mc$ds_spec
```

And a codelist can be recovered by name, from the XML, without the
workbook present at all:

```{r ex4-ct}
get_control_term(mc, LBNRIND)
```

### Why this matters

Consider who can run that code. Not just the sponsor - **anyone holding
the submission**. A reviewer receiving define.xml can reconstruct the
sponsor's entire declared metadata programmatically and check the
datasets against it.

That is what "machine-readable" bought, and it cuts both ways. The
metadata is not documentation attached to the submission; it is a set
of claims the submission can be tested against, by someone who does not
trust you and does not have to ask.

## What Can Go Wrong

**Reading task, about 10 minutes.** No code. Mostly about trusting a
generator that is a version behind the standard.

*Estimated time: ~10 minutes (reading)*

**The tool validates against the wrong schema.** `defineR`'s own
`check = TRUE` produces `check.sdtm.pdf`, and the document validates
cleanly, against 2.0. A green check from a tool tells you the tool is
satisfied. `R/build_define.R` therefore asserts *both* schemas
explicitly, and is written to complain if the 2.1 result ever changes,
so that a future `defineR` gaining 2.1 support is noticed rather than
silently assumed.

**The workbook has no place for what you know.** This is the failure
mode with no error message. You know GLPX-1's lab results come from a
central laboratory. The 2.0 `ORIGIN` column cannot record it. Nothing
warns you, nothing fails, and the information is simply absent from the
deliverable. **A metadata format's limits become your metadata's
limits, silently.**

**Fixing the XML instead of the workbook.** Hand-editing generated
define.xml to correct something is the single most tempting mistake
here, and it does not survive the next regeneration. It also makes the
workbook and the deliverable disagree, which is worse than either being
wrong on its own.

**Links to documents that do not exist.** `EXTERNAL_LINKS` in this
workbook names `blankcrf.pdf` and `sdrg.pdf`. **Neither file is part of
this course**, and the generated define.xml links to both. A real
submission ships them alongside; here they are named and absent, which
is exactly the kind of dangling reference a conformance check is
supposed to catch and a schema will not.

### Common misinterpretations

- **"The define.xml is the source of truth."** The workbook is. The
  XML is generated output, and treating it as the master is how a study
  ends up with two disagreeing versions of its own metadata.
- **"It validated, so the version is fine."** It validated against the
  version it was written for. That is close to a tautology.
- **"2.0 is 2.1 minus some attributes."** `CRF` and `eDT` say
  otherwise. The vocabulary changed, not just the field count.
- **"Value-level metadata is an advanced feature."** Four rows in a
  spreadsheet. It is ordinary, and any Findings domain with mixed units
  needs it.

## Exercises

**Exercise, about 15 minutes.** Do these before the next session.

[Exercise 9: Define-XML from a Specification Workbook](../exercises/define-from-specs.qmd)
adds a domain to the workbook, regenerates, and asks what it would take
to close the gap to 2.1.

## Comprehension Check

**Self-test, about 10 minutes.** Answer before expanding.

::: {.callout-note collapse="true"}
### Questions

1. A reviewer finds `LBSTRESN`'s label is wrong in your submitted
   define.xml. Where do you fix it, and why not in the XML?
2. The workbook has 48 variables but the generated define.xml has 52
   `ItemDef`s. Account for the difference.
3. Why is `Type="eDT"` a more serious problem than `def:Origin` simply
   lacking a `Source` attribute?
4. Your `defineR` output validates and your check report is clean. What
   have you demonstrated, and what have you not?
5. What can a regulatory reviewer do with your define.xml that they
   could not do with a PDF describing the same datasets?
:::

::: {.callout-note collapse="true"}
### Answers

1. **In `data/spec/SDTM_METADATA.xlsx`, then regenerate.** Editing the
   XML fixes one file and leaves the workbook (the source of truth,
   and the thing the next regeneration reads) still wrong. It also
   makes the two disagree, so the next person cannot tell which is
   intended.

2. **Value-level metadata: four extra `ItemDef`s, one per lab analyte.**
   `LBSTRESN` has a different meaning and unit depending on `LBTESTCD`
   (% for HbA1c, mmol/L for glucose, U/L for ALT, µmol/L for
   creatinine), so Define-XML defines an item per value and attaches a
   where-clause to each.

3. **Because a missing `Source` is an omission, and `eDT` is a
   retirement.** Adding `Source` to a 2.0 document would leave `eDT`
   still invalid. It is not in 2.1's `OriginType` enumeration at all.
   The conversion is `eDT` → `Collected` + `Vendor`: one value becomes
   a pair, and you cannot get there by adding an attribute.

4. **You have demonstrated that the document conforms to Define-XML
   2.0.** You have not demonstrated conformance to the current
   standard, nor to the 90 Specification-sourced conformance rules that
   no schema check touches (see [Define-XML](define-xml.qmd)), nor that
   the metadata is *correct*: a confidently wrong label validates
   perfectly.

5. **Check it programmatically.** `define_to_metacore()` turns the
   submission's declared metadata back into data, so a reviewer can
   test the datasets against the sponsor's own claims (codelists,
   types, lengths, derivations) without asking the sponsor anything.
   A PDF is readable; only this is checkable.
:::
