---
title: "Define-XML"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(xml2)

# Local path to the unzipped CDISC Define-XML v2.1 package. See the
# callout below: this is reference material the reader downloads, not
# something the repository can redistribute.
dx <- "refs/DefineV2111_0"
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 105 minutes.**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background: a web held together by arbitrary names | 20 min | Reading |
| Example 1: A define.xml for DM, built by hand | 25 min | Worked example |
| Example 2: Origin grows a second attribute | 15 min | Worked example |
| Example 3: Declaring the standard, and a decision made three sessions ago | 15 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Exercises + Comprehension Check | 15 min | Practice / Self-test |

The session that closes Part 2. Everything built in
[DM](sdtm-dm.qmd), [Events](sdtm-events.qmd) and
[Findings](sdtm-findings.qmd) is a dataset; this session builds the
document that *describes* those datasets, and it is the first time the
course writes something that is not a data frame. Split after Example 1
for self-paced study.
:::

::: {.callout-important}
### You Need the Define-XML Package Before You Start

Unlike every earlier session, this one reads files that this repository
cannot ship. The Define-XML v2.1 package is copyright CDISC and
licensed to you, not redistributable by a course.

Download it yourself. It is free: a CDISC account, not a paid
membership:

- **Define-XML v2.1** -
  <https://www.cdisc.org/standards/data-exchange/define-xml/define-xml-v2-1>

Unzip it into a `refs/` folder at the project root, so that the
schema sits at `refs/DefineV2111_0/schema/cdisc-define-2.1/`. Every
chunk below assumes:

```{r}
#| eval: false
dx <- "refs/DefineV2111_0"
```

Adjust that one line if you put it somewhere else. `refs/` is
gitignored in this repository for exactly this reason.

The package used to write this session is **v2.1.11**, whose `README.md`
is dated 2026-03-27. The specification PDF inside it is **Version 2.1
(Final), 2019-05-15**: the point releases update the schemas, not the
prose. That gap between a frozen document and a maintained schema turns
out to matter, and Example 3 is where it bites.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown.
Where you see a "Run It Yourself" box, stop and run the code above it
before reading the explanation that follows. Run R from the project
root, as in earlier sessions.

The only package this session needs beyond the tidyverse is
**`xml2`**, which you already have. It is a tidyverse dependency. We
build the XML by hand, element by element. That is slower than the
tools a sponsor would actually use, and it is the point: you cannot
review a define.xml you have never assembled.
:::

## When Do You Use This?

::: {.callout-tip}
You have built DM, AE, LB and VS. Then:

- A reviewer opens `lb.xpt` and sees a column called `LBSTRESN`. Nothing
  in the file says what it is, where it came from, or what its units
  mean. Where do they look?
- Your SAP defines HbA1c baseline one way; a second study defined it
  another. Both submissions contain a variable called `BASE`. How does
  anyone downstream tell them apart?
- You coded `SEX` against a CDISC codelist in
  [SDTM Concepts](sdtm-concepts.qmd). How does a reviewer *verify* that,
  without taking your word for it?

**Where does the meaning of a submitted dataset live, if not in the
dataset?**
:::

## Learning Objectives

After completing this session you will be able to:

- Explain what a define.xml document is for, and why a submission is
  incomplete without one
- Describe how `Def` and `Ref` elements and their OIDs turn a flat list
  of definitions into a linked document
- Build a schema-valid Define-XML v2.1 document for an SDTM domain
  using `xml2`, from `ODM` down to `CodeList`
- Populate `def:Origin` with both its `Type` and `Source` attributes,
  and justify the combination for a collected lab result
- Validate a document against the CDISC schema in R, and explain why a
  document that validates can still be wrong

## Background: A Web Held Together by Arbitrary Names

**Reading task, about 20 minutes.** The central idea here is that the
names holding the document together mean nothing. It is strange, and it
is worth sitting with before you build anything.

*Estimated time: ~20 minutes (reading)*

Everything Part 2 has produced so far is a table. `dm.csv` has 400 rows
and 27 columns; `lb.csv` has rows of results with units and reference
ranges. A regulatory reviewer receiving those tables would know the
column names and nothing else. Not what `LBNRIND` means. Not whether
`AGE` was collected or derived. Not which version of the SDTMIG the
sponsor was implementing when they chose those columns.

**define.xml is where all of that lives.** It is a single XML document
submitted alongside the datasets, and it carries the metadata: every
dataset, every variable, every codelist, every derivation, and (the
part this session keeps coming back to) where each value came from.

The specification is blunt about the purpose:

> The purpose of Define-XML is to support the interchange of dataset
> metadata for clinical research applications in a machine-readable
> format (Define-XML v2.1, §4)

**Machine-readable** is doing real work in that sentence. A PDF
describing your datasets would be readable. It would not be *checkable*.
Everything this session does in its last third (validating against a
schema, running conformance rules) is possible only because the format
was designed for a program to read first and a person second.

### Defs and Refs

The structure has one organising idea, and once you see it the rest of
the document stops looking arbitrary:

> In Define-XML, an element whose name ends with "Def" is the
> declaration of an object instance. An element whose name ends with
> "Ref" is a reference to that object from some other entity
> (Define-XML v2.1, §3.5)

So a variable is *declared once*, in an `ItemDef`. A dataset that
contains that variable does not repeat the declaration. It points at
it, with an `ItemRef`. The link is carried by an identifier called an
**OID**.

This is why `STUDYID` is not defined twenty times in a submission with
twenty domains. It is defined once, and every `ItemGroupDef` references
the same `ItemDef`:

> By defining STUDYID outside of any dataset context, we can define it
> only once and reference the definition from every dataset that
> contains the variable. In this sense, the STUDYID definition is being
> shared or reused by all the datasets (Define-XML v2.1, §3.5.1)

### The central intuition

Here is the part that surprises people, and it is worth sitting with
before writing any code.

**The OIDs mean nothing.**

> This Define-XML specification does not mandate a format for OIDs. OIDs
> are only intended as a mechanism for unambiguously linking between a
> definition of an object and references to it. The examples in this
> document use prefixes at the start of OIDs to indicate the object
> type. However, this is not required; it is equally valid to use
> randomly generated identifiers (Define-XML v2.1, §3.5.1)

And, in case that was not clear enough:

> The value of an OID attribute has no intrinsic meaning (Define-XML
> v2.1, §3.5.1)

Every define.xml you will ever open uses OIDs like `IT.DM.USUBJID` and
`IG.LB` and `CL.SEX`. It looks like a naming standard. It is not one. It
is a convention that implementers adopted because it makes the file
readable to humans, and the standard permits it the same way it would
permit `a7f3b91c`.

So the intuition to carry through the rest of this session:

> **A define.xml is a web of definitions held together by names that are
> deliberately meaningless. The structure lives entirely in which name
> points at which, not in what the names say.**

That has a practical consequence you will hit in Example 1 and again in
"What Can Go Wrong": if an `ItemRef` points at an OID that no `ItemDef`
declares, nothing about the *name* will tell you. The link is either
there or it is not, and only checking finds out.

## Example 1: A define.xml for DM, Built by Hand

**Worked example, about 25 minutes.** You build a schema-valid document
element by element. Run each chunk as you reach it; the order the
elements go in is load-bearing, and skipping ahead breaks it.

*Estimated time: ~25 minutes (worked example)*

We will build a small but genuinely valid Define-XML v2.1 document
describing GLPX-1's DM domain, with three variables and one codelist.
Small, but not a toy: it validates against the same CDISC schema a
submission would be checked with.

The document nests like this, outermost first:

| Element | What it is |
|---|---|
| `ODM` | The file itself. Define-XML is an extension of CDISC ODM. |
| `Study` | The study being described |
| `GlobalVariables` | Study name, description, protocol name |
| `MetaDataVersion` | The container for everything below |
| `def:Standards` | Which published standards this submission follows |
| `ItemGroupDef` | A dataset (here, DM) |
| `ItemDef` | A variable |
| `CodeList` | A set of permitted values |

Note where `ItemDef` sits: **beside** `ItemGroupDef`, not inside it.
That is the Def/Ref idea made concrete: the dataset holds `ItemRef`
pointers, and the definitions live at the same level as the dataset,
available to be shared.

### The document shell

`xml2` builds a tree by adding children to nodes. `xml_new_root()`
creates the outermost element, and the three `xmlns` attributes declare
the namespaces: the ODM one for standard elements, the `def` one for
the Define-XML extensions, and `xlink` for links to external files.

```{r ex1-shell}
NS_ODM <- "http://www.cdisc.org/ns/odm/v1.3"
NS_DEF <- "http://www.cdisc.org/ns/def/v2.1"

odm <- xml_new_root(
  "ODM",
  "xmlns"       = NS_ODM,
  "xmlns:def"   = NS_DEF,
  "xmlns:xlink" = "http://www.w3.org/1999/xlink",
  ODMVersion       = "1.3.2",
  FileType         = "Snapshot",
  FileOID          = "GLPX1.define.sdtm.v1",
  CreationDateTime = "2026-09-04T10:00:00",
  Originator       = "CDISC with R course",
  "def:Context"    = "Other"
)

study <- xml_add_child(odm, "Study", OID = "STDY.GLPX1")
gv    <- xml_add_child(study, "GlobalVariables")
xml_add_child(gv, "StudyName", "GLPX-1")
xml_add_child(gv, "StudyDescription",
              "A simulated Phase III trial of GLPX 10 mg in type 2 diabetes")
xml_add_child(gv, "ProtocolName", "GLPX-1")

mdv <- xml_add_child(
  study, "MetaDataVersion",
  OID  = "MDV.GLPX1.SDTMIG.3.4",
  Name = "GLPX-1 SDTM Data Definitions",
  "def:DefineVersion" = "2.1.11"
)
```

Two attributes are worth pausing on. `FileType="Snapshot"` says this
document describes data as of a moment, rather than a set of changes,
which is what a submission always is. And `def:DefineVersion` is the
document declaring which version of Define-XML *it* is written to; not
which version of SDTM it describes. Those are different things, and the
next chunk handles the other one.

### Declaring which standards you followed

```{r ex1-standards}
std <- xml_add_child(mdv, "def:Standards")

xml_add_child(std, "def:Standard", OID = "STD.1",
              Name = "SDTMIG", Type = "IG",
              Version = "3.4", Status = "Final")

xml_add_child(std, "def:Standard", OID = "STD.2",
              Name = "CDISC/NCI", Type = "CT",
              PublishingSet = "SDTM",
              Version = "2023-12-15", Status = "Final")
```

This is the machine-readable version of a claim the course has been
making in prose since [SDTM Concepts](sdtm-concepts.qmd): GLPX-1
implements **SDTMIG v3.4** with SDTM CT. Until now that claim lived in a
sentence. Here it becomes an attribute a validator can check the rest of
the document against.

`def:Standards` is new in v2.1, and it replaced something:

> Note: The def:Standards and def:Standard elements replace the
> def:StandardName and def:StandardVersion attributes used in Define-XML
> Version 2.0 (Define-XML v2.1, §4.1)

The reason for the change is visible in the shape: 2.0 had *attributes*,
so a document could name one standard. 2.1 has *elements*, so a document
can list several, which a real submission needs when some domains were
built to one SDTMIG version and some to another.

### The dataset, and pointers to its variables

```{r ex1-itemgroup}
igd <- xml_add_child(
  mdv, "ItemGroupDef",
  OID = "IG.DM", Name = "DM", Domain = "DM",
  Repeating = "No", IsReferenceData = "No",
  SASDatasetName = "DM", Purpose = "Tabulation",
  "def:Structure"          = "One record per subject",
  "def:StandardOID"        = "STD.1",
  "def:ArchiveLocationID"  = "LF.DM"
)

d <- xml_add_child(igd, "Description")
xml_add_child(d, "TranslatedText", "Demographics", "xml:lang" = "en")

xml_add_child(igd, "ItemRef", ItemOID = "IT.STUDYID",
              Mandatory = "Yes", OrderNumber = "1", KeySequence = "1")
xml_add_child(igd, "ItemRef", ItemOID = "IT.DM.USUBJID",
              Mandatory = "Yes", OrderNumber = "2", KeySequence = "2")
xml_add_child(igd, "ItemRef", ItemOID = "IT.DM.SEX",
              Mandatory = "Yes", OrderNumber = "3")

xml_add_child(igd, "def:Class", Name = "SPECIAL PURPOSE")

lf <- xml_add_child(igd, "def:leaf", ID = "LF.DM", "xlink:href" = "dm.xpt")
xml_add_child(lf, "def:title", "dm.xpt")
```

Read the `ItemRef` lines and notice what is *not* there: no label, no
data type, no length, no origin. An `ItemRef` says "this dataset
contains that variable, in this position, and it is a key": nothing
about what the variable *is*. `def:StandardOID="STD.1"` points back at
the `def:Standard` declared a moment ago, so this dataset is explicitly
an SDTMIG v3.4 dataset.

`def:Class="SPECIAL PURPOSE"` is DM's SDTM class, the same classification
[SDTM Concepts](sdtm-concepts.qmd) introduced. Keep an eye on that
element, Example 3 comes back to it.

`def:leaf` is the link to the actual data file. This is the line that
connects the metadata document to `dm.xpt` sitting beside it in the
submission folder.

### The variable definitions

Three variables, and a small helper so the pattern is visible rather
than buried in repetition:

```{r ex1-items}
add_item <- function(parent, oid, name, type, len, label,
                     origin_type, origin_source = NULL, codelist = NULL) {
  it <- xml_add_child(parent, "ItemDef",
                      OID = oid, Name = name, DataType = type,
                      Length = as.character(len), SASFieldName = name)

  dd <- xml_add_child(it, "Description")
  xml_add_child(dd, "TranslatedText", label, "xml:lang" = "en")

  if (!is.null(codelist)) {
    xml_add_child(it, "CodeListRef", CodeListOID = codelist)
  }

  args <- list(it, "def:Origin", Type = origin_type)
  if (!is.null(origin_source)) args$Source <- origin_source
  do.call(xml_add_child, args)

  invisible(it)
}

add_item(mdv, "IT.STUDYID", "STUDYID", "text", 8,
         "Study Identifier", "Protocol", "Sponsor")

add_item(mdv, "IT.DM.USUBJID", "USUBJID", "text", 15,
         "Unique Subject Identifier", "Derived", "Sponsor")

sex_item <- add_item(mdv, "IT.DM.SEX", "SEX", "text", 1,
                     "Sex", "Collected", "Investigator", codelist = "CL.SEX")
```

Now the `ItemRef`s from the previous chunk have something to point at.
`IT.DM.SEX` also carries a `CodeListRef`, which is the third kind of
link in the document: variable to codelist.

### The codelist

```{r ex1-codelist}
cl <- xml_add_child(mdv, "CodeList", OID = "CL.SEX",
                    Name = "Sex", DataType = "text")

for (v in c("F", "M")) {
  ei <- xml_add_child(cl, "EnumeratedItem", CodedValue = v)
  xml_add_child(ei, "Alias", Context = "nci:ExtCodeID",
                Name = if (v == "F") "C16576" else "C20197")
}

xml_add_child(cl, "Alias", Context = "nci:ExtCodeID", Name = "C66731")
```

That `C66731` should look familiar. It is the NCI code for the CDISC
`SEX` codelist, the same one checked programmatically with
`sdtm.terminology` back in [SDTM Concepts](sdtm-concepts.qmd). The
`Alias` elements on each `EnumeratedItem` carry the codes for the
individual terms.

This is the answer to the third scenario at the top of the session. A
reviewer does not take the sponsor's word that `SEX` was coded to CDISC
terminology. They read `C66731` out of the define.xml and look it up.

### Run It Yourself

```{r ex1-write}
out <- file.path(tempdir(), "define_glpx.xml")
write_xml(odm, out)

schema <- read_xml(file.path(dx, "schema/cdisc-define-2.1/define2-1-0.xsd"))
result <- xml_validate(read_xml(out), schema)

as.logical(result)
```

### Reading the Output Line by Line

`TRUE`. The document you just assembled is a schema-valid Define-XML
v2.1 document.

That is worth a moment. `xml_validate()` checked your file against
`define2-1-0.xsd`: the same schema CDISC publishes and the same one a
sponsor's validation software loads. Nothing about this was simulated or
approximated. Roughly forty lines of R produced a file a validator
accepts.

Which raises the obvious question, and it is the one the rest of the
session is about: **if it validates, is it right?**

## Example 2: Origin Grows a Second Attribute

**Worked example, about 15 minutes.** Short, and it settles a question
left open back in Findings.

*Estimated time: ~15 minutes (worked example)*

[Findings](sdtm-findings.qmd) made a claim about GLPX-1's central lab
data that this session can now finish. It said that a central lab result
has `LBDRVFL` null because the value was not derived, and that the same
fact is recorded a second time in the Define-XML origin metadata, where
the origin is *Collected*.

That was right, and it was incomplete. In Define-XML v2.1, origin is not
one value:

> The def:Origin element includes the Type and Source attributes. The
> Type attribute indicates how the data originated. The Source attribute
> identifies the party responsible for the data's origin (Define-XML
> v2.1, §4.3.2)

And for SDTM, you do not get to skip one:

> For SDTM datasets, both Type and Source attributes are required,
> except in the case of Predecessor where the Source attribute is not
> used (Define-XML v2.1, §4.3.2.1)

So `Collected` is only half an answer. Collected *by whom*?

### The table that decides it

§4.3.2.1 gives a grid of Type against Source, with `X` marking
combinations that are not valid. Two rows matter for LB:

| Type | Subject | Investigator | Vendor | Sponsor |
|---|---|---|---|---|
| `Collected` | ePro | CRF | **Lab data, ECG** | X |
| `Derived` | X | X | Lab data, ECG | **SDTM** |

Read the `Collected` row. A central lab result is not collected by the
investigator. It is produced by the laboratory, which is a **Vendor**.
The spec's own example in that cell is literally "Lab data, ECG". And
`Collected` + `Sponsor` is marked `X`: invalid, because the sponsor
observes nothing.

Now read the `Derived` row for `LBDY`. §4.3.2.1's note says derivation
examples include "calculations performed during data collection (e.g.,
`--DY`)". `LBDY` was computed by the sponsor's SDTM programming, so it
is `Derived` + `Sponsor`.

```{r ex2-lb}
lb_origin <- tribble(
  ~variable,   ~type,        ~source,        ~why,
  "LBORRES",   "Collected",  "Investigator", "Result as reported on the CRF",
  "LBSTRESN",  "Collected",  "Vendor",       "Central lab result, standardised",
  "LBDY",      "Derived",    "Sponsor",      "Study day, computed in SDTM",
  "LBTESTCD",  "Assigned",   "Sponsor",      "Set to complete the SDTM structure",
  "DOMAIN",    "Assigned",   "Sponsor",      "Set to complete the SDTM structure"
)

knitr::kable(lb_origin, col.names = c("Variable", "Type", "Source", "Why"))
```

`LBTESTCD` and `DOMAIN` are `Assigned`, which the spec defines to
include "values set independently of any subject-related data value in
order to complete a dataset", and names `DOMAIN` and `--TESTCD` as its
examples (§4.3.2.1).

### Run It Yourself

Add the LB variable that Part 2 spent the most time on, and confirm the
`Collected`/`Vendor` pairing survives a schema check:

```{r ex2-run}
new_item <- xml_add_sibling(
  sex_item, "ItemDef",
  OID = "IT.LB.LBSTRESN", Name = "LBSTRESN",
  DataType = "float", Length = "8", SASFieldName = "LBSTRESN",
  .where = "after"
)

dd <- xml_add_child(new_item, "Description")
xml_add_child(dd, "TranslatedText",
              "Numeric Result/Finding in Standard Units", "xml:lang" = "en")

xml_add_child(new_item, "def:Origin", Type = "Collected", Source = "Vendor")

out2 <- file.path(tempdir(), "define_glpx2.xml")
write_xml(odm, out2)
as.logical(xml_validate(read_xml(out2), schema))
```

Note what that chunk does **not** do: call `add_item()`. The helper
appends its `ItemDef` to the end of `MetaDataVersion`, which by now
means *after* the `CodeList`, and the schema requires every `ItemDef`
to precede every `CodeList`. Appending returns `FALSE`, for exactly the
reason "What Can Go Wrong" is about to demonstrate deliberately.

So the new `ItemDef` is inserted as a **sibling immediately after an
existing one**, using the node `add_item()` returned in Example 1. This
is the practical face of §3.6's "the correct ordering of elements within
a document is an absolute requirement": in a hand-built define.xml,
*where you attach a node is part of whether the document is valid*, and
a convenience helper that always appends will eventually betray you.

The refinement to carry away: **`LBDRVFL` and `def:Origin` are not two
records of the same fact after all.** `LBDRVFL` says the value was not
derived. `def:Origin` says it was collected *and that a vendor collected
it*. The second is strictly more information, and it is information the
dataset has no column for.

## Example 3: Declaring the Standard, and a Decision Made Three Sessions Ago

**Worked example, about 15 minutes.** This one closes a loop opened in
Part 3. Read the version note carefully; it is the kind of detail that
costs a submission.

*Estimated time: ~15 minutes (worked example)*

[ADAE](adae.qmd) made a structural decision that mattered enough to
argue about at length: ADAE is built as **Class = OCCURRENCE DATA
STRUCTURE**, not as SubClass = ADVERSE EVENT. At the time, that decision
lived in the session's prose and in the way the variables were derived.

Define-XML is where it becomes a fact in a file.

```{r ex3-classes}
enum <- read_xml(file.path(dx, "schema/cdisc-define-2.1/define-enumerations.xsd"))

get_enum <- function(doc, name) {
  node <- xml_find_first(doc, sprintf("//xs:simpleType[@name='%s']", name),
                         ns = c(xs = "http://www.w3.org/2001/XMLSchema"))
  xml_attr(xml_find_all(node, ".//xs:enumeration",
                        ns = c(xs = "http://www.w3.org/2001/XMLSchema")), "value")
}

get_enum(enum, "ItemGroupClass")
```

`OCCURRENCE DATA STRUCTURE` is in that list, alongside
`BASIC DATA STRUCTURE` for ADLB and `SUBJECT LEVEL ANALYSIS DATASET` for
ADSL. And the SubClass:

```{r ex3-subclass}
get_enum(enum, "ItemGroupSubClass")
```

`ADVERSE EVENT` is a **SubClass**, in a different list. So the choice
ADAE made is not a matter of emphasis or documentation style. It is a
choice between two attributes on two different enumerated lists, and
define.xml forces you to declare which one you meant.

### The version trap

Here is where the frozen-PDF-versus-live-schema gap becomes concrete.

```{r ex3-standardname}
get_enum(enum, "StandardName")
```

The course builds ADAE to **OCCDS v1.1**. There is no `ADaM-OCCDSIG` in
that list. You cannot declare OCCDS as a standard in a Define-XML
v2.1.11 document; an OCCDS dataset is declared under `ADaMIG`.

It used to be there. The package's own `README.md` records the removal:

> Enumeration for ADaM-OCCDSIG (C214535) was deleted
> (Define-XML v2.1.11 release notes, changes in v2.1.11)

Added at v2.1.9, deleted at v2.1.11. Two point releases.

Now check the specification PDF against the same schema for SDTM
versions. §4.1.1 lists SDTMIG allowable versions as 3.1.2, 3.2 and 3.3,
and this course uses v3.4, which Example 1 declared:

```{r ex3-sdtmig}
get_enum(enum, "SdtmigVersionResponse")
```

`3.4` is there. The 2019 PDF does not mention it; the schema does.

**This is not the specification being out of date and the schema
contradicting it.** Read what §4.3.2 actually says about its own tables:

> In advance of CDISC publication of controlled terminology, the terms
> in the following tables can be used... Users are encouraged to refer
> to the most recent version of CDISC Controlled Terminology
> (Define-XML v2.1, §4.3.2)

The specification designated its own lists as provisional and pointed at
Controlled Terminology as the authority. The schema *is* that authority,
regenerated for each CT package. So the rule to carry away:

> **For any enumerated value in Define-XML, the schema is current and
> the PDF is a snapshot. Check `define-enumerations.xsd`, not §4.1.1.**

The same discipline the course applied to SDTM terminology with
`sdtm.terminology` rather than reading the CT PDF, applied one level up.

## What Can Go Wrong

**Reading task, about 10 minutes.** Includes a document that validates
perfectly and is still broken, and one of our own files failing a real
conformance rule.

*Estimated time: ~10 minutes (reading)*

Example 1 ended on a question: if it validates, is it right? The
specification answers it directly, and the answer is no.

> Once a Define-XML document is valid according to the schema,
> validation software should consider all other Define-XML requirements
> in the specification. These include rules about conditionally required
> components or other business rules in this document. **The Define-XML
> schema can only enforce some of the standard** (Define-XML v2.1, §3.6)

CDISC publishes the rest as a separate deliverable: **Conformance Rules
for Define-XML v2.1**, a spreadsheet of 225 rules. Each rule records
where it comes from, and the split is the whole story:

| Source | Rules | Who catches it |
|---|---|---|
| Schema | 135 | `xml_validate()` |
| Specification | 90 | Nothing, unless you check |

**Ninety rules that no schema validation will ever catch.** A document
can validate `TRUE` and violate any of them.

### Three failures the schema does catch

**Case sensitivity.** The spec warns about it explicitly: "XML is
case-sensitive... `<def:Class Name="findings" />` is not valid, but
`<def:Class Name="FINDINGS" />` is valid" (§3.6). Lowercasing our DM
class produces:

```
Element '{http://www.cdisc.org/ns/def/v2.1}Class', attribute 'Name':
[facet 'enumeration'] The value 'special purpose' is not an element of
the set {'ADAM OTHER', 'BASIC DATA STRUCTURE', ..., 'SPECIAL PURPOSE',
'STUDY REFERENCE', 'SUBJECT LEVEL ANALYSIS DATASET', 'TRIAL DESIGN'}.
```

Note that the error prints the entire permitted set. Schema errors in
Define-XML are unusually informative, worth reading rather than
skimming.

**Element ordering.** Moving `def:Class` above the `Description` inside
`ItemGroupDef` gives:

```
Element '{http://www.cdisc.org/ns/odm/v1.3}Description': This element is
not expected. Expected is one of ( {http://www.cdisc.org/ns/def/v2.1}Class,
{http://www.cdisc.org/ns/def/v2.1}leaf ).
```

Order is not cosmetic here: "The correct ordering of elements within a
document is an absolute requirement for the document to be valid with
respect to the Define-XML schema" (§3.6).

**A misspelled enumerated value**, as in the version trap above.

### The failure the schema does not catch

Take the document from Example 1 and break the link, point an `ItemRef`
at an OID that no `ItemDef` declares:

```{r wcg-dangling}
broken <- read_xml(out)
node <- xml_find_first(broken, "//d1:ItemRef[@ItemOID='IT.DM.SEX']")
xml_set_attr(node, "ItemOID", "IT.DM.DOES_NOT_EXIST")

xml_validate(broken, schema)
```

`TRUE`. The document is schema-valid and structurally broken: a dataset
claims a variable that is defined nowhere. This is exactly the
consequence of the central intuition: the schema knows `ItemOID` must
be a string, because that is all a schema can know. It cannot know
whether the string points at anything, because OIDs have no meaning to
check against.

Conformance rule 65 is the one that catches it, and its Source Type is
`Specification`:

> The ItemOID attribute on the ItemRef element must reference an
> existing ItemDef element (Conformance Rules for Define-XML v2.1, rule
> 65)

Which you can check yourself in three lines:

```{r wcg-rule65}
refs <- xml_attr(xml_find_all(broken, "//d1:ItemGroupDef/d1:ItemRef"), "ItemOID")
defs <- xml_attr(xml_find_all(broken, "//d1:ItemDef"), "OID")
setdiff(refs, defs)
```

### The one your own file already fails

Rule 65 needed a deliberately broken document. This one does not.

> The MethodOID attribute must be provided on the ItemRef element when
> the Origin Type attribute on the corresponding ItemDef element is
> "Derived" (Conformance Rules for Define-XML v2.1, rule 73)

In Example 1, `USUBJID` was given `Origin Type = "Derived"`. Its
`ItemRef` has no `MethodOID`. Nothing declares *how* it was derived:

```{r wcg-rule73}
doc <- read_xml(out)

derived <- xml_attr(
  xml_find_all(doc, "//d1:ItemDef[def:Origin/@Type='Derived']"), "OID")

missing_method <- Filter(function(oid) {
  ref <- xml_find_first(
    doc, sprintf("//d1:ItemGroupDef/d1:ItemRef[@ItemOID='%s']", oid))
  !is.na(ref) && is.na(xml_attr(ref, "MethodOID"))
}, derived)

missing_method
```

The file you built and validated in Example 1 is non-conformant, and
`xml_validate()` returned `TRUE` on it. That is not a mistake in the
example. It is the most useful thing this session can show you.
Fixing it means adding a `MethodDef` that states the derivation and a
`MethodOID` on the `ItemRef` that points at it, which is what
[Exercise 8](../exercises/define-xml.qmd) asks you to do.

### Common misinterpretations

- **"The define.xml is documentation."** It is a submission deliverable
  with its own conformance rules, not a README. A submission with
  conformant datasets and a non-conformant define.xml is a
  non-conformant submission.
- **"Validation means conformance."** 135 of 225 rules, at best. Say
  "schema-valid" when you mean schema-valid.
- **"OID naming conventions are part of the standard."** They are not
  (§3.5.1). Use them (they make files readable) but never rely on an
  OID's *shape* to infer what it points at. Follow the link.
- **"Origin is one value."** For SDTM and ADaM, Type *and* Source are
  both required (§4.3.2.1, §4.3.2.2).
- **"The spec PDF is the authority on permitted values."** For
  enumerated values it is a 2019 snapshot; the schema is regenerated
  with each CT package.

## Exercises

**Exercise, about 15 minutes.** Do these before the next session. You
will need the CDISC Define-XML package downloaded first.

[Exercise 8: Define-XML](../exercises/define-xml.qmd) extends this
document: a `MethodDef` to resolve rule 73, an `ItemGroupDef` for LB
with value-level metadata, and the variable this course cannot populate
at all.

## Comprehension Check

**Self-test, about 10 minutes.** Answer before expanding.

::: {.callout-note collapse="true"}
### Questions

1. A colleague proposes renaming every OID in your define.xml to a UUID,
   arguing it is cleaner. Is the resulting document still conformant?
   What is actually lost?
2. GLPX-1's HbA1c results come from a central laboratory. What
   `Type` and `Source` belong on `LBSTRESN`, and why is
   `Collected`/`Sponsor` invalid?
3. Your define.xml returns `TRUE` from `xml_validate()`. What proportion
   of the published conformance rules have you demonstrated compliance
   with?
4. Why can an OCCDS-structured ADAE not be declared with
   `Name="ADaM-OCCDSIG"` in a v2.1.11 document, and what do you declare
   instead?
5. `ItemDef` elements sit beside `ItemGroupDef`, not inside it. What
   does that arrangement make possible?
:::

::: {.callout-note collapse="true"}
### Answers

1. **Yes, still conformant.** The specification does not mandate an OID
   format and states that random identifiers are equally valid (§3.5.1).
   What is lost is human readability: nothing machine-checkable. This
   is the central intuition: the structure lives in which OID points at
   which, not in what the OIDs say.

2. **`Type="Collected"`, `Source="Vendor"`.** The value is observed and
   recorded, so the Type is `Collected`; a central laboratory is a
   vendor, and §4.3.2.1's example in that cell is "Lab data, ECG".
   `Collected`/`Sponsor` is marked `X` in the table, invalid, because
   the sponsor observes nothing. Both attributes are required for SDTM.

3. **At most 135 of 225: the Schema-sourced rules.** The remaining 90
   come from the specification and no schema validation touches them
   (§3.6). Rule 73 is the example: the Example 1 document validates and
   still fails it.

4. **`ADaM-OCCDSIG` is not in the `StandardName` enumeration in
   v2.1.11**. It was added at v2.1.9 and deleted at v2.1.11 (package
   release notes). Declare `Name="ADaMIG"` instead, and record the
   occurrence structure through `def:Class="OCCURRENCE DATA STRUCTURE"`
   on the `ItemGroupDef`, which is where that decision belongs anyway.

5. **Definition once, reference many times.** `STUDYID` appears in
   almost every domain but is declared in a single `ItemDef` that every
   `ItemGroupDef` references by OID (§3.5.1). Nesting definitions inside
   datasets would force a copy per dataset, and copies drift.
:::
