---
title: "SDTM: Findings (Labs & Vital Signs)"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(sdtm.terminology)
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 100 minutes.**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background: what a Findings row is | 20 min | Reading |
| Example 1: Building LB — results, units, and a duplicate | 25 min | Worked example |
| Example 2: Reference ranges, a real outlier, and baseline | 25 min | Worked example |
| The finished SDTM, saved | 10 min | Reading |
| What Can Go Wrong | 10 min | Reading |
| Exercises + Comprehension Check | 15 min | Practice / Self-test |

The sibling of [Events](sdtm-events.qmd) and the last domain-building
session in Part 2. It resolves three deliberate defects (a duplicate
record, a missing unit, a real out-of-range value) and completes the
persisted SDTM the ADaM sessions will read. Split after Example 1 for
self-paced study.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown.
Where you see a "Run It Yourself" box, stop and run the code above it
before reading the explanation that follows. Run R from the project
root, as in earlier sessions.

Controlled-terminology checks use the **`sdtm.terminology`** package,
as in [SDTM Concepts](sdtm-concepts.qmd).
:::

## When Do You Use This? {#sec-sdtm-findings}

::: {.callout-tip}
A lab transfer lands and you have to make it submission-ready: it has a
record the vendor sent twice, a result whose unit field is blank, and a
value so high it looks like a typo — except it isn't. Then the analysis
team asks for a finished SDTM they can build ADaM from.

All of it is the same task: representing *measurements*. This session
asks: **once you know a dataset is about results rather than
occurrences, how do you build it — and how does the standard help you
tell a data error from a real extreme?**

Builds on [SDTM Concepts](sdtm-concepts.qmd) (the result quartet;
controlled terminology), [The DM Domain](sdtm-dm.qmd) (RFSTDTC and the
study-day rule), and [Events](sdtm-events.qmd) (its sibling). After
this session the SDTM is complete.
:::

## Learning Objectives

After completing this session you will be able to:

- Recognise a Findings-class domain from what one row represents, and
  predict its result variables
- Build an LB dataset with original and standardised results and units
- Use a domain's natural key to detect and remove a duplicate record
- Recover a missing unit from documented sources rather than guessing
- Derive a reference-range indicator, and distinguish a mechanical
  range flag from the judgement of whether a value is real
- Derive an operational baseline flag, and say why the authoritative
  baseline lives in ADaM
- Explain how the raw data becomes the committed SDTM through a
  reproducible build

## Background: What a Findings Row Is

*Estimated time: ~20 minutes (reading)*

::: {.callout-important}
### The one thing to remember

**A Findings row records what a measurement was — its test and its
result, kept twice as collected and standardised — and the domain's
natural key is what makes a duplicate detectable and a real outlier
keepable.**
:::

Laboratory Test Results (LB) is a Findings-class domain, and its
structure is **one record per lab test per time point per visit per
subject** (SDTMIG v3.4, §6.3.5.6). That sentence is also the domain's
**natural key**: the combination of subject, visit, and test that
should identify a row uniquely. Hold onto it — it is what turns "the
vendor sent a duplicate" from a vague worry into a checkable fact.

The **topic** is `LBTESTCD` (the test), and the result is not one
column but a small family, all Expected (SDTMIG v3.4, §6.3.5.6;
§4.5.1.1): `LBORRES` (the result as collected) with `LBORRESU` (its
unit), and the standardised pair `LBSTRESC` (character) / `LBSTRESN`
(numeric) with `LBSTRESU`. You met this in [SDTM
Concepts](sdtm-concepts.qmd): the original is always kept beside the
standardised version. Units are controlled terminology — `LBORRESU`
and `LBSTRESU` draw on the UNIT codelist, and a unit the vendor uses
that is not a submission value must be mapped to an equivalent or added
in the Define-XML (SDTMIG v3.4, §6.3.5.6, assumption 7).

Findings also carry a **reference range** and an indicator of where the
result falls in it (SDTMIG v3.4, §6.3.5.6): `LBORNRLO`/`LBORNRHI` (the
range in original units), `LBSTNRLO`/`LBSTNRHI` (in standard units),
and `LBNRIND` — the "Reference Range Indicator", which "indicates where
the value falls with respect to reference range defined by LBORNRLO and
LBORNRHI" (SDTMIG v3.4, §6.3.5.6), taking values from the NRIND codelist
(`NORMAL`, `LOW`, `HIGH`, `ABNORMAL`).

::: {.callout-important}
### LBNRIND is a comparison, not a verdict

`LBNRIND` is a **mechanical** comparison of a number against a range. It
is *not* a clinical-significance flag and *not* a judgement about
whether the value is a data error (SDTMIG v3.4, §6.3.5.6, assumption 3;
§4.5.5). Whether a `HIGH` value is a genuine extreme or a transcription
mistake is a separate question, answered by data review, not by the
flag. Example 2 turns on exactly this distinction.
:::

One more, because it recurs: a value produced by a central lab
following its own procedures is considered **collected, not derived**,
so its `LBDRVFL` (derived flag) is null (SDTMIG v3.4, §6.3.5.6,
assumption 6; §4.1.8.1). That is why a byte-perfect duplicate from the
vendor is a *transmission artefact* to remove — not a real second
measurement to keep.

## Example 1: Building LB — Results, Units, and a Duplicate

*Estimated time: ~25 minutes (worked example)*

### The natural key, and the duplicate that breaks it

Read the raw lab file and test the natural key directly — is any
(subject, visit, test) combination present more than once?

```{r read-key}
#| eval: true

lb_raw <- read_csv("data/raw/lb_raw.csv", show_col_types = FALSE,
                   col_types = cols(.default = col_character()))

nrow(lb_raw)

lb_raw |> count(subjid, visit, test) |> filter(n > 1)
```

One key value occurs twice: subject `101-010`'s screening HbA1c. Look
at the two rows:

```{r show-dup}
#| eval: true

lb_raw |>
  filter(subjid == "101-010", visit == "SCREENING", test == "HBA1C")
```

::: {.callout-note}
### Reading the Output Line by Line

The two rows are **byte-for-byte identical** — same result, same date,
same everything. Because a central-lab value is collected once (SDTMIG
v3.4, §6.3.5.6, assumption 6), two identical rows for one key cannot
both be real measurements; the second is a transmission artefact. This
is the one case where removing a row is correct — and the natural key
is what licenses it. Contrast this with the out-of-range value in
Example 2, which is also unusual but must be *kept*: the difference is
that the duplicate violates the key and the outlier does not.
:::

Remove it with `distinct()`, which drops only fully identical rows:

```{r dedup}
#| eval: true

lb <- distinct(lb_raw)

c(before = nrow(lb_raw), after = nrow(lb), removed = nrow(lb_raw) - nrow(lb))
```

Exactly one row goes. Had the two rows differed in any field — a
different result, a different date — `distinct()` would have kept both,
and you would have a genuine discrepancy to investigate rather than a
duplicate to drop. De-duplication is safe *only* because the rows are
identical.

### The result quartet, and a missing unit

Now map the results. The collected value and unit become
`LBORRES`/`LBORRESU`; the standardised versions become
`LBSTRESC`/`LBSTRESN`/`LBSTRESU`. These analytes need no unit
conversion, so standardising is a copy — but first, defect D7: one
glucose result arrived with a blank unit.

```{r missing-unit}
#| eval: true

lb |>
  filter(subjid == "101-037", visit == "WEEK 8", test == "GLUC") |>
  select(subjid, test, result, unit)

# a blank is not a valid unit value
is_term("", clst_code = "C71620")
```

The unit is missing, and a blank is not a legal UNIT value. You met
this row in [SDTM Concepts](sdtm-concepts.qmd); here you resolve it. The
resolution is not a guess: the vendor's documented panel measures
glucose in `mmol/L`, and every other glucose result in the file confirms
it, so the unit is **recovered from the documented source** and the gap
is logged as a query to the vendor — the same discipline as the sex
conflict in [The DM Domain](sdtm-dm.qmd):

```{r resolve-unit}
#| eval: true

panel_unit <- c(HBA1C = "%", GLUC = "mmol/L", ALT = "U/L", CREAT = "umol/L")

lb <- lb |>
  mutate(unit = if_else(is.na(unit) | unit == "", unname(panel_unit[test]), unit))

# the recovered unit is valid CT
is_term("mmol/L", clst_code = "C71620")
```

::: {.callout-tip}
### Run It Yourself

Run the chunks. The blank returns `FALSE` (not a unit), and `mmol/L`
returns `TRUE`. Notice the shape of the fix: the value was **recovered
from a documented source and logged**, not invented to fill the column.
If the vendor's panel had *not* pinned the unit, the honest step would
be to leave it null and query — never to assume.
:::

## Example 2: Reference Ranges, a Real Outlier, and Baseline

*Estimated time: ~25 minutes (worked example)*

### Deriving the range indicator

The raw file carries the vendor's reference range per result
(`ref_lo`/`ref_hi`). Map them to `LBORNRLO`/`LBORNRHI`, and **derive**
`LBNRIND` by comparing the result to the range — never hard-code it:

```{r nrind}
#| eval: true

lb <- lb |>
  mutate(
    LBORRES  = result,
    LBSTRESN = as.numeric(result),
    LBORNRLO = ref_lo,
    LBORNRHI = ref_hi,
    LBNRIND  = case_when(
      as.numeric(result) > as.numeric(ref_hi) ~ "HIGH",
      as.numeric(result) < as.numeric(ref_lo) ~ "LOW",
      TRUE                                    ~ "NORMAL"
    )
  )

lb |> filter(test == "HBA1C") |> count(LBNRIND)
```

::: {.callout-note}
### Reading the Output Line by Line

Almost every HbA1c reads `HIGH`. That is correct and expected: the
reference range (4.0–5.6%) is the *non-diabetic* normal, and this is a
trial of people with type 2 diabetes, so nearly all of them sit above
it — exactly as they would on a real lab report. `LBNRIND` is doing
precisely its job: a mechanical comparison against the range. It is not
telling you anything is wrong. Whether HbA1c should have flagged against
a diabetic-appropriate range is an *analysis* decision (ADaM); SDTM
records the range the lab supplied.
:::

### The value that looks like an error but isn't

Now defect D5. One screening HbA1c is `13.9%`. Against the range it is
`HIGH` — like every other HbA1c, so the flag alone does not single it
out. What singles it out is its *magnitude*:

```{r outlier}
#| eval: true

screening <- lb |>
  filter(test == "HBA1C", visit == "SCREENING") |>
  mutate(result = as.numeric(result))

c(median = median(screening$result),
  max    = max(screening$result),
  n_at_or_above_13.9 = sum(screening$result >= 13.9))
```

::: {.callout-note}
### Reading the Output Line by Line

`13.9%` is the single highest screening HbA1c in the trial — one value
of 400, against a median of about 8.4. That is the kind of extreme an
**edit check** flags automatically: it looks like it could be a
transcription error (a slipped digit, an 8.9 mistyped). The flag routes
the value into **data review** — a process, not a hunch — which checks
it against the source record. Here the source confirms it: a genuinely,
severely uncontrolled patient. So the value is **kept**.

This is the whole lesson of D5, and it is a *process*, not a person's
opinion: an automated check flags an implausible-looking value → data
review investigates it against source → the value is confirmed real →
it is retained, unchanged. A range flag (`LBNRIND = HIGH`) and a
plausibility flag are different signals; neither is permission to delete
or "correct" a value. Silently changing `13.9` to something tidier would
be fabricating data — the opposite of what the review exists to prevent.
:::

### The operational baseline flag

Finally, `LBLOBXFL` — the flag SDTM uses for a consistent pre-treatment
reference value, defined as the **last non-missing value prior to
RFXSTDTC** (SDTMIG v3.4, §4.5.9). It needs each subject's first-exposure
date, which lives in the DM you built last session:

```{r baseline}
#| eval: true

dm <- read_csv("data/sdtm/dm.csv", show_col_types = FALSE,
               col_types = cols(.default = col_character()))

lb <- lb |>
  left_join(dm |> select(subjid = SUBJID, RFXSTDTC), by = "subjid") |>
  group_by(subjid, test) |>
  mutate(
    is_pre   = as.Date(colldt) < as.Date(RFXSTDTC),
    last_pre = suppressWarnings(max(as.Date(colldt)[is_pre])),
    LBLOBXFL = if_else(is_pre & is.finite(last_pre) & as.Date(colldt) == last_pre,
                       "Y", NA_character_)
  ) |>
  ungroup()

lb |>
  filter(subjid == "101-010", test == "HBA1C") |>
  select(visit, colldt, LBLOBXFL)
```

::: {.callout-note}
### Reading the Output Line by Line

Exactly one row per subject per test is flagged `Y` — here the
screening draw, the last HbA1c taken before this subject's first dose.
Everything after treatment starts is null. `LBLOBXFL` gives every study
a *consistent* pre-treatment reference, computed the same way
everywhere. But note the standard's own caution (SDTMIG v3.4, §4.5.9):
this is an operational flag, and the **authoritative** baseline for
analysis — which visit counts as baseline, how it is chosen — is
defined later, in ADaM. You will set it in [ADSL](adsl.qmd). SDTM marks
a consistent reference; ADaM decides what baseline *means* for the
analysis.
:::

## The Finished SDTM, Saved

*Estimated time: ~10 minutes (reading)*

You have now built all four of GLPX-001's domains across three sessions
— DM, AE, and here LB and VS. Rebuilding them by hand every time a later
session needs them would invite drift: the DM logic in one session
subtly disagreeing with another. So the course reifies exactly these
mappings in one script, [`R/build_sdtm.R`](https://github.com/sufyansuleman/cdisc-with-r/blob/main/R/build_sdtm.R),
which reads `data/raw/` and writes the finished domains to `data/sdtm/`:

```
data/raw/*.csv   --  simulate_trial.R  (the EDC + vendor extracts)
      |
      |  R/build_sdtm.R   (the mappings these sessions taught)
      v
data/sdtm/{dm,ae,lb,vs}.csv   --  conformant, committed, reproducible
```

Two properties make this more than a convenience, and both are the
subject matter of the course, not incidental to it:

- **Reproducibility.** `simulate_trial.R` is fixed-seed and
  `build_sdtm.R` has no randomness, so `data/sdtm/` is fully determined
  by those two scripts. Anyone who clones the repository regenerates
  byte-identical datasets. Reproducibility is not a nicety in regulated
  work; it is a requirement, and here the pipeline demonstrates it.
- **Traceability.** Every value in `data/sdtm/lb.csv` traces back
  through `build_sdtm.R` to a row in `data/raw/lb_raw.csv` — the
  provenance a reviewer needs to follow a result from a table back to
  its source. You will meet traceability again, formally, in
  [define.xml](define-xml.qmd) and [ADRG and
  Submission](adrg-and-submission.qmd).

From here on, the ADaM sessions read `data/sdtm/` rather than rebuilding
from raw. The committed CSVs are the hand-off point between "what
happened in the trial" (SDTM) and "what the trial shows" (ADaM).

```{r read-built}
#| eval: true

built <- read_csv("data/sdtm/lb.csv", show_col_types = FALSE,
                  col_types = cols(.default = col_character()))

c(rows = nrow(built), subjects = n_distinct(built$USUBJID))
```

## What Can Go Wrong

*Estimated time: ~10 minutes (reading)*

**Deleting the out-of-range value as an error**

The 13.9% HbA1c is the most tempting mistake in the session. It looks
wrong; it is real. A range flag and a plausibility flag both send it to
review, and review — checking source — keeps it. Deleting or "correcting"
a value because it is extreme is fabricating data.

**De-duplicating rows that are not duplicates**

`distinct()` is safe only for byte-perfect duplicates. Two rows for one
key that differ in *any* field are a discrepancy to investigate, not a
duplicate to drop — collapsing them silently discards a real, conflicting
measurement.

**Guessing a missing unit**

The blank glucose unit was recovered from the vendor's documented panel
and logged. Typing `mmol/L` because it "must be" that is a different
act — an assumption dressed as data. Recover from a source, or leave
null and query.

**Reading LBNRIND as a quality or significance flag**

`LBNRIND = HIGH` on nearly every HbA1c does not mean the data is bad or
the finding clinically significant. It is a mechanical comparison
against the supplied range (SDTMIG v3.4, §6.3.5.6, assumption 3).
Clinical significance and analysis-appropriate ranges are separate,
later, and elsewhere.

**Treating LBLOBXFL as the analysis baseline**

`LBLOBXFL` is an operational SDTM flag; the baseline that drives
change-from-baseline analysis is defined in ADaM (SDTMIG v3.4, §4.5.9).
Using the SDTM flag as if it were the analysis baseline skips the
decision ADaM exists to record.

::: {.callout-warning}
### Common misinterpretations

**"An out-of-range value is a data error"**
`LBNRIND` marks where a value sits relative to a range; it says nothing
about whether the value is correct. The 13.9% is out of range *and*
real. Range position and data quality are different questions.

**"Every HIGH flag needs action"**
In a diabetes trial almost every HbA1c is HIGH against the non-diabetic
range. A flag that fires for nearly everyone is describing the
population, not signalling individual problems.

**"The vendor file is clean because it is machine-generated"**
This file had a duplicate transmission and a blank unit. Consistent
formatting is not the same as correct data — the same lesson as [Why
Standards](why-standards.qmd).

**"data/sdtm/ is just cached output I can edit"**
It is generated. Editing a CSV by hand breaks the traceability the whole
pipeline exists to provide; the way to change it is to change
`build_sdtm.R` and regenerate.
:::

## Exercises

The Findings exercise builds the **VS** domain — the structural twin of
LB, with one deliberate difference to catch: [Exercise — SDTM: Findings
(Labs & Vital Signs)](../exercises/sdtm-findings.qmd). Solutions are in
the paid tier.

## Comprehension Check

*Estimated time: ~10 minutes*

1. What is the natural key of the LB domain, and how did it let you
   remove exactly one row safely?
2. A glucose result arrived with a blank unit. What did you put in
   `LBORRESU`, and what makes that a resolution rather than a guess?
3. Nearly every HbA1c in the trial reads `LBNRIND = HIGH`. Is that a
   data problem? What is `LBNRIND` actually telling you?
4. The 13.9% HbA1c is flagged, reviewed, and kept. Describe the process
   that led to "kept", and say why deleting it would be wrong.
5. `LBLOBXFL` marks one pre-treatment value per subject per test. Why is
   it not the baseline used for the analysis?

::: {.callout-note collapse="true"}
### Answers

1. The natural key is subject + visit + test — "one record per lab test
   per time point per visit per subject" (SDTMIG v3.4, §6.3.5.6). One
   key value (subject `101-010`, screening, HbA1c) appeared twice, and
   the two rows were byte-for-byte identical, so one was a transmission
   artefact of a collected-once value (assumption 6) and `distinct()`
   removed exactly it. Removal was safe *only* because the rows were
   identical.

2. `mmol/L` — recovered from the vendor's documented panel, which
   measures glucose in `mmol/L`, as every other glucose result confirms;
   the gap was logged as a query. It is a resolution rather than a guess
   because the value came from a documented source, not from assuming
   what the unit "should" be. Had no source pinned it, the honest step
   is null plus a query.

3. Not a data problem. `LBNRIND` is a mechanical comparison of the result
   against the supplied reference range (SDTMIG v3.4, §6.3.5.6,
   assumption 3). The range is the non-diabetic normal, so a diabetic
   cohort reads `HIGH` almost everywhere — the flag is describing the
   population relative to that range, not signalling errors or clinical
   significance.

4. An automated edit check flagged 13.9% as an implausible extreme (the
   highest of 400 screening values, against a median near 8.4); that
   routed it to data review; review checked it against the source record
   and confirmed a genuinely uncontrolled patient; so it was retained
   unchanged. Deleting or "correcting" it would fabricate data — the
   value is real, and its being extreme is not evidence that it is wrong.

5. `LBLOBXFL` is an *operational* flag — the last non-missing value
   before first exposure, computed identically across studies (SDTMIG
   v3.4, §4.5.9). The *authoritative* baseline for analysis (which visit
   counts, how it is chosen, how missing baselines are handled) is a
   decision recorded in ADaM, not SDTM. You set it in [ADSL](adsl.qmd).
:::
