---
title: "SDTM Concepts"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(sdtm.terminology)
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 90 minutes.**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background: the model | 25 min | Reading |
| Example 1: Anatomy of a domain | 20 min | Worked example |
| Example 2: Original result vs standardised result | 20 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Exercises + Comprehension Check | 10 min | Practice / Self-test |

This is the first session of Part 2 and the first that is dense with
standard-specific detail. For self-paced study, take the Background in
one sitting and the two worked examples in another. It underpins every
domain you build afterwards.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown.
Where you see a "Run It Yourself" box, stop and run the code above it
before reading the explanation that follows. Run R from the project
root, as in earlier sessions.

This session also uses the **`sdtm.terminology`** package to check
controlled-terminology values programmatically, rather than trusting
anyone's memory of what a codelist contains. If `renv::restore()` has
not already provided it, install it once with
`install.packages("sdtm.terminology")`.
:::

## When Do You Use This? {#sec-sdtm-concepts}

::: {.callout-tip}
You have the five raw GLPX-001 files and you have been told to "map
them to SDTM." Before you rename a single column you hit the questions
this session answers: how many datasets should there be, and which raw
file becomes which? What is a single row supposed to represent in each?
Which columns are mandatory, which are optional, and who decides? And
when the lab sends a result in one form but the standard expects
another, where does each version live?

Every one of those is a **modelling question, not a coding question.**
This session is the vocabulary and the mental model you need before
[The DM Domain](sdtm-dm.qmd), [Events](sdtm-events.qmd) and
[Findings](sdtm-findings.qmd), where you build real domains from
the raw data.

It builds on [Why Standards](why-standards.qmd) (why a shared structure
exists) and [The Trial: GLPX-001](the-trial-glpx001.qmd) (the raw data
you will now reshape).
:::

## Learning Objectives

After completing this session you will be able to:

- Explain what SDTM standardises and how the implementation guide
  relates to the general model
- Classify a dataset into its observation class from what one row
  represents
- Identify the role a variable plays in a record — who it identifies,
  what it is about, when it happened, and how it qualifies the result
- Read a variable's core designation and say what obligation it imposes
- Distinguish a result as originally collected from its standardised
  form, and say why both are kept
- Check a value against CDISC Controlled Terminology programmatically
  with `sdtm.terminology`, instead of recalling codelist contents

## Background: The Model Behind the Datasets

*Estimated time: ~25 minutes (reading)*

::: {.callout-important}
### The one thing to remember

**SDTM doesn't tell you new facts about the trial — it puts the facts
you already have into fixed boxes, so that what one row means, what
each column's job is, and what may go in it are the same in every
study.**
:::

### The model and the guide

Two documents govern this work, and they are not the same thing. The
**SDTM** (Study Data Tabulation Model) is the general framework: it
defines the kinds of boxes that exist. The **SDTMIG** (the
implementation guide for human clinical trials) is what turns that
framework into concrete datasets with named variables you can actually
build. The guide is *based on* the model (SDTMIG v3.4, §2.1); this
course uses **SDTMIG v3.4**, which is meant to be used together with
**SDTM v2.0** (SDTMIG v3.4, §1.1).

Which version is not a matter of taste. The applicable SDTMIG version
for a study is fixed by the study's start date against the FDA Data
Standards Catalog [@fdadsc11]. Reading that catalogue for clinical
study datasets:

- **SDTMIG v3.2** — FDA support **ended** 2023-12-13.
- **SDTMIG v3.3** — supported since 2020-07-07; **required** for
  studies starting on or after 2022-03-15.
- **SDTMIG v3.4** — supported since 2023-12-13; **required** for
  studies starting on or after 2025-03-15; pairs with SDTM v2.0.

GLPX-001 starts in 2024, so its version is settled by that start date
falling in the v3.4 support window — not by preference. This is why
"SDTM is required" is too loose to be useful: what is required is a
*specific implementation-guide version over a specific window of study
start dates*, and the catalogue's "date support begins / date support
ends / date requirement begins" columns are the whole point. (The
study's start date itself travels in the submission in a small dataset
called `ts.xpt`; you will meet it in [ADRG and
Submission](adrg-and-submission.qmd).)

### The three general observation classes

Here is the first big idea. Most of what a trial collects about its
subjects fits into one of **three general observation classes**, and
which one a dataset belongs to is decided by what one row *is* (SDTMIG
v3.4, §2.3):

| Class | One row captures… | GLPX example |
|---|---|---|
| **Interventions** | something done to or taken by the subject | study drug exposure (EX) |
| **Events** | something that happened to the subject | an adverse event (AE) |
| **Findings** | the result of an observation or measurement | a lab test (LB), a vital sign (VS) |

The distinction is not cosmetic — it determines which variables a
dataset is even allowed to have. A findings dataset is built around a
*result*; an events dataset is built around a *thing that occurred*;
an interventions dataset around a *treatment*. Most trial data, being
measurements and responses, lands in Findings (SDTMIG v3.4, §2.3).

Not everything fits the three classes. SDTM also defines
**special-purpose** domains, trial-design datasets, relationship
datasets, and study-reference datasets (SDTMIG v3.4, §2.4).
**Demographics (DM)** is the one that matters most here: it is a
special-purpose domain — one record per subject (SDTMIG v3.4, §5.2) —
not a member of any general observation class.

### Domains, and the two-character code

A **domain** is a collection of logically related observations with a
common topic, and each domain is one dataset (SDTMIG v3.4, §2.2). Every
domain has a unique **two-character code** — and that code does four
jobs at once: it is the dataset name, the value of the `DOMAIN`
variable on every row, the prefix on most of the domain's variable
names, and the domain's reference in relationship datasets (SDTMIG
v3.4, §2.2).

That third job is the one to internalise. The implementation guide
writes variable names generically with a `--` placeholder — `--TESTCD`,
`--ORRES` — and you replace the `--` with the domain code to get the
real name: `LBTESTCD` in LB, `VSTESTCD` in VS (SDTMIG v3.4, §2.2,
§4.2.2). So `--ORRES` is not itself a variable; it is a *pattern*. When
this course writes `--STRESC`, read it as "the STRESC variable of
whichever domain we're in."

The two-character codes are themselves controlled terminology — they
come from a published codelist, not from your imagination. Example 1
checks GLPX's codes against it.

### Identifiers: the spine of every record

Four **identifier** variables are required in every domain built on a
general observation class: `STUDYID`, `DOMAIN`, `USUBJID`, and `--SEQ`
(SDTMIG v3.4, §2.5). Two of them carry rules worth stating precisely,
because they are exactly the rules that the raw-data problems from
earlier sessions violate:

- **`USUBJID`** — the unique subject identifier — must be unique for
  each subject across *all trials in the submission*, not merely within
  one study (SDTMIG v3.4, §4.2.3). This is the rule that the
  duplicate-enrolment subject from [The DM Domain](sdtm-dm.qmd) has to
  be reconciled against.
- **`--SEQ`** — the sequence number — must be unique within a
  `USUBJID` within the domain, so that every record has a stable
  identity even when a subject has many (SDTMIG v3.4, §4.1.7).

### Variable roles, and the topic of a dataset

Every SDTM variable plays a **role**. The model defines five: Identifier,
Topic, Timing, Qualifier, and Rule — and Qualifiers subdivide further
into Grouping, Result, Synonym, Variable, and Record qualifiers (SDTMIG
v3.4, §2.1). The Rule role belongs to the trial-design datasets and
need not concern you while building subject data; the other four are
the working vocabulary of every domain:

- **Identifier** — who and which record (`USUBJID`, `--SEQ`).
- **Topic** — what the observation is *about*.
- **Timing** — when it happened (`--DTC`, `VISIT`).
- **Qualifier** — everything that describes or completes the result
  (units, severity, the result itself).

The **topic** variable is the heart of a dataset, and it differs by
class (SDTMIG v3.4, §2.1, §2.6): in Findings it is `--TESTCD` (the
thing measured), in Events it is `--TERM` (the verbatim event, with
`--DECOD` its dictionary-coded companion), and in Interventions it is
`--TRT` (the treatment). Identify the topic and you have identified
what the dataset is fundamentally a list of.

### Core: which variables you must include

Finally, every variable in a domain model carries a **core
designation** that says how obligatory it is (SDTMIG v3.4, §4.1.5):

- **Required** — must be in the dataset and must never be null on any
  record. These are the identifying and topic variables that make a
  record meaningful.
- **Expected** — must be in the dataset, but a value may be null on
  some records. If the study did not collect the item at all, the
  column is still included (empty) with an explanation in the
  Define-XML.
- **Permissible** — include it when the study collected the relevant
  data, and leave it out entirely when it did not.

"Required" and "Expected" both mean the column is always present; the
difference is whether a value is mandatory on every row. Confusing the
two is one of the most common conformance findings, and Example 2
depends on the distinction.

## Example 1: Anatomy of a Domain

*Estimated time: ~20 minutes (worked example)*

Take the five raw GLPX files and, using nothing but the ideas above,
answer three questions about each: which SDTM domain does it become,
which observation class is that domain in, and what is its topic
variable? This is the first thing a programmer does with a new study,
and it is done on paper before any code:

```{r classify}
#| eval: true

glpx_map <- tribble(
  ~raw_file,    ~domain, ~observation_class, ~topic_variable,
  "dm_raw.csv", "DM",    "Special-Purpose",  "(none — one row per subject)",
  "ex_raw.csv", "EX",    "Interventions",    "EXTRT",
  "ae_raw.csv", "AE",    "Events",           "AETERM",
  "lb_raw.csv", "LB",    "Findings",         "LBTESTCD",
  "vs_raw.csv", "VS",    "Findings",         "VSTESTCD"
)

glpx_map
```

Every domain code in that table is a claim about controlled
terminology, so check it rather than trust it. The SDTM Domain
Abbreviation codelist is `C66734`; ask the terminology package whether
each code is a real member:

```{r classify-ct}
#| eval: true

glpx_map |>
  mutate(valid_domain_code = map_lgl(domain, \(d) is_term(d, clst_code = "C66734"))) |>
  select(raw_file, domain, valid_domain_code)
```

::: {.callout-tip}
### Run It Yourself

Run both chunks. Every row of the second should read `TRUE`: `DM`,
`EX`, `AE`, `LB`, `VS` are all valid domain codes. Now change one code
to something plausible-but-wrong — `"LBS"`, say, or `"VITALS"` — and
re-run. Watch it come back `FALSE`. That is the whole discipline of
this course in one line: **you did not have to know the codelist, you
had to check it.**
:::

::: {.callout-note}
### Reading the Output Line by Line

| Column | What it tells you |
|---|---|
| `domain` | The two-character code each file becomes. It will be the dataset name, the `DOMAIN` value on every row, and the prefix on the domain's variables. |
| `observation_class` | Decided purely by what one row is. LB and VS are both Findings because each row is a measurement result; AE is Events because each row is something that happened; EX is Interventions because each row is a treatment given. |
| `topic_variable` | What the dataset is a list *of*. Notice the topic follows the class: `--TESTCD` for the two Findings domains, `--TERM` for the Events domain, `--TRT` for the Interventions domain. |
| `DM` row | Special-purpose, so it has no observation-class topic variable; its structure is simply one record per subject. You build it in [The DM Domain](sdtm-dm.qmd). |

Whatever the domain, four identifier variables form its spine —
`STUDYID`, `DOMAIN`, `USUBJID`, `--SEQ` — before a single
domain-specific column is added.
:::

## Example 2: Original Result vs Standardised Result

*Estimated time: ~20 minutes (worked example)*

This is where Findings pay off the modelling. A findings result is not
one value in one column; it is kept twice — as collected, and
standardised — and both are preserved. Take one subject's glucose
across visits from the raw lab file:

```{r ex2-raw}
#| eval: true

lb <- read_csv("data/raw/lb_raw.csv", show_col_types = FALSE,
               col_types = cols(.default = col_character()))

gluc <- lb |>
  filter(subjid == "101-037", test == "GLUC") |>
  select(subjid, visit, test, result, unit)

gluc
```

Now map those raw columns onto the SDTM findings result variables. The
result as collected goes to `LBORRES`, with its unit in `LBORRESU`
(SDTMIG v3.4, §4.5.1.1). The standardised character result goes to
`LBSTRESC`; where that value is numeric it is also written, as a true
number, to `LBSTRESN` (SDTMIG v3.4, §4.5.1.1). Glucose here is already
in the standard unit, so standardising is a copy — but the *structure*
is the lesson, not the arithmetic:

```{r ex2-map}
#| eval: true

mapped <- gluc |>
  transmute(
    USUBJID  = subjid,
    LBTESTCD = test,
    LBORRES  = result,              # result as originally collected
    LBORRESU = unit,                # original unit
    LBSTRESC = result,              # standardised, character
    LBSTRESN = as.numeric(result),  # standardised, numeric
    LBSTRESU = unit                 # standard unit
  )

mapped
```

Both the test code and the unit are controlled terminology, so confirm
them against the published codelists — `C65047` for the lab test code,
`C71620` for units — instead of assuming:

```{r ex2-ct}
#| eval: true

c(
  gluc_is_valid_testcd = is_term("GLUC",   clst_code = "C65047"),
  mmolL_is_valid_unit  = is_term("mmol/L", clst_code = "C71620")
)
```

::: {.callout-tip}
### Run It Yourself

Run the three chunks. Both CT checks return `TRUE` — the vendor's
`GLUC` and `mmol/L` happen to be legal SDTM values. But look carefully
at the mapped table's week-8 row before reading on. Something is
missing, and it is not the result.
:::

::: {.callout-note}
### Reading the Output Line by Line

| Variable | Value at week 8 | What it means |
|---|---|---|
| `LBORRES` | `9.6` | The result as collected. It is present. |
| `LBORRESU` | *(blank)* | The **unit is missing** — this is defect D7 from the trial simulator, a genuinely blank unit on one transferred record. |
| `LBSTRESC` / `LBSTRESN` | `9.6` / `9.6` | The result standardises fine; the *number* was never the problem. |
| `LBSTRESU` | *(blank)* | The missing unit propagates, because you cannot standardise a unit you do not have. |

`LBORRES` is an **Expected** variable — always present, occasionally
null (SDTMIG v3.4, §4.5.1.1). The unit's blankness is not a
model violation; it is a data problem the model makes *visible*. And
crucially, a blank is not a valid unit value:
:::

```{r ex2-blank}
#| eval: true

is_term("", clst_code = "C71620")
```

The check returns `FALSE`: an empty string is not a member of the UNIT
codelist. So the week-8 glucose cannot be silently accepted — its unit
has to be recovered from the lab's documentation or queried, a piece of
work you do in [Findings](sdtm-findings.qmd). The
point for now: **the standard did not fix the data, it located the
problem** — and it located it in a specific variable with a specific
rule, not in a vague sense that "something looks off."

## What Can Go Wrong

*Estimated time: ~10 minutes (reading)*

**Treating SDTM as "rename the columns"**

The recurring error from [Why Standards](why-standards.qmd), now with a
name. Renaming is a fraction of the work; classifying each dataset into
its observation class, giving each record the right identifier spine,
and constraining values to controlled terminology are the substance. If
your mental model is a lookup table of old name → new name, the DM and
findings sessions will not make sense.

**Overwriting the original result with the standardised one**

`--ORRES` and `--STRESC` are different variables on purpose (SDTMIG
v3.4, §4.5.1.1). The original is kept even when standardising changes
nothing, so that a reviewer can always see what was collected. Collapse
them into one column and you have destroyed traceability, which is one
of the things the whole tabulation layer exists to preserve.

**Inventing a test code or a unit**

Every `--TESTCD` and every unit is drawn from a published codelist. The
temptation to write `"HGBA1C"` or `"mmol/l"` because it looks right is
exactly the failure this course is built to prevent — check with
`is_term()` and let the package be right so you don't have to remember.

**Computing study day with a day zero**

Timing has its own trap, addressed in the next section's cross-links:
there is no study day 0 (SDTMIG v3.4, §4.4.4). More on this in [The DM
Domain](sdtm-dm.qmd), where `RFSTDTC` is set.

**Assuming one subject means one row, everywhere**

True only in DM, the special-purpose domain (SDTMIG v3.4, §5.2). In a
findings domain a subject has one row per test per visit; in an events
domain, one row per event. "How many rows should this subject have?" is
answered by the domain's observation class, never by a habit.

::: {.callout-warning}
### Common misinterpretations

**"SDTM is a database schema"**
It is a *submission* format: flat, one dataset per domain, built to be
read and archived by a regulator, not to be queried efficiently. Its
shape follows that purpose, not database-normalisation instincts.

**"Required and Expected are the same — both are mandatory"**
Both mean the column is always present. Only Required forbids a null
value on every row; an Expected variable may be null where the datum is
genuinely absent (SDTMIG v3.4, §4.1.5). The week-8 missing unit in
Example 2 is legal precisely because `LBORRESU` is not Required.

**"The `--` is part of the variable name"**
It is a placeholder the guide uses for the domain prefix. The real
variable is `LBTESTCD` or `VSTESTCD`; `--TESTCD` names the pattern
across domains (SDTMIG v3.4, §2.2).

**"If a value passes `is_term()`, the record is correct"**
It means the *value* is legal controlled terminology, nothing more. A
legal `LBTESTCD` on the wrong row, or a valid unit that is simply the
wrong unit, both pass. Conformance is necessary, not sufficient — the
lesson that closes [Why Standards](why-standards.qmd).
:::

## Exercises

### Exercise 1 (Guided): Classify Two More Domains

*Estimated time: ~10 minutes*

A later GLPX analysis will need two datasets this course has not built:
concomitant medications (what other drugs subjects were taking) and
disposition (how and when each subject left the study).

1. For each, name the general observation class you would put it in,
   and justify the choice from what *one row* represents.
2. Name the topic variable each would use (as a `--` pattern), based on
   its class.
3. Propose a plausible two-character domain code for each, then — this
   is the point — describe how you would check whether your proposed
   code is real, without stating from memory whether it is.

Worked solutions are in the paid tier; the habit the exercise builds
(class first, then topic, then *check* the code) is the whole method.

## Comprehension Check

*Estimated time: ~10 minutes*

1. What is the difference between the SDTM and the SDTMIG, and why does
   this course use v3.4 specifically?
2. A dataset has one row per subject per test per visit. Which
   observation class is it, and what is its topic variable?
3. `--STRESC` and `--STRESN` are both standardised results. Why are
   there two, and when is `--STRESN` null?
4. A variable is marked **Expected**. The study never collected that
   item. What must appear in the submitted dataset, and what must not?
5. You need the SDTM unit value for "beats per minute." A colleague
   says it is `"bpm"`. How do you settle it without trusting either of
   your memories?

::: {.callout-note collapse="true"}
### Answers

1. The SDTM is the general model (the kinds of boxes that exist); the
   SDTMIG is the implementation guide that turns it into named
   datasets and variables (SDTMIG v3.4, §2.1). The version is not a
   preference: the FDA Data Standards Catalog ties the required SDTMIG
   version to the study's start date, and GLPX-001's 2024 start falls
   in the v3.4 support window [@fdadsc11].

2. Findings — one row is the result of a measurement (SDTMIG v3.4,
   §2.3). Its topic variable is `--TESTCD` (SDTMIG v3.4, §2.1), i.e.
   `LBTESTCD` or `VSTESTCD` once the domain prefix is applied.

3. `--STRESC` holds the standardised result in character form, so it
   can carry any result at all; `--STRESN` holds it as a true number
   for the results that are numeric (SDTMIG v3.4, §4.5.1.1). `--STRESN`
   is null whenever the standardised result is not numeric (a category,
   a `>10,000`, a blank), which is why the character version exists.

4. The column must still be present, empty, with a comment in the
   Define-XML stating the study did not collect the item; a value is
   not required on any row (SDTMIG v3.4, §4.1.5). What must *not* happen
   is dropping the column — that is the difference between Expected and
   Permissible.

5. Check it against the UNIT codelist with the terminology package
   rather than debating: `is_term("bpm", clst_code = "C71620")` versus
   `is_term("beats/min", clst_code = "C71620")`. One is a member and one
   is not, and the package settles it in a line — the same move Example
   2 used for the glucose unit.
:::
