---
title: "SDTM: The DM Domain"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(sdtm.terminology)
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 100 minutes.**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background: the DM contract | 20 min | Reading |
| Example 1: Building DM from the raw files | 30 min | Worked example |
| Example 2: RFSTDTC and the study-day rule | 15 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Exercises | 15 min | Practice |
| Comprehension Check | 10 min | Self-test |

This is the busiest session in Part 2: it builds the course's first
real SDTM dataset, resolves three of the raw data's deliberate defects,
and sets the reference date that every later study-day calculation
uses. For self-paced study, split after Example 1.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown.
Where you see a "Run It Yourself" box, stop and run the code above it
before reading the explanation that follows. Run R from the project
root, as in earlier sessions.

Controlled-terminology checks use the **`sdtm.terminology`** package,
as introduced in [SDTM Concepts](sdtm-concepts.qmd).
:::

## When Do You Use This? {#sec-sdtm-dm}

::: {.callout-tip}
Every mapping effort starts here: DM is the first domain anyone builds,
because every other domain assumes it exists. You will also recognise
this session's problems personally by now — the demographics file with
401 rows for 400 subjects, and the subject whose sex differs between
the EDC and the lab file, have both been sitting unresolved since [Why
Standards](why-standards.qmd). And any question of the form "what study
day was that?" is unanswerable until someone sets each subject's
reference date.

So the question this session answers: **which dataset must exist,
correctly, before any other dataset can be built — and what does
building it actually involve?**

Builds on [SDTM Concepts](sdtm-concepts.qmd) (classes, cores,
controlled terminology) and [The Trial: GLPX-001](the-trial-glpx001.qmd)
(the raw files and their defects).
:::

## Learning Objectives

After completing this session you will be able to:

- Build a conformant DM dataset from raw EDC data, one record per
  subject
- Construct USUBJID and explain why its format is a sponsor decision
  with consequences
- Resolve a multi-site enrolment and a cross-source sex conflict into
  single DM values, and record the decisions
- Represent partial and missing birth dates at their true precision,
  without imputation
- Set RFSTDTC from first exposure and derive the reference-date pairs
- Compute study days across the reference date, including why there is
  no day zero
- Check every CT-governed DM value programmatically before submission

## Background: The DM Contract

*Estimated time: ~20 minutes (reading)*

::: {.callout-important}
### The one thing to remember

**DM is the contract: one row per subject that fixes who they are,
when their study clock starts, and what they were assigned — and every
other dataset inherits those answers.**
:::

### One row per person

Demographics (DM) is a special-purpose domain — one record per subject
— and the implementation guide calls it "the parent domain for all
other observations for human clinical subjects" (SDTMIG v3.4, §5.2).
*Parent* is meant structurally: every AE record, every lab result,
every dosing record in the study will carry a `USUBJID` that must
correspond to exactly one DM row. If DM is wrong, everything downstream
inherits the error.

The one-row rule has real teeth. The guide addresses our duplicated
subject directly: subjects occasionally change sites during a trial,
sponsors must decide how to populate `USUBJID`, `SUBJID` and `SITEID`
in that situation — "but only 1 DM record should be submitted for each
subject", with the Supplemental Qualifiers dataset available for
additional detail (SDTMIG v3.4, §5.2, assumption 2). Note what the rule
is really saying: **the DM row represents the person, not the enrolment
event.** A subject who moves sites is still one subject; the move is a
fact *about* them, not a second them.

### The identifier block

Four identifiers open the dataset, all Required (SDTMIG v3.4, §5.2):
`STUDYID`, `DOMAIN` (the constant `"DM"`), `USUBJID`, and `SUBJID`.
Two details are easy to miss:

- **`SUBJID` is DM's topic variable** (SDTMIG v3.4, §5.2) — the thing
  each row is *about* is the subject, as identified on the CRF. It must
  be unique within the study.
- **`USUBJID` has no mandated format.** The guide offers
  STUDYID-SITEID-SUBJID as one possibility, not a rule (SDTMIG v3.4,
  §5.2; §4.2.3). What *is* mandated is uniqueness across every study in
  the submission. Format is therefore a sponsor decision — and subject
  `101-012` is about to show why it has consequences.

### The subject's clock

DM carries a set of reference dates, and one of them matters more than
all the others. `RFSTDTC`, the subject reference start date, is
"usually equivalent to date/time when subject was first exposed to
study treatment" and is the anchor from which every study-day variable
in every domain is calculated (SDTMIG v3.4, §5.2, assumption 9). It is
null for subjects who were never treated — screen failures, subjects
never assigned (SDTMIG v3.4, §5.2, assumption 9) — which GLPX-001 does
not have, but the principle stands: the clock starts at first exposure,
so a subject with no exposure has no clock.

Around it sit the pairs (SDTMIG v3.4, §5.2): `RFENDTC` (the subject's
reference end — usually trial end, often last exposure; required for
all randomised subjects), `RFXSTDTC`/`RFXENDTC` (first and last
exposure to protocol-specified treatment, defined as the earliest
`EXSTDTC` and latest `EXENDTC`), `RFICDTC` (informed consent) and
`RFPENDTC` (end of participation — the last known date of contact).

One modelling subtlety: these look like timing variables, but in DM
they are classified as Record Qualifiers, because they are not intended
for use in the general observation classes (SDTMIG v3.4, §5.2,
assumption 7). The only true Timing variables DM may carry are
`DMDTC`/`DMDY` and, if needed, `VISITNUM`/`VISIT`/`VISITDY` (SDTMIG
v3.4, §5.2, assumption 8).

### The demographic facts

The rest of the row is the subject's stable description, each variable
with its core designation from the specification (SDTMIG v3.4, §5.2):

| Variable | Core | Governed by |
|---|---|---|
| `SITEID` | Required | sponsor-defined |
| `BRTHDTC` | Permissible | ISO 8601 |
| `AGE`, `AGEU` | Expected | AGEU codelist |
| `SEX` | **Required** | SEX codelist |
| `RACE` | Expected | RACE codelist |
| `ETHNIC` | Permissible | ETHNIC codelist |
| `COUNTRY` | **Required** | ISO 3166-1 alpha-3 |

Three notes with direct GLPX consequences. `BRTHDTC` is only
Permissible, with an explicit privacy caution (SDTMIG v3.4, §5.2,
assumption 3) — and partial birth dates are represented at their true
precision by right-truncating the ISO 8601 string, never by inventing
the missing part (SDTMIG v3.4, §4.4.2). `AGE` "may be derived from
RFSTDTC and BRTHDTC, but BRTHDTC may not be available in all cases"
(SDTMIG v3.4, §5.2) — so the partial-date subjects will force a
decision. And `SEX` is Required with controlled terminology: exactly
one legal value per subject, which is what turns the EDC-versus-lab
conflict from an awkward observation into something that must be
resolved. `COUNTRY` is Required and "generally represented using ISO
3166-1 Alpha-3" (SDTMIG v3.4, §5.2) — an external standard, not an NCI
codelist, so `sdtm.terminology` cannot check it and this course does
not pretend otherwise.

### The arms

Finally, treatment assignment: `ARM`/`ARMCD` (planned) and
`ACTARM`/`ACTARMCD` (actual), all Expected (SDTMIG v3.4, §5.2). The
codes are sponsor-defined, not controlled terminology: their values
must come from the Trial Arms (TA) dataset — the trial-design dataset
where a study declares its arms, which this course notes but does not
build — with ARM and ARMCD in a strict one-to-one relationship (SDTMIG
v3.4, §5.2, assumption 4). `ARMCD` is limited to 20 characters and is
free of the `--TESTCD` character restrictions (SDTMIG v3.4, §4.2.1).
`ARMNRS` exists for subjects whose arm variables are null — and may
**not** be populated when both ARMCD and ACTARMCD have values (SDTMIG
v3.4, §5.2, assumption 4a). In a trial like GLPX-001 where everyone was
randomised and treated as assigned, `ARMNRS` is a column of nulls —
present because it is Expected, empty because it has nothing to say.

## Example 1: Building DM from the Raw Files

*Estimated time: ~30 minutes (worked example)*

### One person, one row

Start where [The Trial](the-trial-glpx001.qmd) left off — 401 rows for
400 subjects:

```{r read-401}
#| eval: true

dm_raw <- read_csv("data/raw/dm_raw.csv", show_col_types = FALSE,
                   col_types = cols(.default = col_character()))

nrow(dm_raw)

dm_raw |> filter(SUBJID == "101-012")
```

Subject `101-012` was enrolled at site 101, moved, and was enrolled
again at site 108 three weeks later. Both rows record true events; the
person is nevertheless one person, and the standard is unambiguous:
one DM record per subject (SDTMIG v3.4, §5.2, assumption 2). We keep
one row — carrying the original enrolment's site — and note that the
site change itself is information *about* this subject that belongs in
a supplemental dataset, not in a second row:

```{r one-row}
#| eval: true

dm_person <- dm_raw |>
  arrange(SUBJID, RANDDT) |>
  distinct(SUBJID, .keep_all = TRUE)   # one row per person; site move -> SUPPDM

c(rows = nrow(dm_person), subjects = n_distinct(dm_person$SUBJID))
```

::: {.callout-tip}
### Run It Yourself

Run both chunks. You should see 401 collapse to 400, with rows and
subjects now equal. Before moving on, note what we did *not* do: we
did not delete a false record. Both raw rows were true. We decided —
as the standard requires the sponsor to decide — how one person's
site change is represented, and the answer is "in one row, with the
detail carried elsewhere."
:::

### Identifiers, and why USUBJID format matters

```{r identifiers}
#| eval: true

dm_ids <- dm_person |>
  mutate(
    STUDYID = "GLPX001",
    DOMAIN  = "DM",
    USUBJID = paste("GLPX001", SUBJID, sep = "-")
  )

dm_ids |> select(STUDYID, DOMAIN, USUBJID, SUBJID, SITEID) |> head(3)
```

::: {.callout-note}
### Reading the Output Line by Line

| Column | What it is |
|---|---|
| `STUDYID` | The study, constant. Sponsor-assigned. |
| `DOMAIN` | The constant `"DM"` — the two-character code doing one of its four jobs (SDTMIG v3.4, §2.2). |
| `USUBJID` | `GLPX001-101-012` — built as STUDYID + SUBJID, **deliberately not** from `SITEID`. |
| `SUBJID` | The subject as recorded on the CRF, unique within the study — DM's topic variable (SDTMIG v3.4, §5.2). |

The format decision is the lesson. The guide suggests
STUDYID-SITEID-SUBJID as one option (SDTMIG v3.4, §5.2) — and subject
`101-012` shows the trap in it: their site is now 108, and an
identifier rebuilt from current site would *change*, which is the one
thing an identifier must never do. We build from `SUBJID` alone. (The
`101` inside `101-012` is historical — it records where the number was
issued, like the area code of a phone number that moved. It is never
recomputed.)
:::

### The subject's clock, from the dosing records

`RFSTDTC` is usually the first exposure date (SDTMIG v3.4, §5.2,
assumption 9), and exposure lives in the raw dosing file. The
reference pairs come from the same place: `RFXSTDTC` is defined as the
earliest `EXSTDTC` and `RFXENDTC` as the latest `EXENDTC` (SDTMIG
v3.4, §5.2). GLPX-001 has one dosing record per subject, so earliest
and latest are simply its two dates:

```{r ref-dates}
#| eval: true

ex_raw <- read_csv("data/raw/ex_raw.csv", show_col_types = FALSE,
                   col_types = cols(.default = col_character()))

dm_ref <- dm_ids |>
  left_join(ex_raw |> select(SUBJID, EXSTDT, EXENDT), by = "SUBJID") |>
  mutate(
    RFSTDTC  = EXSTDT,   # clock starts at first exposure
    RFENDTC  = EXENDT,
    RFXSTDTC = EXSTDT,   # earliest exposure start
    RFXENDTC = EXENDT,   # latest exposure end
    RFICDTC  = NA_character_,   # consent date not collected -> null
    RFPENDTC = EXENDT    # last known contact in this study's data
  )

dm_ref |> select(USUBJID, RFSTDTC, RFENDTC, RFXSTDTC, RFXENDTC) |> head(3)
```

`RFICDTC` deserves a pause: GLPX-001's raw data never collected consent
dates, so the column is present and entirely null. That is the
Expected discipline from [SDTM Concepts](sdtm-concepts.qmd) in the
flesh — the column must exist (with the absence explained in the
Define-XML), but no value is invented (SDTMIG v3.4, §4.1.5).

### Birth dates at their true precision

Now defect D1. Four subjects' birth dates arrived incomplete, and the
standard's instruction is to represent reduced precision by
right-truncating the ISO 8601 string — not to fill in the missing part
(SDTMIG v3.4, §4.4.2):

```{r brthdtc}
#| eval: true

dm_birth <- dm_ref |> mutate(BRTHDTC = na_if(BIRTHDT, ""))

dm_birth |>
  filter(SUBJID %in% c("102-005", "110-023", "111-057", "103-141")) |>
  select(SUBJID, BIRTHDT, BRTHDTC)
```

::: {.callout-note}
### Reading the Output Line by Line

`1967-04` stays `1967-04` — a real ISO 8601 value meaning "April 1967,
day unknown." `1969` stays `1969`. The empty string becomes a true
null, the representation for a completely missing date (SDTMIG v3.4,
§4.4.2). Nothing was guessed. If an analysis later needs a complete
date, imputing one is an *analysis* decision that happens in the ADaM
layer with documented rules — [ADSL](adsl.qmd) picks this up. SDTM's
job is to preserve exactly what is known, no more.
:::

### AGE: derived where honesty allows

`AGE` is Expected, expressed in `AGEU`, and "may be derived from
RFSTDTC and BRTHDTC" (SDTMIG v3.4, §5.2). *May*, when the birth date
supports it. For the four D1 subjects it does not — deriving an age
from "1969" means inventing a birthday — so their AGE is honestly
null:

```{r age}
#| eval: true

dm_age <- dm_birth |>
  mutate(
    age_derivable = !is.na(BRTHDTC) & nchar(BRTHDTC) == 10,
    AGE  = if_else(age_derivable,
                   floor(as.numeric(as.Date(RFSTDTC) - as.Date(BRTHDTC)) / 365.25),
                   NA_real_),
    AGEU = if_else(age_derivable, "YEARS", NA_character_)
  )

dm_age |> count(age_derivable)

dm_age |> filter(!age_derivable) |> select(SUBJID, BRTHDTC, AGE, AGEU)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk. 396 subjects get a derived age; the four D1 subjects
get `NA` in both AGE and AGEU. Ask yourself which you would rather
defend to a reviewer: four honest nulls, or four ages computed from
invented birthdays. (In a real study, ages for such subjects are often
*collected* on the CRF precisely because the birth date is withheld —
GLPX-001's CRF did not, so nulls it is.)
:::

### SEX: one subject, one value

Defect D3 has been waiting since [Why Standards](why-standards.qmd).
`SEX` is Required with controlled terminology (SDTMIG v3.4, §5.2) —
one legal value per subject — and we have a subject with two:

```{r sex-check}
#| eval: true

lb_raw <- read_csv("data/raw/lb_raw.csv", show_col_types = FALSE,
                   col_types = cols(.default = col_character()))

dm_age |>
  select(SUBJID, edc_sex = SEX) |>
  inner_join(lb_raw |> distinct(subjid, lab_sex = sex),
             by = join_by(SUBJID == subjid)) |>
  filter(edc_sex != lab_sex)
```

The data cannot arbitrate this; a rule about sources can. The EDC
value is what the site — the people with the subject and the source
documents in front of them — entered; the lab file's copy is
downstream demographic data supplied *to* the vendor along with the
samples. For a stable demographic fact, the site-entered record is the
authoritative source, so DM takes the EDC value, and the disagreement
is logged as a data query to the lab rather than silently absorbed:

```{r sex-resolve}
#| eval: true

dm_queries <- tribble(
  ~USUBJID,          ~variable, ~issue,
  "GLPX001-111-003", "SEX",     "EDC records F; central-lab file records M"
) |>
  mutate(resolution = paste(
    "SEX taken from the EDC (site-entered source);",
    "discrepancy sent to the lab as a data query"
  ))

dm_queries |> select(USUBJID, issue, resolution)

dm_age |> count(SEX)
```

The DM column itself needed no edit — the EDC value was already `F` —
but do not mistake that for nothing having happened. A decision was
made, by rule, and recorded. The alternative (quietly preferring
whichever file was loaded last) produces datasets whose values nobody
can later explain.

### The remaining demographics, checked not trusted

`RACE`, `ETHNIC` and `AGEU` are controlled terminology; `SEX` we have
just resolved. Check every value the dataset actually contains against
the published codelists — the habit from [SDTM
Concepts](sdtm-concepts.qmd), now at dataset scale:

```{r ct-demog}
#| eval: true

ct_checks <- tibble(
  variable = c("SEX", "RACE", "ETHNIC", "AGEU"),
  codelist = c("C66731", "C74457", "C66790", "C66781"),
  values   = list(unique(dm_age$SEX), unique(dm_age$RACE),
                  unique(dm_age$ETHNIC), unique(na.omit(dm_age$AGEU)))
) |>
  mutate(all_valid = map2_lgl(values, codelist,
                              \(v, cl) all(map_lgl(v, \(x) is_term(x, clst_code = cl)))))

ct_checks |> mutate(values = map_chr(values, paste, collapse = ", "))
```

::: {.callout-note}
### Reading the Output Line by Line

All four rows come back `TRUE` — including RACE's `OTHER`, which is a
genuine member of the RACE codelist (`C74457`). GLPX-001's EDC was
configured with codelist values, so the checks pass; in a study whose
CRF allowed free text they would not, and this one table would be
where you found out. Passing is not the point. *Having checked* is
the point — the alternative is stating codelist contents from memory,
which is how plausible-but-wrong values enter submissions.

`COUNTRY` (`DNK`, `DEU`, `GBR`, `ESP`, `POL`) is governed by ISO
3166-1 alpha-3 (SDTMIG v3.4, §5.2), an external standard the
terminology package does not carry — so it is *not* in this table,
and its check is a different one (against the ISO list) rather than
pretending `is_term()` covers it.
:::

### Arms

The raw file carries the randomised arm's name; the codes are ours to
define — sponsor-defined values that must come from the Trial Arms
dataset with ARM and ARMCD one-to-one (SDTMIG v3.4, §5.2, assumption
4), ARMCD within 20 characters (SDTMIG v3.4, §4.2.1). Every GLPX
subject was treated as randomised, so actual mirrors planned, and
`ARMNRS` — which may not be populated when both code variables have
values (SDTMIG v3.4, §5.2, assumption 4a) — stays null:

```{r arms}
#| eval: true

dm_arms <- dm_age |>
  mutate(
    ARMCD    = case_when(ARM == "GLPX 10 mg" ~ "GLPX10",
                         ARM == "Placebo"    ~ "PBO"),
    ACTARM   = ARM,
    ACTARMCD = ARMCD,
    ARMNRS   = NA_character_,
    ACTARMUD = NA_character_,
    DTHDTC   = NA_character_,   # no deaths in GLPX-001
    DTHFL    = NA_character_
  )

dm_arms |> count(ARMCD, ARM, ACTARMCD, ACTARM)
```

### Assemble and check

Put the columns in the specification's order and let the dataset prove
itself — one row per subject, Required variables never null, CT values
legal:

```{r assemble}
#| eval: true

dm <- dm_arms |>
  select(STUDYID, DOMAIN, USUBJID, SUBJID,
         RFSTDTC, RFENDTC, RFXSTDTC, RFXENDTC, RFICDTC, RFPENDTC,
         DTHDTC, DTHFL, SITEID, BRTHDTC, AGE, AGEU, SEX, RACE, ETHNIC,
         ARMCD, ARM, ACTARMCD, ACTARM, ARMNRS, ACTARMUD, COUNTRY)

required_vars <- c("STUDYID", "DOMAIN", "USUBJID", "SUBJID",
                   "SITEID", "SEX", "COUNTRY")

c(
  one_row_per_subject  = nrow(dm) == n_distinct(dm$USUBJID),
  required_never_null  = all(map_lgl(required_vars, \(v) !any(is.na(dm[[v]])))),
  domain_code_valid_ct = is_term("DM", clst_code = "C66734"),
  expected_cols_present = all(c("ARMNRS", "RFICDTC", "DTHFL") %in% names(dm))
)
```

::: {.callout-tip}
### Run It Yourself

Run the two chunks: four checks, four `TRUE`s, and `dm` is a 400-row,
26-column DM dataset. Notice which columns you built by *copying*
(SITEID, RACE), which by *deciding* (the one-row rule, USUBJID's
format, SEX's source, AGE's nulls), and which by *deriving* (the
reference dates). The copying was trivial. The dataset's value is
entirely in the deciding — which is why those decisions were made
explicitly and, in a real study, live in the study's documentation.

Also absent, deliberately: the Permissible variables the study never
collected (`INVID`, `INVNAM`, `DMDTC`, `DMDY`) are left out entirely
— included-if-collected is exactly what Permissible means (SDTMIG
v3.4, §4.1.5) — while the Expected-but-empty columns (`RFICDTC`,
`DTHFL`, `ARMNRS`) are present and null. The difference between those
two fates is the difference between Permissible and Expected.
:::

## Example 2: RFSTDTC and the Study-Day Rule

*Estimated time: ~15 minutes (worked example)*

DM now gives every subject a clock. [The
Trial](the-trial-glpx001.qmd) promised the precise rule for reading
it, and here it is, verbatim from the guide (SDTMIG v3.4, §4.4.4): the
reference start date is study day 1; the study day increments by 1 for
each date after it; dates before it count down from **−1**; **there is
no study day 0**. As formulas:

- on or after the reference date:
  `--DY = (date portion of --DTC) − (date portion of RFSTDTC) + 1`
- before the reference date:
  `--DY = (date portion of --DTC) − (date portion of RFSTDTC)`

Watch what that does across the boundary, for subject `101-026`
(RFSTDTC = 2024-05-06):

```{r jump-demo}
#| eval: true

ref <- dm |> filter(SUBJID == "101-026") |> pull(RFSTDTC) |> as.Date()

study_day <- function(date, ref) {
  d <- as.integer(date - ref)
  if_else(date >= ref, d + 1L, d)
}

tibble(date = ref + (-2:2)) |>
  mutate(diff_in_days = as.integer(date - ref),
         DY           = study_day(date, ref))
```

::: {.callout-note}
### Reading the Output Line by Line

| Date | Difference | Study day |
|---|---|---|
| two days before | −2 | **−2** |
| one day before | −1 | **−1** |
| the reference date | 0 | **1** |
| one day after | +1 | **2** |

The day sequence runs −2, −1, 1, 2 — the calendar difference of 0
never appears as a study day. This mirrors how people naturally count
("day 1 of treatment" is the first day, not day 0), and it has a hard
consequence the guide states itself: study days are unusable for
duration arithmetic, precisely because the scale has a missing number.
Durations are computed from the raw dates, never from `--DY` (SDTMIG
v3.4, §4.4.4).
:::

Now the same function against real data — the subject's HbA1c
collection dates from the raw lab file:

```{r real-days}
#| eval: true

lb_raw |>
  filter(subjid == "101-026", test == "HBA1C") |>
  mutate(colldt = as.Date(colldt),
         LBDY   = study_day(colldt, ref)) |>
  select(visit, colldt, LBDY)
```

::: {.callout-tip}
### Run It Yourself

Run both chunks. The screening draw lands on study day **−15** —
before the clock started, so negative, with no zero crossed on the way
— and the baseline draw on day **2**, because the sample was taken the
day after first dose. Nominal week 26 lands on day 184. All three are
the reality behind the tidy visit labels: `--DY` records when things
*actually* happened on the subject's own clock. When you build full
findings domains in [Findings](sdtm-findings.qmd), this function is
applied to every dated record in the study.
:::

## What Can Go Wrong

*Estimated time: ~10 minutes (reading)*

**Keeping both rows for a re-enrolled subject "to be safe"**

Safety is exactly what it isn't. Two DM rows for one person breaks the
parent-domain contract — joins duplicate, counts inflate, and every
downstream domain inherits the ambiguity. The standard's rule is one
record per subject with the detail carried supplementally (SDTMIG
v3.4, §5.2, assumption 2). Keeping both rows is not caution; it is
deferring a decision to everyone who uses the data after you.

**Imputing partial birth dates in SDTM**

Turning `1969` into `1969-07-01` feels helpful and destroys
information: the dataset now asserts a birthday nobody stated, and no
reader can tell it apart from a real one. SDTM represents reduced
precision honestly by truncation (SDTMIG v3.4, §4.4.2); imputation for
analysis is done, with documented rules, in the analysis datasets.

**Resolving a source conflict by tidiness**

The lab file is machine-formatted and consistent; the temptation is to
trust it. The resolution rule must come from *provenance*, not
formatting: for a stable demographic, the site-entered source record
outranks a downstream copy. Whatever the rule, it is applied
consistently and the conflict is logged as a query — a dataset whose
values cannot be traced to a decision is unreviewable.

**Building USUBJID from mutable facts**

Any identifier constructed from a value that can change — site, arm,
name fragments — will eventually change, and an identifier that
changes is not an identifier. Subject `101-012` is the standing
counterexample to STUDYID-SITEID-SUBJID in this study.

**Study-day arithmetic with a day 0**

`as.integer(date - ref)` alone is off by one for every on-study
record, and `+ 1` everywhere is off by one for every pre-study record.
The rule is two-branched because the scale skips zero (SDTMIG v3.4,
§4.4.4) — and the same section is why `--DY` must never be used to
compute durations.

**Dropping the empty Expected columns**

`ARMNRS`, `RFICDTC` and `DTHFL` are all null in GLPX-001's DM, and a
tidy instinct says remove them. Expected means the column stays, with
the study-level absence explained in the metadata (SDTMIG v3.4,
§4.1.5). Only Permissible variables disappear when uncollected.

::: {.callout-warning}
### Common misinterpretations

**"DM holds whatever was on the demographics CRF page"**
DM holds the standard's set of subject-level variables, several of
which (the reference dates) come from *other* raw sources entirely —
GLPX's clock came from the dosing file. The CRF page and the domain
are different units of organisation.

**"USUBJID and SUBJID are basically the same thing"**
`SUBJID` is the subject as the study knew them, unique within the
study. `USUBJID` is the person across every study of the product,
unique across the whole submission (SDTMIG v3.4, §5.2; §4.2.3). The
same person in a later GLPX trial would get a new SUBJID and must get
the same USUBJID.

**"AGE missing for four subjects is a data quality problem"**
It is the honest representation of what is known. The quality problem
would be four plausible ages derived from invented birthdays. Null,
here, is the *correct value*.

**"The subject's study day 0 is their randomisation day"**
There is no study day 0 (SDTMIG v3.4, §4.4.4) — the reference date is
day 1, and GLPX-001's reference date is first *exposure*, which for
some subjects is not the randomisation day at all.
:::

## Exercises

The DM exercise set is in [Exercise 1 — SDTM: The DM
Domain](../exercises/sdtm-dm.qmd): finishing the build for a column
subset, defending the AGE nulls, and writing the decision log for the
two resolved conflicts. Solutions are in the paid tier.

## Comprehension Check

*Estimated time: ~10 minutes*

1. Subject `101-012` has two genuine enrolment records. Why does DM get
   exactly one row, and where does the site-change information belong?
2. This course built `USUBJID` as `GLPX001-<SUBJID>` rather than the
   guide's STUDYID-SITEID-SUBJID example. Justify the choice.
3. `RFICDTC`, `DTHFL` and `ARMNRS` are all completely null in the
   finished dataset, while `INVNAM` and `DMDTC` are absent entirely.
   Explain both fates.
4. A colleague fills subject `111-057`'s birth date (`1969`) in as
   `1969-06-15` so that AGE can be computed. What has gone wrong, and
   where would completing a date be legitimate?
5. A lab sample was collected the day before a subject's first dose,
   and another on the day of first dose. What are the two study days,
   and what rule produced them?

::: {.callout-note collapse="true"}
### Answers

1. The DM row represents the person, not the enrolment event, and the
   guide requires exactly one DM record per subject even when sites
   change (SDTMIG v3.4, §5.2, assumption 2). Both raw rows are true;
   the representation decision is the sponsor's, with the site change
   carried in the Supplemental Qualifiers dataset rather than a second
   row.

2. An identifier must never change, and site can (subject `101-012`
   moved from 101 to 108). Building from study + within-study subject
   number uses only immutable facts. The guide mandates uniqueness
   across the submission, not a format (SDTMIG v3.4, §5.2; §4.2.3), so
   this is a legitimate sponsor decision — made and documented once.

3. The null trio are **Expected** variables: the column must be present
   even when the study has nothing to put in it, with the absence
   explained in the metadata. The missing pair are **Permissible**
   variables the study did not collect, and uncollected Permissible
   variables are omitted entirely (SDTMIG v3.4, §4.1.5). Same emptiness,
   opposite obligations.

4. The dataset now asserts a birthday that was never stated, and
   nothing distinguishes it from a real one — traceability to source is
   broken. SDTM represents partial dates at their true precision by
   truncation (SDTMIG v3.4, §4.4.2). Completing a date is legitimate as
   a documented *imputation* in the analysis layer (ADaM), where the
   rule is stated and the imputed value is flagged as such.

5. Day **−1** and day **1**. The reference date (first dose, RFSTDTC)
   is study day 1 and the day before it is −1, because there is no
   study day 0 (SDTMIG v3.4, §4.4.4): before the reference date the
   study day is the plain date difference; on or after it, the
   difference plus one.
:::
