---
title: "SDTM: Events (Adverse Events)"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(sdtm.terminology)
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 75 minutes.**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background: what an Events row is | 15 min | Reading |
| Example 1: From raw events to AE rows | 20 min | Worked example |
| Example 2: Qualifying and timing the event | 20 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Exercises + Comprehension Check | 15 min | Practice / Self-test |

This is the first of two domain-building sessions in the general
observation classes; [Findings](sdtm-findings.qmd) is its sibling. It
fits one sitting. It resolves one of the raw data's deliberate defects
(an adverse event with no start date).
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown.
Where you see a "Run It Yourself" box, stop and run the code above it
before reading the explanation that follows. Run R from the project
root, as in earlier sessions.

Controlled-terminology checks use the **`sdtm.terminology`** package,
as in [SDTM Concepts](sdtm-concepts.qmd).
:::

## When Do You Use This? {#sec-sdtm-events}

::: {.callout-tip}
A safety reviewer asks "how many serious adverse events were there?"
and you need to know that seriousness is a regulatory category, not a
measure of how bad the event felt. A site reports an event by phone
and never sends a start date. A subject reports no adverse events at
all — and you have to know that "none" is represented by *no row*, not
a row that says none.

All three are the same underlying task: representing *things that
happened to subjects*. This session asks: **once you know a dataset is
about occurrences rather than measurements, how do you build it?**

Builds on [SDTM Concepts](sdtm-concepts.qmd) (the three observation
classes; controlled terminology) and [The DM Domain](sdtm-dm.qmd)
(RFSTDTC and the study-day rule, both reused here). Its sibling is
[Findings](sdtm-findings.qmd).
:::

## Learning Objectives

After completing this session you will be able to:

- Recognise an Events-class domain from what one row represents, and
  predict its key variables
- Build an AE dataset from raw events, keeping the verbatim term and
  naming (not faking) the dictionary-coding step
- Distinguish seriousness from severity and represent each correctly
- Map raw outcome text to controlled terminology, and check it
- Handle an adverse event with no start date the way the standard
  intends
- Derive event study days from the subject's reference date

## Background: What an Events Row Is

*Estimated time: ~15 minutes (reading)*

::: {.callout-important}
### The one thing to remember

**An Events row records that something happened — carried by a verbatim
term and a timing. Everything else in the row qualifies that
occurrence; it never turns the row into a measurement.**
:::

Adverse Events (AE) is an Events-class domain, and its structure is
**one record per adverse event per subject** (SDTMIG v3.4, §6.2.1). The
sponsor defines what counts as one event, but the unit is the event: a
subject with three events has three rows, a subject with none has no
rows at all.

The **topic** variable — the thing each row is about — is `AETERM`, the
Required, verbatim name of the event exactly as it was reported (SDTMIG
v3.4, §6.2.1, assumption 2a). "Verbatim" is doing real work: `AETERM`
holds what the site actually wrote, messiness and all, and it is never
cleaned in place. The cleaned, standardised version lives in a
*separate* variable, `AEDECOD` — the dictionary-derived preferred term,
also Required, produced by coding each verbatim term against a medical
dictionary, normally MedDRA (SDTMIG v3.4, §6.2.1, assumption 2c).

::: {.callout-warning}
### This course stops at the verbatim term

Coding `AETERM` into `AEDECOD` requires **MedDRA**, a licensed external
dictionary that is not part of CDISC Controlled Terminology and not
available in the `sdtm.terminology` package. This course therefore
builds `AETERM` honestly and leaves `AEDECOD` empty, *naming* the
coding step rather than performing it. It will never invent a preferred
term, because a plausible-looking MedDRA code that is wrong is exactly
the kind of silent error the whole course is built to avoid. In a real
study, coding is done by trained coders (or a supervised process) and
the dictionary name and version are recorded in the Define-XML.
:::

Around the topic sit **qualifiers** — and two of them are constantly
confused. `AESEV` is the event's *severity* (its intensity: MILD,
MODERATE, SEVERE). `AESER` is whether the event is *serious* — a
regulatory category with a Y/N answer, defined by criteria like death,
hospitalisation, or life-threatening risk. They are independent: a
severe headache may not be serious, and a "mild" event can be serious.
Getting these two straight is the single most important idea in AE
data, and Example 2 shows the trap directly.

Finally, **timing**. `AESTDTC` and `AEENDTC` are the ISO 8601 start and
end of the event — and both are *Expected*, meaning the column is
always present but a value may be missing (SDTMIG v3.4, §6.2.1). That
core designation is not a technicality: it is exactly what lets the
standard represent an event whose start date was never captured.
Study-day versions (`AESTDY`, `AEENDY`) are derived from the DM
reference date using the rule from [The DM Domain](sdtm-dm.qmd).

## Example 1: From Raw Events to AE Rows

*Estimated time: ~20 minutes (worked example)*

Start with the raw adverse-event file and the DM dataset you built and
saved in the previous session:

```{r read}
#| eval: true

ae_raw <- read_csv("data/raw/ae_raw.csv", show_col_types = FALSE,
                   col_types = cols(.default = col_character()))

# DM, built in the previous session and saved to data/sdtm/ (the build
# script that produces it is covered in the Findings session)
dm <- read_csv("data/sdtm/dm.csv", show_col_types = FALSE,
               col_types = cols(.default = col_character()))

c(ae_rows = nrow(ae_raw),
  subjects_with_events = n_distinct(ae_raw$SUBJID),
  subjects_total = nrow(dm))
```

::: {.callout-note}
### Reading the Output Line by Line

389 event rows across 262 subjects, out of 400 enrolled. The gap — 138
subjects — is the point: those subjects have **no AE rows**, and that
is the correct representation of "no adverse events." A dataset never
carries a row that says a subject had none; absence is the record
(SDTMIG v3.4, §6.2.1, assumption 4d). If you counted rows expecting
400, this is where you would learn that Events data is one-row-per-
occurrence, not one-row-per-subject.
:::

Now build the identifier spine and the topic. `USUBJID` comes from DM —
the AE file only knows the study's `SUBJID`, so DM is where the
across-submission identifier is joined in. `AESEQ` numbers each
subject's events. `AETERM` is copied verbatim; `AEDECOD` is created and
left empty for the coding step:

```{r build-ae}
#| eval: true

ae <- ae_raw |>
  left_join(dm |> select(SUBJID, USUBJID, RFSTDTC), by = "SUBJID") |>
  arrange(USUBJID, AESTDT, AETERM) |>
  group_by(USUBJID) |>
  mutate(AESEQ = row_number()) |>
  ungroup() |>
  mutate(STUDYID = "GLPX001", DOMAIN = "AE",
         AEDECOD = NA_character_)   # MedDRA preferred term — coding step, not done here

ae |>
  filter(USUBJID == "GLPX001-101-037") |>
  select(USUBJID, AESEQ, AETERM, AEDECOD)
```

::: {.callout-tip}
### Run It Yourself

Run both chunks. Subject `101-037` should come back with two events,
`AESEQ` 1 and 2, each with a verbatim `AETERM` and an empty `AEDECOD`.
Look back at [Why Standards](why-standards.qmd): the raw `AETERM`
column is where `Nausea`, `NAUSEA`, `Diarrea` and `nausea and
vomitting` all live. Every one of those is preserved here exactly as
written — because `AETERM` is verbatim, and fixing the spelling is the
coder's job, recorded in `AEDECOD`, not ours to guess.
:::

## Example 2: Qualifying and Timing the Event

*Estimated time: ~20 minutes (worked example)*

### Severity is not seriousness

The raw file carries both the intensity of each event and whether it
was serious. Map them straight across — `AESEV` for severity, `AESER`
for the serious flag — then look at what they say together:

```{r sev-ser}
#| eval: true

ae <- ae |>
  mutate(AESEV = AESEV,    # MILD / MODERATE / SEVERE (intensity)
         AESER = AESER)    # Y / N (regulatory seriousness)

ae |> count(AESEV, AESER)
```

::: {.callout-note}
### Reading the Output Line by Line

Read the cross-tabulation carefully. There are `SEVERE` events that are
**not** serious (`AESER = "N"`), and exactly one event that **is**
serious (`AESER = "Y"`). Severity and seriousness are different axes:
severity is how intense the event was; seriousness is whether it met a
regulatory criterion (here, a hypoglycaemia that required
hospitalisation). Most severe events are not serious, and — though
GLPX-001 has no such case — a mild event *could* be serious. Confusing
the two is how a safety table ends up with the wrong SAE count.
:::

Look at the one serious event directly, and confirm both qualifier
columns hold only legal controlled-terminology values:

```{r sae}
#| eval: true

ae |> filter(AESER == "Y") |> select(USUBJID, AETERM, AESEV, AESER)

# AESEV -> codelist C66769; AESER -> the NY (No/Yes) codelist C66742
c(
  aesev_valid = all(map_lgl(unique(ae$AESEV), \(x) is_term(x, clst_code = "C66769"))),
  aeser_valid = all(map_lgl(unique(ae$AESER), \(x) is_term(x, clst_code = "C66742")))
)
```

### Outcome: a term that needs mapping

The raw outcome column reads `RECOVERED`, but that is *not* a legal
value of the AE outcome codelist — the published term is
`RECOVERED/RESOLVED`. This is the controlled-terminology discipline
from [SDTM Concepts](sdtm-concepts.qmd) in miniature: a raw string that
looks right is not automatically a codelist member, so it is mapped and
the mapping is checked:

```{r aeout}
#| eval: true

ae <- ae |>
  mutate(AEOUT = case_when(
    AEOUT == "RECOVERED"     ~ "RECOVERED/RESOLVED",
    AEOUT == "NOT RECOVERED" ~ "NOT RECOVERED/NOT RESOLVED"
  ))

c(
  raw_recovered_valid = is_term("RECOVERED", clst_code = "C66768"),
  mapped_valid        = is_term("RECOVERED/RESOLVED", clst_code = "C66768")
)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk. The raw `RECOVERED` returns `FALSE` — it is not in the
outcome codelist — while the mapped `RECOVERED/RESOLVED` returns `TRUE`.
Every GLPX-001 event resolved, so after mapping, `AEOUT` is a single
value across the file; there are no ongoing events to represent. (Had
there been, an event still ongoing at last contact would carry a null
`AEENDTC`; the relative-timing variable `AEENRF` exists for expressing
"ongoing" against the study reference period, per SDTMIG v3.4, §4.4.7,
but this trial gives us no occasion to use it.)
:::

### Timing, and the event with no start date

Now the dates — and defect D4. Map `AESTDTC`/`AEENDTC` from the raw
start and end, turning blanks into true nulls, and derive the study
day from each subject's `RFSTDTC` with the same rule as [The DM
Domain](sdtm-dm.qmd):

```{r timing}
#| eval: true

study_day <- function(date, ref) {
  date <- as.Date(date); ref <- as.Date(ref)
  d <- as.integer(date - ref)
  if_else(is.na(date) | is.na(ref), NA_integer_,
          if_else(date >= ref, d + 1L, d))
}

ae <- ae |>
  mutate(
    AESTDTC = na_if(AESTDT, ""),
    AEENDTC = na_if(AEENDT, ""),
    AESTDY  = study_day(AESTDTC, RFSTDTC),
    AEENDY  = study_day(AEENDTC, RFSTDTC)
  )

ae |>
  filter(is.na(AESTDTC)) |>
  select(USUBJID, AETERM, AESTDTC, AEENDTC, AESTDY, AEENDY)
```

::: {.callout-note}
### Reading the Output Line by Line

One event comes back — subject `103-199`'s constipation — with a null
`AESTDTC`. It was reported by phone and the start date was never
obtained (defect D4). The standard handles this without complaint,
because `AESTDTC` is **Expected**, not Required: the column is present,
and this one value is legitimately null (SDTMIG v3.4, §6.2.1). The
study day `AESTDY` is null too — you cannot compute days-from-reference
without a date — while `AEENDTC` is present, so `AEENDY` is not. Note
what the standard did *not* make you do: it did not force you to invent
a start date to fill the column, and it did not make you drop the
event. The missing value is carried honestly, exactly as with the
partial birth dates in DM.
:::

## What Can Go Wrong

*Estimated time: ~10 minutes (reading)*

**Conflating seriousness and severity**

The most consequential AE error there is. A count of `AESEV = "SEVERE"`
is not a count of serious adverse events; the SAE count comes from
`AESER = "Y"`. In GLPX-001 the two numbers are wildly different — many
severe events, one serious one — and reporting the wrong column
misstates the trial's safety profile.

**Inventing preferred terms**

Filling `AEDECOD` by guessing what a verbatim term "should" code to,
rather than leaving it for the dictionary step, plants exactly the kind
of authoritative-looking error a reviewer cannot catch. If you do not
have MedDRA and a coding process, `AEDECOD` stays empty — and the
Define-XML says so.

**A row that says "no events"**

If a subject had no adverse events, they get no AE rows. A row created
to state "none" is not just unnecessary; it is wrong, and it will
inflate every event count and every join (SDTMIG v3.4, §6.2.1,
assumption 4d).

**Cleaning `AETERM` in place**

Correcting `Diarrea` to `Diarrhoea` in `AETERM` destroys the verbatim
record. The verbatim term is a source value; the corrected concept
belongs in `AEDECOD`, produced by coding. Keep both.

**Copying raw outcome text without checking it**

`RECOVERED` is not a codelist value. Carrying raw text straight into a
CT-governed variable produces data that fails conformance — the map to
`RECOVERED/RESOLVED` is not optional.

::: {.callout-warning}
### Common misinterpretations

**"389 adverse events means the drug caused 389 problems"**
It means 389 were *recorded*, across both arms, most unrelated to
treatment. Counts describe what was captured, not what the drug did —
causality is a separate variable (`AEREL`) and a separate analysis.

**"A missing start date is a data-quality failure to fix"**
It is a fact about collection, carried honestly in an Expected
variable. The failure would be inventing a date to make the column
look complete.

**"Severe and serious are synonyms"**
Severity (`AESEV`) is intensity; seriousness (`AESER`) is a regulatory
category. They vary independently, and the SAE count depends only on
`AESER`.

**"AETERM should be tidied so the counts come out right"**
Tidying belongs in `AEDECOD` via coding, which maps `Nausea`, `NAUSEA`
and `nausea and vomitting` to their preferred terms while preserving
the verbatim originals. Editing `AETERM` throws away the source.
:::

## Exercises

The AE exercise set is in [Exercise — SDTM:
Events](../exercises/sdtm-events.qmd): completing the domain slice,
deriving the study days, and articulating the severity/seriousness
distinction on real rows. Solutions are in the paid tier.

## Comprehension Check

*Estimated time: ~10 minutes*

1. A subject had no adverse events. How many AE rows do they have, and
   why?
2. `AETERM` for one event reads `nausea and vomitting`. What do you put
   in `AETERM`, what in `AEDECOD`, and why does this course leave the
   second empty?
3. GLPX-001 has many `SEVERE` events and one `AESER = "Y"` event.
   Explain, to a colleague who thinks they are the same thing, how both
   can be true.
4. Subject `103-199`'s constipation has a null `AESTDTC`. Why is the
   dataset still conformant, and what must you *not* do to "fix" it?
5. The raw outcome value is `RECOVERED`. Why can it not go straight into
   `AEOUT`, and how do you find the value that can?

::: {.callout-note collapse="true"}
### Answers

1. Zero rows. Adverse-event data is one record per event per subject,
   and "no events" is represented by the absence of records, never by a
   row asserting none (SDTMIG v3.4, §6.2.1, assumption 4d). Creating a
   "none" row would inflate counts and joins.

2. `AETERM` holds the verbatim string `nausea and vomitting` exactly as
   reported — it is a Required, verbatim topic variable (SDTMIG v3.4,
   §6.2.1, assumption 2a). `AEDECOD` would hold the dictionary-derived
   preferred term(s) from coding (SDTMIG v3.4, §6.2.1, assumption 2c).
   This course leaves `AEDECOD` empty because coding needs MedDRA, a
   licensed dictionary it does not use — and guessing a preferred term
   would risk a confident, invisible error.

3. Severity (`AESEV`) measures how intense the event was; seriousness
   (`AESER`) is a yes/no regulatory category based on criteria such as
   hospitalisation or life-threatening risk. They are independent axes,
   so a study can have many intense-but-not-serious events and very few
   serious ones. The SAE count comes only from `AESER = "Y"`.

4. `AESTDTC` is an Expected variable: the column must be present, but a
   value may be null (SDTMIG v3.4, §6.2.1). The start date was never
   collected, so null is the honest value, and `AESTDY` is null in
   consequence. You must not invent a start date to fill the column,
   and you must not drop the event — both would misrepresent what was
   recorded.

5. `RECOVERED` is not a member of the AE outcome codelist; the published
   value is `RECOVERED/RESOLVED`, confirmed with
   `is_term("RECOVERED/RESOLVED", clst_code = "C66768")`. Controlled-
   terminology variables take only published values, so the raw string
   is mapped to the codelist term, and the mapping is checked with the
   package rather than assumed.
:::
