---
title: "The Trial: GLPX-001"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
source("R/palette.R")   # course colours (Okabe-Ito); see R/palette.R
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 85 minutes.**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background | 20 min | Reading |
| Example 1: Predict the Data Before Opening It | 20 min | Worked example |
| Example 2: Reading the Trial's Story in the Raw Data | 20 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Comprehension Check | 10 min | Self-test |

For self-paced study this splits naturally into two sittings:
Background and Example 1 first, the rest later. Every later session
uses this trial, so time spent here repays itself for the remainder of
the course.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown.
Where you see a "Run It Yourself" box, stop and run the code above it
before reading the explanation that follows.

As in [Why Standards](why-standards.qmd), run R from the project root,
so that `data/raw/dm_raw.csv` means what it says.
:::

## When Do You Use This? {#sec-the-trial}

::: {.callout-tip}
A data transfer arrives from the lab vendor and someone asks you: "is
it complete?" Or you are handed a study's raw data on your first day
and need to know what you are looking at before the Monday meeting. Or
a listing you produced has 181 subjects at the final visit and a
reviewer wants to know where the other 219 went.

None of these questions is answerable from the data alone. They are
answerable from the data *plus the design*. This session asks: **what
must you know about a trial before you can judge whether its data is
complete?**

This session builds directly on [Why Standards](why-standards.qmd),
which showed that raw data cannot be read without trial-specific
knowledge. Here you acquire that knowledge for the one trial this
course uses throughout.
:::

## Learning Objectives

After completing this session you will be able to:

- Describe the design of GLPX-001 — population, arms, duration,
  endpoints, visit schedule — from its summary
- Predict the expected shape of every raw file from the design alone,
  before opening it
- Account for every gap between prediction and reality, and say which
  gaps are the trial's history and which are data problems
- Name the two organisations that produced the raw files and the
  conventions each brought with it
- Read the emerging treatment effect out of raw lab data — and explain
  why that number is not yet an analysis

## Background: The Protocol Is a Prediction Machine

*Estimated time: ~20 minutes (reading)*

::: {.callout-important}
### The one thing to remember

**The design tells you what the data must look like before you open a
single file. Every gap between that prediction and the files is a
piece of the trial's real history — and your job is to account for
all of it.**
:::

A clinical trial is one of the most pre-specified activities in
science. Who can enrol, what they receive, when they are seen, what is
measured at each visit — all of it is written down before the first
subject walks in. That has a consequence which this session exploits:
**you can compute the shape of the data from the design.** If 400
subjects each attend 8 visits and give 4 lab tests per visit, the lab
file has 12,800 rows. Before you open it.

Reality then edits that prediction. Subjects leave early. Someone is
enrolled twice. A vendor transmits a record twice. Each edit leaves an
arithmetic trace, and the discipline of this session — predict first,
then look, then account for the difference — is the single most useful
habit you can bring to a new study. It converts "here are five
mysterious files" into "here are two discrepancies I need to explain."

### The trial

GLPX-001 is a simulated Phase III, randomised, double-blind,
placebo-controlled trial of **GLPX**, a fictional GLP-1-class agent,
in adults with type 2 diabetes.

| Design element | GLPX-001 |
|---|---|
| Phase | III |
| Population | Adults with type 2 diabetes |
| Arms | GLPX 10 mg once weekly vs placebo, randomised 1:1 |
| Blinding | Double-blind |
| Treatment duration | 26 weeks |
| Primary endpoint | Change in HbA1c from baseline to week 26 (central lab) |
| Key secondary endpoint | Percent change in body weight (vital signs) |
| Planned enrolment | 400 subjects across 12 sites |

Every value in the data is synthetic, generated by
[`R/simulate_trial.R`](https://github.com/sufyansuleman/cdisc-with-r/blob/main/R/simulate_trial.R)
with a fixed seed. No real patient information exists anywhere in this
course. The *structure*, however, is deliberately realistic — this is
what a study team receives.

### The visit schedule

Subjects are screened two weeks before dosing, then seen at baseline
and every four weeks (with a final gap of six) until week 26:

| Visit | Nominal study day |
|---|---|
| SCREENING | −14 |
| BASELINE | 1 |
| WEEK 4 | 29 |
| WEEK 8 | 57 |
| WEEK 12 | 85 |
| WEEK 16 | 113 |
| WEEK 20 | 141 |
| WEEK 26 | 182 |

"Nominal" is doing real work in that column heading: day 29 is when
the week-4 visit is *supposed* to happen. Actual visit dates scatter
around the nominal day, because subjects have lives. You will see this
in Example 2.

### Two sources, two sets of habits

The five raw files come from two different organisations, and
everything about their formatting reflects that:

| File | Source | One row is… |
|---|---|---|
| `dm_raw.csv` | EDC (site-entered) | one subject's demographics and randomisation |
| `ae_raw.csv` | EDC (site-entered) | one adverse event |
| `vs_raw.csv` | EDC (site-entered) | one vital-sign measurement at one visit |
| `ex_raw.csv` | EDC (site-entered) | one subject's dosing record |
| `lb_raw.csv` | Central laboratory vendor | one lab result at one visit |

The **EDC** (electronic data capture) system is where site staff type
what happens: demographics, adverse events, vital signs, dosing. The
**central lab** is a separate company that receives blood samples and
returns results in its own file, with its own column conventions — you
saw in [Why Standards](why-standards.qmd) that it lower-cases its
column names and carries its own copy of each subject's sex. Four lab
tests are measured at every visit: HbA1c, glucose, ALT and
creatinine.

This split matters beyond formatting. The trial's *primary endpoint*
lives in the vendor's file, not in the EDC. When you standardise this
trial's data in Part 2, the lab file's conventions — not the EDC's —
will be the ones you spend the most time reconciling.

::: {.callout-note collapse="true"}
### Why measure the primary endpoint centrally?

Twelve sites could each measure HbA1c locally. The trial instead
ships every sample to one laboratory. The reason is comparability: a
single assay, calibrated once, produces values that can be compared
across sites and visits without asking whether site 104's analyser
runs high. The cost is operational — samples travel, results arrive on
the vendor's schedule in the vendor's format, and the study team must
reconcile a file it does not control. That trade — scientific
consistency for logistical dependence — is typical of how trials
allocate measurement, and it is why "which source produced this
number?" is always worth asking.
:::

## Example 1: Predict the Data Before Opening It

*Estimated time: ~20 minutes (worked example)*

The design table above is enough to predict four of the five file
shapes. Do the arithmetic first, honestly, before looking:

- `dm_raw` — one row per subject: **400 rows**
- `ex_raw` — one dosing record per subject: **400 rows**
- `lb_raw` — 400 subjects × 8 visits × 4 tests: **12,800 rows**
- `vs_raw` — 400 subjects × 8 visits × 4 vitals, plus height measured
  once at screening: **13,200 rows**

(`ae_raw` is deliberately absent — hold that thought for the end of
this example.)

Now look:

```{r predict-shapes}
#| eval: true

raw_paths <- list.files("data/raw", pattern = "\\.csv$", full.names = TRUE)
raw <- raw_paths |>
  set_names(\(p) basename(p)) |>
  map(\(p) read_csv(p, show_col_types = FALSE,
                    col_types = cols(.default = col_character())))

n_subj <- 400

tibble(
  file      = c("dm_raw", "ex_raw", "lb_raw", "vs_raw"),
  predicted = c(n_subj, n_subj, n_subj * 8 * 4, n_subj * 8 * 4 + n_subj),
  actual    = c(nrow(raw[["dm_raw.csv"]]), nrow(raw[["ex_raw.csv"]]),
                nrow(raw[["lb_raw.csv"]]), nrow(raw[["vs_raw.csv"]]))
) |>
  mutate(gap = actual - predicted)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. Two files match the prediction and two miss by
hundreds: `dm_raw` is **+1**, `ex_raw` is exactly right, `lb_raw` is
**−411**, and `vs_raw` is **−412**.

Before reading on, try to explain the pattern yourself. Why would the
two *visit-driven* files be short by almost exactly the same amount?
And why is `dm_raw` — the simplest file in the study — the one with
too *many* rows?
:::

::: {.callout-note}
### Reading the Output Line by Line

| Row | What the gap means |
|---|---|
| `dm_raw +1` | 401 rows for 400 subjects. You met this in [Why Standards](why-standards.qmd): one subject enrolled at two sites. A *data* problem, resolved in [The DM Domain](sdtm-dm.qmd). |
| `ex_raw 0` | One dosing record per randomised subject, and every subject was dosed. The tidiest file in the trial. |
| `lb_raw −411` | Hundreds of lab results that "should" exist don't. This is not a data problem — it is the trial's history, and the next chunk accounts for it. |
| `vs_raw −412` | The same history, seen by the other measuring instrument — and one row *more* missing than the lab file. That difference of exactly 1 matters. |
:::

### Accounting for every missing row

The prediction assumed every subject attends every visit. Check that
assumption:

```{r visits-attended}
#| eval: true

lb <- raw[["lb_raw.csv"]]

visits_per_subject <- lb |>
  distinct(subjid, visit) |>
  count(subjid, name = "visits")

visits_per_subject |> count(visits, name = "subjects")
```

Not everyone reached week 26: **367 subjects completed all 8 visits,
and 33 left early** — after as few as 3 visits. Trials call this early
discontinuation, and at roughly 8% of subjects over 26 weeks it is
unremarkable. Now make the corrected prediction:

```{r account-rows}
#| eval: true

total_visits <- lb |> distinct(subjid, visit) |> nrow()
total_visits

c(lb_expected = total_visits * 4,
  lb_actual   = nrow(lb),
  vs_expected = total_visits * 4 + 400,
  vs_actual   = nrow(raw[["vs_raw.csv"]]))
```

::: {.callout-note}
### Reading the Output Line by Line

The 400 subjects actually attended **3,097** subject-visits between
them. Four lab tests per attended visit predicts 3,097 × 4 = 12,388
rows; the file has **12,389** — one row too many. Four vitals per
attended visit plus 400 height measurements predicts 12,788; the file
has exactly **12,788**.

So after accounting for dropout, `vs_raw` reconciles to the row and
`lb_raw` carries **exactly one surplus record**. You do not yet know
which record it is or why it is there — that is a job for [Events and
Findings](sdtm-events-findings.qmd) — but you know it exists, you
found it with arithmetic rather than by staring at 12,000 rows, and
you know the vendor file is the one carrying it.

This is the method in miniature: dropout explained 411 missing rows
and 412 missing rows in one stroke, and what refused to reconcile —
one single row — is precisely the thing worth investigating.
:::

And `ae_raw`? Its row count was never predictable. Adverse events are
not on the visit schedule; they happen when they happen, to some
subjects and not others. The file has 389 rows covering 262 subjects —
138 subjects have no rows at all, and nothing about the design could
have told you those numbers in advance. Scheduled data can be
reconciled against the protocol; event data cannot. That distinction —
*findings on a schedule* versus *events as they occur* — becomes a
load-bearing idea in [SDTM Concepts](sdtm-concepts.qmd).

## Example 2: Reading the Trial's Story in the Raw Data

*Estimated time: ~20 minutes (worked example)*

Shape is only the first layer. The values tell the trial's story:
who was randomised where and when, who stayed, and what the drug did.

### Randomisation and enrolment

```{r arms-enrolment}
#| eval: true

dm <- raw[["dm_raw.csv"]]

dm |> distinct(SUBJID, ARM) |> count(ARM)

range(dm$RANDDT)
```

Exactly 200 per arm — but note what the code had to do to show it.
`count(ARM)` on the raw file gives 200 and **201**, because of the
duplicated subject; only after `distinct(SUBJID, ARM)` does the true
1:1 randomisation appear. The first subject was randomised in
mid-January 2024 and the last in mid-July: enrolment took about six
months, which means the trial's calendar spans well over a year and no
two subjects are at the same visit on the same date. Keep that in mind
whenever you are tempted to think of "week 26" as a point in time. It
is a point in *each subject's* time.

### The primary endpoint, emerging

The primary endpoint is the change in HbA1c from baseline to week 26.
The raw lab file already contains its raw material. Look at the group
means by visit — remembering from [Why Standards](why-standards.qmd)
that the lab file names things its own way, and noting that because we
read everything as text, `result` must be converted before it can be
averaged:

```{r hba1c-by-visit}
#| eval: true

visit_order <- c("SCREENING", "BASELINE", "WEEK 4", "WEEK 8",
                 "WEEK 12", "WEEK 16", "WEEK 20", "WEEK 26")

hba1c <- lb |>
  filter(test == "HBA1C") |>
  mutate(result = as.numeric(result)) |>
  inner_join(dm |> distinct(SUBJID, ARM),
             by = join_by(subjid == SUBJID)) |>
  mutate(visit = factor(visit, levels = visit_order))

hba1c_means <- hba1c |>
  summarise(mean_hba1c = mean(result), n = n(), .by = c(ARM, visit)) |>
  arrange(visit, ARM)

hba1c_means
```

```{r hba1c-plot}
#| eval: true
#| fig-width: 7
#| fig-height: 4.5

week_lookup <- tibble(
  visit = factor(visit_order, levels = visit_order),
  week  = c(-2, 0, 4, 8, 12, 16, 20, 26)
)

hba1c_means |>
  left_join(week_lookup, by = "visit") |>
  ggplot(aes(x = week, y = mean_hba1c, colour = ARM, shape = ARM)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = arm_colours) +
  scale_x_continuous(breaks = c(-2, 0, 4, 8, 12, 16, 20, 26)) +
  labs(
    title  = "Mean HbA1c by visit and arm (raw central-lab data)",
    x      = "Nominal week",
    y      = "Mean HbA1c (%)",
    colour = NULL, shape = NULL
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

::: {.callout-tip}
### Run It Yourself

Run both chunks. In the table, both arms should start indistinguishable
(screening means 8.37 vs 8.45) and then separate: by week 26 the GLPX
arm mean is **6.90%** against placebo's **8.14%**. The plot shows the
same story as two diverging lines.

Now look at the `n` column and answer honestly: what is quietly
happening to the denominators as the weeks pass?
:::

::: {.callout-note}
### Reading the Output Line by Line

| What you see | What it means |
|---|---|
| Screening means 8.37 / 8.45 | The arms start balanced — this is what successful randomisation looks like, the same baseline comparability you would check in any trial. |
| GLPX falls to 6.90, placebo drifts to 8.14 | A treatment effect of roughly 1.2 percentage points is emerging, alongside a small placebo response. For a 26-week glucose-lowering trial, an entirely plausible picture. |
| `n` falls from 200/201 to 181/186 | The denominators shrink as subjects discontinue. Each week's mean describes *the subjects still present that week* — a subtly different population at every visit. |
| Placebo `n` is 201 at screening | The duplicated subject again, still inflating any count that forgets about them. |
:::

The `n` column is the important one, and it is why this plot is a
*description*, not an analysis. If subjects who leave early differ
from those who stay — sicker, less tolerant of side effects — then
comparing week-26 means among survivors quietly changes the question
being asked. Deciding what "change from baseline to week 26" means
when some week-26 values are missing is a genuine analytical decision,
made in advance and documented — and building the datasets that carry
such decisions is exactly what Part 3 of this course is about. For
now, the honest statement is: *the raw data shows the arms
separating.* How much, in whom, under what handling of the missing —
that is what the rest of the pipeline exists to answer.

### Nominal versus actual dates

One last look, this time at a single subject. The schedule says the
week-4 blood draw happens on study day 29. Did it?

```{r visit-dates}
#| eval: true

sched <- tibble(visit = visit_order,
                nominal_day = c(-14, 1, 29, 57, 85, 113, 141, 182))

lb |>
  filter(subjid == "101-026", test == "HBA1C") |>
  inner_join(sched, by = "visit") |>
  inner_join(dm |> select(SUBJID, RANDDT),
             by = join_by(subjid == SUBJID)) |>
  mutate(
    actual_day = as.integer(as.Date(colldt) - as.Date(RANDDT)) + 1
  ) |>
  select(visit, nominal_day, colldt, actual_day)
```

::: {.callout-note}
### Reading the Output Line by Line

Subject `101-026` attended every visit, but almost never on the
nominal day: the draws land a day or two either side of schedule.
Nothing is wrong here — visit windows exist precisely because humans
cannot be scheduled to the day. But it means "week 4" is a *label*,
not a date, and any analysis that needs elapsed time must compute it
from actual dates, per subject. The `+ 1` in the study-day arithmetic
is itself a convention — here, counting the randomisation day itself
as day 1 — and the exact rule (including how days before the reference
date are numbered) is one of the things a standard pins down so that
every programmer computes it identically. You will meet study days
formally, and precisely, in [SDTM Concepts](sdtm-concepts.qmd).
:::

## What Can Go Wrong

*Estimated time: ~10 minutes (reading)*

**Treating the data as the trial**

The files record what was *captured*, not everything that happened.
`ae_raw` holds the adverse events that sites entered; a symptom a
subject never mentioned is nowhere. The 33 early discontinuations
appear only as absent rows — the data does not say *why* any of them
left. Statements like "there were 389 adverse events" should always be
heard as "389 were recorded."

**Assuming the schedule describes the data**

Every prediction in Example 1 that used "8 visits" was wrong until
corrected for dropout. Any code that hard-codes the visit count — a
loop over 8 visits, a join against a complete subject-visit grid —
will silently invent rows for visits that never happened. Build the
attended-visit set from the data, then reconcile it against the
schedule; never the other way round.

**Averaging over shrinking denominators without noticing**

The week-26 means in Example 2 describe 367 of 400 subjects. Nothing
in `mean(result)` warns you. Whenever a summary spans visits, put `n`
beside it — as a habit, not a flourish. A mean whose denominator you
have not checked is a number you do not yet understand.

**Trusting the tidy file more**

`ex_raw` reconciled perfectly and `lb_raw` is machine-generated, yet
the lab file carries the surplus record and (as you saw in [Why
Standards](why-standards.qmd)) a sex value that contradicts the EDC.
Tidiness measures formatting discipline, not truth.

::: {.callout-warning}
### Common misinterpretations

**"400 subjects means 400 rows"**
Only in subject-level files, and even there only after the
one-row-per-subject rule is actually enforced — `dm_raw` has 401.
Every file's row count follows from *its* granularity: per subject,
per event, or per subject-visit-measurement.

**"The gap between predicted and actual rows means the transfer is
broken"**
Most of the gap (411 of 412 rows here) was the trial's legitimate
history — dropout. The skill is not "expect the prediction to hold";
it is "account for every unit of the difference, and investigate the
remainder."

**"The treatment effect is visible in the raw data, so the analysis
is basically done"**
The raw means answer "what did surviving subjects average, per
visit?" The trial's question — the effect of treatment on change from
baseline, in everyone randomised, with missing data handled as
pre-specified — is a different and harder question. The distance
between those two questions is Parts 2 through 4 of this course.

**"Screening values and baseline values are the same thing"**
GLPX-001 measures HbA1c at both, two weeks apart, and the means
differ (8.37 at screening, 8.28 at baseline for GLPX). Which one is
"baseline" for analysis is a definition, not an observation — and
defining it is a decision you will make explicitly in
[ADSL](adsl.qmd).
:::

## Exercises

This session has no exercise set — its method *is* the exercise. If
you want to practise it now: re-derive `vs_raw`'s row count from
`total_visits` without looking back at Example 1, and confirm your
arithmetic against `nrow()`. Then try the same style of reconciliation
on `ae_raw` and articulate, in one sentence, why it cannot be done.

The first full exercise set arrives with [The DM Domain](sdtm-dm.qmd).

## Comprehension Check

*Estimated time: ~10 minutes*

1. Why did `ex_raw` match its predicted row count exactly while
   `lb_raw` fell 411 rows short? Both cover the same 400 subjects.
2. `count(ARM)` on `dm_raw` gives 200 and 201. Is the trial's
   randomisation unbalanced? What is the correct check?
3. After correcting for attended visits, `vs_raw` reconciled exactly
   and `lb_raw` had one extra row. Why was the surplus *findable*
   without inspecting a single record?
4. Where does the primary endpoint's raw data live, which organisation
   produced it, and why does that arrangement exist?
5. A colleague reports "mean HbA1c at week 26 was 6.90% on GLPX" and
   moves on. What should you ask before letting that number into a
   report?

::: {.callout-note collapse="true"}
### Answers

1. Their granularities respond differently to reality. `ex_raw` has
   one row per *subject*, and every randomised subject has a dosing
   record, so dropout does not remove rows — it only shortens the
   dosing period recorded in `EXENDT`. `lb_raw` has one row per
   subject per *attended visit* per test, so every visit a subject
   missed removes four rows. Same subjects, different units of
   observation.

2. No — it is exactly 200 per arm. The raw count is inflated by the
   subject with two DM rows, who happens to sit in the placebo arm.
   The correct check counts *subjects*, not rows:
   `distinct(SUBJID, ARM) |> count(ARM)`. The general habit: know a
   file's granularity before trusting any count over it.

3. Because both findings files are driven by the same attended visits.
   Once `total_visits` was known, each file had an exact expected
   count (× 4 tests; × 4 vitals + 400 heights). `vs_raw` landing
   exactly on its prediction confirmed the accounting was right, which
   made `lb_raw`'s +1 a real anomaly rather than a rounding error in
   the method. Arithmetic isolated one suspect row out of 12,389
   without reading any of them.

4. HbA1c is measured by the central laboratory and arrives in
   `lb_raw.csv`, the vendor's file — not the EDC. One lab, one assay,
   one calibration makes values comparable across 12 sites and 8
   visits. The price is a file whose conventions the study team does
   not control, which is where much of Part 2's reconciliation work
   comes from.

5. At minimum: *what was the denominator?* (181 of 200 randomised —
   the survivors), *compared to what?* (placebo also fell, to 8.14,
   so the drug's effect is the difference, not the drop), and *change
   from which baseline, handled how for the missing?* None of these
   has a wrong answer in the raw data; they have *undefined* answers
   until the analysis datasets define them. That is why the pipeline
   continues past the raw files.
:::
