---
title: "Tables, Listings and Figures"
---

<!--
SCOPE NOTE: TLF format has no CDISC Implementation Guide the way
SDTM/ADaM do. Every dataset-level claim in this session cites back to
ADaMIG v1.3 / OCCDS v1.1 sections already verified in adsl/adae/adlb.
The one exception is the AE-frequency-display convention, which OCCDS
v1.1 itself attributes to ICH Guideline E3 Sections 12.2.2 and 14.3.1
(OCCDS v1.1, §4, opening line of Example 1), cited as OCCDS reporting
what E3 says, NOT a direct E3 citation, since ICH E3 has not been read
directly. If the ICH E3 PDF arrives, upgrade this to a direct citation.

Deliberately no gt/flextable/rtables in renv.lock, tables are tidy
tibbles rendered with knitr::kable(); no inferential statistics.
-->

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(knitr)
source("R/palette.R")
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 100 minutes.**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background: the shape nobody standardised | 15 min | Reading |
| Example 1: The demographics and disposition table | 20 min | Worked example |
| Example 2: A table and a listing, from the same events | 25 min | Worked example |
| Example 3: The figure the baseline machinery was for | 20 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Exercises + Comprehension Check | 5 min | Practice / Self-test |

The first session of Part 4, and the one that closes the course's data
arc: raw, SDTM, ADaM, and now the document a reviewer actually reads.
For self-paced study, split after Example 1, and again after Example 2.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown. Run R
from the project root, as in earlier sessions.

This session reads `data/adam/adsl.csv`, `data/adam/adae.csv` and
`data/adam/adlb.csv`: nothing new is built here. Every number in this
session is arithmetic on datasets you have already built and cited.
:::

## When Do You Use This? {#sec-tlf}

::: {.callout-tip}
Nobody in a regulatory submission reads ADSL. They read a table that
says how many subjects were randomised, how many completed, and how the
two arms compare. Nobody reads all 12,388 rows of ADLB either; they read
a figure that shows HbA1c falling faster on drug than on placebo. A
medical reviewer scanning for a single serious event does not filter a
dataset. They read a listing built to show exactly that.

Every one of those documents is built from datasets you have already
finished and already cited: ADSL in [ADSL](adsl.qmd), ADAE in
[ADAE](adae.qmd), ADLB in [ADLB](adlb.qmd). None of them need a new
derivation. They need a **question**, asked precisely enough that
someone else can check the answer against the source. That raises this
session's question: **once the data is standardised, who decides what
the report looks like, and how do you prove the numbers in it are
real?**

Builds on [ADSL](adsl.qmd), [ADAE](adae.qmd) and [ADLB](adlb.qmd),
which built the datasets this session only reads.
:::

## Learning Objectives

After completing this session you will be able to:

- Build a demographics and disposition table from ADSL with the correct
  denominator declared up front
- Build a treatment-emergent adverse event summary table from ADAE
  without double-counting a subject who had several events
- Build a listing, and explain in one sentence how a listing differs
  from a table doing the same underlying count
- Build a figure from ADLB's `CHG` and connect what it shows back to
  the `ABLFL` definition that produced it
- Trace any number in a table, listing or figure back to the row of the
  ADaM dataset it came from

## Background: The Shape Nobody Standardised

**Reading task, about 15 minutes.** Everything up to now had a standard
to follow. Outputs do not, and that is the subject of this session.

*Estimated time: ~15 minutes (reading)*

::: {.callout-important}
### The one thing to remember

**A table, listing or figure is not a new dataset. It is a single,
precise question asked of the ADaM datasets already built, answered in
a shape a reviewer can check line by line back to source. CDISC
standardised the data three sessions ago; nobody standardised the shape
of the answer, and that is deliberate.**
:::

Every session since [SDTM Concepts](sdtm-concepts.qmd) has cited a
specific Implementation Guide with numbered sections: SDTMIG v3.4,
ADaMIG v1.3, OCCDS v1.1. This session cannot do that, because **no such
document exists for the shape of a table, a listing or a figure.**
There is no TLFIG.

That is not a gap in this course. It is a fact about the standards. What
a demographics table looks like (which rows, in which order, with what
column headers) comes from the protocol, the statistical analysis
plan, the sponsor's house style, and the structure a clinical study
report is expected to follow. CDISC's own OCCDS guide makes the
boundary explicit when it reaches for a display convention it does not
own:

> The basic summary of adverse event frequencies described in ICH
> Guideline E3 Sections 12.2.2 and 14.3.1 should be used to display
> frequencies in treatment and control groups (OCCDS v1.1, §4).

Read that sentence carefully. OCCDS defines `TRTEMFL`, `AOCCFL` and
every variable this session uses, and then, for the one place it
touches *display*, it points outside CDISC entirely, to an ICH
guideline about clinical study reports. That is the honest answer to
"where is the standard for what a table looks like": there mostly isn't
one, and the document that comes closest is not a CDISC document.

This is the same shape of lesson you have met three times already, at
progressively larger scale. `AGEGR1` had no codelist because an age
grouping is a sponsor decision. `PARAMCD` had no codelist because an
analysis parameter is a sponsor decision. Now: **an entire report** has
no CDISC-mandated layout, because reporting is a sponsor decision too.
CDISC's job ends at making the data trustworthy and traceable. What you
build from it is yours to design, and yours to be able to defend.

### What that means for the tools in this session

Production TLFs are typically built with a package that handles
typesetting: `gt`, `flextable`, or the pharma-specific `rtables`. None
of those are used here, deliberately. This session builds every table as
a plain tibble and prints it with `knitr::kable()`.

That is not a shortcut around the real thing. It is a decision about
what this session is teaching. `gt`/`flextable`/`rtables` are a
**formatting layer over the same derived numbers**. They control
borders, fonts, page breaks and titles, none of which changes what a
number means or where it came from. The lesson here is what question a
table answers and how its numbers trace back to source. A prettier table
answering the wrong question, or one nobody can check, is worse than a
plain one that is right.

### One rule that governs every example below

You met it in [ADAE](adae.qmd) as the denominator trap, and it applies
identically here: **a denominator comes from ADSL, never from the
dataset you are summarising.** A subject with no adverse event has no
row in ADAE; a subject with no lab result at a visit has no row in
ADLB. Count from either of those and you have quietly narrowed the
population to "people something happened to." Every table below states
its denominator from ADSL before it states anything else.

## Example 1: The Demographics and Disposition Table

**Worked example, about 20 minutes.** One stop, four chunks. Every
number in the finished table has to be traceable to a row, and that is
harder than it sounds.

*Estimated time: ~20 minutes (worked example)*

Read the one dataset every table in this session ultimately answers to:

```{r read-adsl}
#| eval: true

adsl <- read_csv("data/adam/adsl.csv", show_col_types = FALSE)

n_arm <- adsl |> count(TRT01P, name = "N")
n_arm
```

200 and 200. Every percentage in this table divides by one of these two
numbers, and every percentage in this table states which one.

```{r age-row}
#| eval: true

age_tbl <- adsl |>
  group_by(TRT01P) |>
  summarise(
    `Mean (SD)` = sprintf("%.1f (%.1f)", mean(AGE, na.rm = TRUE), sd(AGE, na.rm = TRUE)),
    Missing     = as.character(sum(is.na(AGE))),
    .groups = "drop"
  )

kable(age_tbl, caption = "Age (years)")
```

```{r sex-row}
#| eval: true

sex_tbl <- adsl |>
  count(TRT01P, SEX) |>
  left_join(n_arm, by = "TRT01P") |>
  mutate(stat = sprintf("%d (%.1f%%)", n, 100 * n / N)) |>
  select(TRT01P, SEX, stat) |>
  pivot_wider(names_from = TRT01P, values_from = stat)

kable(sex_tbl, caption = "Sex, n (%)")
```

```{r race-row}
#| eval: true

race_tbl <- adsl |>
  count(TRT01P, RACE) |>
  left_join(n_arm, by = "TRT01P") |>
  mutate(stat = sprintf("%d (%.1f%%)", n, 100 * n / N)) |>
  select(TRT01P, RACE, stat) |>
  pivot_wider(names_from = TRT01P, values_from = stat)

kable(race_tbl, caption = "Race, n (%)")
```

```{r disposition-row}
#| eval: true

disp_tbl <- adsl |>
  count(TRT01P, EOSSTT) |>
  left_join(n_arm, by = "TRT01P") |>
  mutate(stat = sprintf("%d (%.1f%%)", n, 100 * n / N)) |>
  select(TRT01P, EOSSTT, stat) |>
  pivot_wider(names_from = TRT01P, values_from = stat)

kable(disp_tbl, caption = "Disposition, n (%)")
```

::: {.callout-tip}
### Run It Yourself

**Your turn, about 6 minutes.** Pick one number in the finished table
and trace it back to the rows it came from. If you cannot, neither can
a reviewer.

Run all four chunks. Every row of every one of these tables is a
`count()` or a `summarise()` on `ADSL` alone: no new logic, only
presentation of a value derived in [ADSL](adsl.qmd) and cited there.

Two things worth checking on the numbers themselves. First, `AGE`'s two
missing subjects per arm are exactly the four subjects from D1, whose
incomplete birth dates you met in [The DM Domain](sdtm-dm.qmd). They
are still missing here, four sessions later, because nothing between
then and now had a reason to impute them. Second, `EOSSTT` here reads
181/19 for GLPX and 186/14 for Placebo, which is the same 367/33 split
you first counted in [The GLPX-1 Trial](the-trial-glpx001.qmd), now
shown by arm rather than pooled. Same subjects, same numbers, different
question asked of them.
:::

### Traceability, made concrete

A table entry is only as good as your ability to point at the row it
came from. Pick a cell (GLPX1-101-010, `Placebo`, `WHITE`,
`COMPLETED`) and check it:

```{r trace-one}
#| eval: true

adsl |>
  filter(USUBJID == "GLPX1-101-010") |>
  select(USUBJID, TRT01P, AGE, SEX, RACE, EOSSTT)
```

That one row is why the `WHITE` count under `Placebo` includes this
subject, and why the `COMPLETED` count does too. A number that cannot be
walked back to a row like this one is not a result. It is a claim.

## Example 2: A Table and a Listing, from the Same Events

**Worked example, about 25 minutes.** Two stops. The same events
summarised two ways, and the denominator is the whole argument.

*Estimated time: ~25 minutes (worked example)*

A **table** counts. A **listing** enumerates. They can describe exactly
the same underlying records and still look nothing alike, because they
answer different questions: "how many?" against "which ones, in full?"
GLPX-1 gives you a case where you can watch both happen to the same
event.

### The table: how many subjects, not how many events

```{r ae-safety-n}
#| eval: true

adae <- read_csv("data/adam/adae.csv", show_col_types = FALSE)

n_saf <- adsl |> filter(SAFFL == "Y") |> count(TRT01P, name = "N")
n_saf
```

Declare the denominator first, as above: 200 and 200, the safety
population from ADSL, not a count of who happens to have a row in ADAE.

```{r ae-teae-table}
#| eval: true

teae_tbl <- adae |>
  filter(TRTEMFL == "Y") |>
  distinct(USUBJID, TRT01A) |>
  count(TRT01A, name = "n") |>
  rename(TRT01P = TRT01A) |>
  left_join(n_saf, by = "TRT01P") |>
  mutate(stat = sprintf("%d (%.1f%%)", n, 100 * n / N)) |>
  select(TRT01P, stat)

kable(teae_tbl, col.names = c("Treatment", "Subjects with ≥1 TEAE, n (%)"))
```

::: {.callout-important}
### `distinct(USUBJID, TRT01A)` is doing the entire job of this table

Look back at the code. `TRTEMFL == "Y"` filters to treatment-emergent
event *rows*. `distinct(USUBJID, TRT01A)` collapses those rows to one
per *subject* before counting. Drop that one line and a subject with
four events is counted four times, and the percentages could exceed
100%.

This is `AOCCFL`'s job by another route. [ADAE](adae.qmd) derived
`AOCCFL` for exactly this purpose, flagging one row per subject as the
"first occurrence" so a table can filter to it directly instead of
de-duplicating by hand. Either approach gives the same answer; using
`AOCCFL` is closer to how a production table would be built, because the
flag, not an inline `distinct()`, is what a reviewer expects to find
documented in the analysis metadata.
:::

Now severity, where `AOCCIFL`: the *worst*-severity occurrence flag,
does the same job for a different question:

```{r ae-severity-table}
#| eval: true

sev_tbl <- adae |>
  filter(TRTEMFL == "Y", AOCCIFL == "Y") |>
  count(TRT01A, ASEV) |>
  rename(TRT01P = TRT01A) |>
  left_join(n_saf, by = "TRT01P") |>
  mutate(stat = sprintf("%d (%.1f%%)", n, 100 * n / N)) |>
  select(TRT01P, ASEV, stat) |>
  pivot_wider(names_from = TRT01P, values_from = stat)

kable(sev_tbl, caption = "Subjects by Worst Treatment-Emergent Severity, n (%)")
```

::: {.callout-tip}
### Run It Yourself

**Your turn, about 5 minutes.** Work out what the denominator is before
you read on. Getting it wrong changes these percentages substantially,
and the wrong version looks perfectly reasonable.

75.5% of GLPX subjects and 55.0% of Placebo subjects had at least one
TEAE: the same two numbers [ADAE](adae.qmd)'s denominator-lesson chunk
produced, arrived at the same way, now sitting in the shape a CSR table
would actually publish them in.

Filtering to `AOCCIFL == "Y"` for severity, rather than `TRTEMFL == "Y"`
alone, is what stops a subject with one Mild and one Severe event from
appearing in both rows of the severity table. Every subject contributes
to exactly one severity cell: the same "exactly one flag per subject"
discipline `ABLFL` enforced for baseline in [ADLB](adlb.qmd).

Worth pausing on, because it is the wrong instinct made concrete: 151
GLPX subjects had *any* TEAE, but only 13 + 67 + 71 = 151 appear across
the three severity rows: the arithmetic checks, because `AOCCIFL` gives
exactly one severity per subject who had any event. Had the table instead
filtered only on `TRTEMFL == "Y"` and counted every severity a subject
ever reached, those three numbers would sum to more than 151, because a
subject can be Mild once and Moderate later.
:::

::: {.callout-note collapse="true"}
### For contrast: the wrong denominator, one more time

You have seen this exact failure mode before, but it is worth one more
look now that it sits beside the correct table:

```{r ae-wrong-denom}
#| eval: true

adae |>
  filter(TRTEMFL == "Y") |>
  distinct(USUBJID, TRT01A) |>
  count(TRT01A, name = "n") |>
  mutate(pct_of_subjects_with_ae = round(100 * n / sum(n), 1))
```

57.9% and 42.1%. Identical `n` values to the correct table above,
different percentages, and these two sum to 100%, which is the signal
that this column answers "what share of *affected* subjects were on
each arm," not "what share of *each arm* was affected." Same records,
same code up to the last `mutate()`, a materially different (and
wrong) clinical claim.
:::

### The listing: every instance, none of them collapsed

GLPX-1 has exactly one serious adverse event. That is a small enough
number to show you the entire listing, and it is the best possible size
for seeing what a listing is *for*:

```{r sae-listing}
#| eval: true

sae_listing <- adae |>
  filter(AESER == "Y") |>
  transmute(
    Subject           = USUBJID,
    `Adverse Event`   = AETERM,
    `Start Date`      = ASTDT,
    Severity          = ASEV,
    `Treatment Emerg.` = TRTEMFL,
    Arm               = TRT01A
  )

kable(sae_listing, caption = "Listing of Serious Adverse Events")
```

::: {.callout-tip}
### Run It Yourself

**Your turn, about 4 minutes.** One subject, one row. Ask what a listing
gives a reviewer that the table above it cannot.

One row: `GLPX1-110-004`, Hypoglycaemia, 2024-03-23, Severe,
treatment-emergent, GLPX 10 mg. You met this subject in
[ADAE](adae.qmd) as the study's single serious event.

Now count the same event, the way a table would:

```{r sae-as-count}
#| eval: true

adae |> filter(AESER == "Y") |> count(TRT01A, name = "n_serious_events")
```

One row either way, because there is only one event, which makes this
a poor example of the *volume* difference between a table and a
listing, and a very good one for the *purpose* difference. The count
tells a reviewer "there was 1 serious event on GLPX 10 mg." It does not
say what it was, when it happened, or how severe. The listing exists
because "1" is not an answerable safety question on its own: a medical
reviewer's next question is always "which one, and show me," and a
table cannot answer that. A listing is built to be read row by row, by
a person deciding whether to worry; a table is built to be scanned,
by a person deciding where to look next.
:::

## Example 3: The Figure the Baseline Machinery Was For

**Worked example, about 20 minutes.** One stop. This figure is the
payoff for four sessions of baseline work.

*Estimated time: ~20 minutes (worked example)*

Every argument [ADLB](adlb.qmd) made about `ABLFL`, why "the BASELINE
visit" was the wrong definition, why the SAP's "last value on or before
first dose" was the right one, why it mattered enough to check against
the data rather than assume, existed to produce one chart: the mean
change from baseline in HbA1c over time, by arm. This is that chart.

```{r read-adlb}
#| eval: true

adlb <- read_csv("data/adam/adlb.csv", show_col_types = FALSE)

hba1c <- adlb |>
  filter(PARAMCD == "HBA1C", !is.na(CHG), AVISIT != "SCREENING")

plot_data <- hba1c |>
  group_by(AVISIT, AVISITN, TRTA) |>
  summarise(n = n(), mean_chg = mean(CHG), se = sd(CHG) / sqrt(n()),
            .groups = "drop")

plot_data |> arrange(AVISITN, TRTA) |>
  mutate(across(c(mean_chg, se), \(x) round(x, 3)))
```

`SCREENING` is excluded on purpose: it is a pre-dose visit, and `CHG` is
only non-null there for the subset of subjects whose own baseline
happens to fall on the screening record, plotting it would show a
partial, artificially-flat point that means something different from
every point after it.

::: {.callout-important}
### Why `BASELINE` is not exactly 0.000, and this is the whole course arc closing

Look at the `BASELINE` row above: `-0.051` for GLPX, `-0.003` for
Placebo. Not zero. If you have been following the baseline thread since
[The DM Domain](sdtm-dm.qmd), that number should stop you, and it is
worth stopping on rather than rounding away.

`CHG` is exactly 0 on the row where `ABLFL == "Y"`, that was checked
and confirmed in [ADLB](adlb.qmd). But `ABLFL` does not always land on
the visit *labelled* `BASELINE`. Split this trial's HbA1c subjects by
where their own baseline actually fell:

```{r why-not-zero}
#| eval: true

bl_source <- adlb |>
  filter(PARAMCD == "HBA1C") |>
  group_by(USUBJID) |>
  summarise(baseline_landed_at = AVISIT[which(ABLFL == "Y")], .groups = "drop")

count(bl_source, baseline_landed_at)
```

244 subjects have their true baseline at `SCREENING`, recall from
[ADLB](adlb.qmd) that this happens whenever the scheduled `BASELINE`
visit fell after first dose. For those 244, the row *labelled*
`BASELINE` is not their baseline at all. It is an early post-baseline
measurement, with a real, usually small, `CHG` already on it. Only the
other 156 have `CHG = 0` at the `BASELINE` visit. Average all 400
together and you get a small non-zero number: the visible fingerprint,
in this figure, of the exact mechanism [ADLB](adlb.qmd) spent an entire
session deriving correctly.

**This is not noise to smooth over.** A course (or a study) that
plotted `BASELINE` and saw a clean 0.000 either got lucky with its
schedule or made the mistake ADLB warned against: using the visit label
instead of the dosing date. Seeing a small, explainable non-zero number
here is confirmation the baseline logic is doing its job, not a sign
that something is wrong.
:::

```{r hba1c-figure}
#| eval: true
#| fig-alt: "Line chart of mean change from baseline in HbA1c by visit, GLPX 10 mg versus placebo. GLPX 10 mg declines steadily from near zero at baseline to about -1.4 percentage points at week 26; placebo declines more gradually to about -0.3. Error bars show standard error at each visit."

ggplot(plot_data, aes(x = AVISITN, y = mean_chg, colour = TRTA, shape = TRTA)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  geom_errorbar(aes(ymin = mean_chg - se, ymax = mean_chg + se), width = 0.15) +
  scale_colour_manual(values = arm_colours) +
  scale_x_continuous(
    breaks = sort(unique(plot_data$AVISITN)),
    labels = plot_data |> distinct(AVISITN, AVISIT) |> arrange(AVISITN) |> pull(AVISIT)
  ) +
  labs(x = NULL, y = "Mean Change from Baseline in HbA1c (%)",
       colour = "Treatment", shape = "Treatment") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

::: {.callout-tip}
### Run It Yourself

**Your turn, about 5 minutes.** Before reading our commentary, decide
what this figure does not tell you. It is persuasive, and persuasive is
when to be careful.

Run the chunk. GLPX 10 mg separates from placebo almost immediately and
the gap widens every visit, reaching -1.41 against -0.30 by week 26:
the same treatment effect [ADLB](adlb.qmd) reported as a table of
numbers, now as the chart a clinical study report would actually
publish.

Colour and shape both carry treatment arm, using `R/palette.R`'s
`arm_colours`: the same Okabe-Ito palette convention established for
the very first figure in [The GLPX-1 Trial](the-trial-glpx001.qmd),
carried through every figure in this course without exception. A
colourblind reader loses nothing; identity never rests on colour alone.

Error bars are standard error at each visit, computed from the same `n`
and `CHG` values in the table above them, another number you can trace
back, this time to a formula rather than a single row, but no less
checkable for it.
:::

### Conformance, one last time

```{r final-checks}
#| eval: true

teae_n <- adae |> filter(TRTEMFL == "Y") |> distinct(USUBJID) |> nrow()
sev_n  <- adae |> filter(TRTEMFL == "Y", AOCCIFL == "Y") |> nrow()

tibble(
  check = c("every table's denominator stated and sourced from ADSL",
            "AE table has exactly one row per subject with a TEAE",
            "severity rows sum to the same subject count as the AE table",
            "figure's SCREENING row excluded (partial data, would mislead)",
            "figure uses the course's established arm_colours palette"),
  pass = c(
    TRUE,   # by construction: every *_tbl above joins n_arm / n_saf first
    teae_n == nrow(adae |> filter(TRTEMFL == "Y") |> distinct(USUBJID, TRT01A)),
    sev_n == teae_n,
    !("SCREENING" %in% plot_data$AVISIT),
    all(arm_colours %in% okabe_ito)
  )
)
```

With this session, every dataset built in Parts 2 and 3 has been read
back out into a document a reviewer could actually be handed: a
demographics table, a safety table and its listing, and the efficacy
figure the entire baseline argument existed to produce.

## What Can Go Wrong

**Reading task, about 10 minutes.** No code. Mostly about tables that
are correct and still mislead.

*Estimated time: ~10 minutes (reading)*

**Computing a denominator from the dataset being summarised.** The
single most consequential error in this session, exactly as it was in
[ADAE](adae.qmd). A table's `N` comes from ADSL, stated before anything
else is computed.

**Counting events instead of subjects.** A TEAE table answers "how many
*subjects*," not "how many *events*." Skip the de-duplication (`distinct()` or an occurrence flag) and a subject with several events
inflates the count and can push a percentage past 100%.

**Filtering severity on the wrong flag.** `TRTEMFL == "Y"` alone lets a
subject appear in more than one severity row. `AOCCIFL == "Y"` restricts
to each subject's single worst occurrence, so the severity rows sum to
the same total as the any-TEAE table.

**Treating a listing as a small table.** A listing does not
de-duplicate, does not aggregate, and does not have a denominator in the
way a table does. Running `count()` on a listing's source and reporting
that instead defeats its purpose, which is to show every instance in
full for individual review.

**Plotting a visit label instead of asking what baseline actually is.**
The `BASELINE`-visit mean in this session's figure is not exactly zero,
and that is correct. A figure that assumed it should be, and "corrected"
it, would be hiding the exact mechanism [ADLB](adlb.qmd) argued for at
length: baseline is a dosing-date decision, not a visit-label one.

**Reaching for a CDISC citation that does not exist.** There is no CDISC
standard for what a table, listing or figure must look like. Citing one
anyway (inventing a section number because every other session had
one) is worse than citing nothing.

### Common misinterpretations

**"This isn't a real TLF because it's not in `gt`/`flextable`."** The
numbers, the denominators and the traceability are the real content of
a TLF. Typesetting is a formatting layer applied on top of exactly these
derived values, in a later step this course does not build.

**"CDISC must specify the table layout somewhere; I just haven't found
it."** It does not, and OCCDS says so implicitly by reaching outside
CDISC to ICH E3 for the one display convention it discusses. The
standardisation stops at the data.

**"A listing is just a table without the aggregation."** A listing is
built for a different *reading pattern*, row by row, in full, by
someone deciding whether to act on any single record, not merely a
table with one fewer `group_by()`.

## Exercises

**Exercise, about 15 minutes.** Do these before the next session.

Practice: [Exercise 7: Tables, Listings and Figures](../exercises/tlf.qmd).

## Comprehension Check

**Self-test, about 5 minutes.** Answer before expanding.

1. Every table in this session states its `N` before it states any
   percentage, and that `N` always comes from ADSL. Why not compute it
   from whichever dataset the table is summarising?
2. The severity table uses `AOCCIFL == "Y"`, not `TRTEMFL == "Y"` alone.
   What would go wrong with the severity row totals if you filtered on
   `TRTEMFL` alone?
3. GLPX-1's single serious adverse event appears as `n = 1` in both a
   table count and the listing. In what sense are these two views of
   the same event still doing different jobs?
4. This session's figure excludes `SCREENING` and shows a `BASELINE`
   mean of -0.051 rather than 0. Explain both decisions from what you
   already know about `ABLFL` from [ADLB](adlb.qmd).
5. A colleague asks which CDISC guide specifies how a demographics table
   should be laid out. What is the honest answer, and what document
   does OCCDS itself point to instead?

::: {.callout-note collapse="true"}
### Answers

1. Because the dataset being summarised only contains subjects
   *something happened to*: an adverse event, a lab result. A subject
   with none has no row there at all, so a denominator drawn from it
   silently excludes exactly the subjects who had nothing to report,
   inflating every percentage. ADSL has one row per subject regardless
   of what happened to them, which is what makes it the only correct
   source for a denominator (OCCDS v1.1, §3.2.1; the same lesson from
   [ADAE](adae.qmd)).
2. A subject can reach more than one severity level across their
   treatment-emergent events, Mild once, Moderate later. Filtering on
   `TRTEMFL` alone and counting every severity a subject ever reached
   would let that subject appear in two rows of the table, so the three
   severity counts would sum to more than the number of subjects with
   any TEAE. `AOCCIFL` restricts to exactly one row, the subject's
   single worst occurrence, so the rows sum correctly.
3. The count answers "how many": one serious event occurred, on GLPX 10
   mg. It does not say what the event was, when it started, how severe
   it was, or whether it resolved. The listing answers "which one, in
   full": every field a reviewer would need to assess the case on its
   own. A table is built to be scanned for where to look; a listing is
   built to be read for what happened.
4. `SCREENING` is excluded because `CHG` is only defined there for
   subjects whose own baseline happens to be the screening record,
   plotting it would show a partial, misleadingly flat data point built
   from a different subset of subjects than every later visit. The
   non-zero `BASELINE` mean follows from the same fact from the other
   direction: 244 of 400 subjects have `ABLFL == "Y"` at `SCREENING`,
   not `BASELINE`, because their scheduled `BASELINE` visit landed after
   first dose ([ADLB](adlb.qmd)). For those subjects the row labelled
   `BASELINE` is a small post-baseline measurement, not their baseline,
   so it carries a small non-zero `CHG` that pulls the arm mean away
   from exactly zero.
5. There is no CDISC guide for TLF layout, that is not an oversight but
   a fact about the standards, which stop at making the data trustworthy
   and traceable. OCCDS v1.1 itself, when it needs a display convention
   for adverse event frequencies, points outside CDISC entirely to ICH
   Guideline E3, Sections 12.2.2 and 14.3.1 (OCCDS v1.1, §4).
:::
