---
title: "Why Standards"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 75 minutes.**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background | 15 min | Reading |
| Example 1: Five Files, Five Conventions | 20 min | Worked example |
| Example 2: One Question, Three Answers | 15 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Comprehension Check | 10 min | Self-test |

This session is shorter than most and has no exercise set: it is the
argument for everything that follows, not a technique to practise. It
fits one sitting. Read it before [The GLPX-001
Trial](the-trial-glpx001.qmd), which introduces the data you will work
with for the rest of the course.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown.
Where you see a "Run It Yourself" box, stop and run the code above it
before reading the explanation that follows.

Run R from the **project root** (open the repository's `.Rproj` file,
or `setwd()` to the folder containing `_quarto.yml`). Every path in
this course is written relative to that root, so `data/raw/dm_raw.csv`
means what it says.
:::

## When Do You Use This? {#sec-why-standards}

::: {.callout-tip}
You have joined a study team and been handed five CSV files and a
question: "how many patients had nausea?" Or a reviewer has asked why
the subject count in your table is 400 when the demographics file has
401 rows. Or someone says the word "SDTM" in a meeting and you nod.

In each case the underlying problem is the same: **the data cannot be
read correctly by anyone who does not already know this particular
trial.** That is the problem standards solve, and this session
demonstrates it before naming a single CDISC variable.

This is the first session of the course and assumes no CDISC
knowledge. It assumes you can read `dplyr`.
:::

## Learning Objectives

After completing this session you will be able to:

- Explain why clinical data standards exist, in terms of the specific
  problems they solve rather than the regulations that mandate them
- Demonstrate, in R, that raw trial data carries structure and
  vocabulary that only the study team understands
- Distinguish the three things a standard fixes: variable naming,
  dataset structure, and controlled vocabulary
- Explain why "we will standardise it at the end" is a false economy
- Describe the shape of the SDTM → ADaM → TLF pipeline the rest of
  this course builds

## Background: Data That Can Only Be Read by Its Author

*Estimated time: ~15 minutes (reading)*

Here is the single idea that ties this session together, and arguably
the whole course:

::: {.callout-important}
### The one thing to remember

**Raw trial data can only be read by someone who already knows the
trial. A standard is what lets it be read by someone who doesn't.**
:::

Every dataset carries two kinds of information. There is the data
itself: the numbers, the dates, the terms. And there is the knowledge
required to interpret it: that `colldt` means the date the sample was
collected, that a blank `AEENDT` means the event was still ongoing
rather than that someone forgot, that subject `101-012` appears twice
for a reason.

In a normal research project, that second kind of information lives in
someone's head, and that is survivable. You wrote the data; you can
read the data. If a colleague needs it, you explain it to them over
coffee.

Clinical trials break this arrangement in three ways at once.

**Scale.** A regulator receives study data from many sponsors, for many
trials, in many therapeutic areas, every year. Nobody at the agency can
have coffee with every study team. If each sponsor invents its own
layout, then reviewing $n$ trials requires understanding $n$ different
conventions, and no tool built for one submission works on the next.

**Time.** A trial's data outlives the people who made it. Regulators
may return to a submission years later; a safety signal may surface
long after the study team has moved on. "Ask the programmer who built
it" stops being an available answer.

**Consequence.** If a reviewer misreads your adverse event data, the
result is not an awkward correction in a journal. It is a wrong
conclusion about whether a drug harms people.

So the industry did the only thing that scales: it agreed, in advance,
on what the columns are called, how the datasets are shaped, and which
words are allowed in them. That agreement is what CDISC publishes.

### A standard fixes three separate things

It is worth separating these now, because learners routinely collapse
them into "renaming columns," and then wonder why the rest of the
course is hard.

| What is fixed | The question it answers | Where this course covers it |
|---|---|---|
| **Variable naming** | What is this column called, in every trial? | [SDTM Concepts](sdtm-concepts.qmd) |
| **Dataset structure** | What is one row, and which datasets exist? | [SDTM Concepts](sdtm-concepts.qmd), [The DM Domain](sdtm-dm.qmd) |
| **Controlled vocabulary** | Which values is this column allowed to hold? | [SDTM Concepts](sdtm-concepts.qmd), [Events and Findings](sdtm-events-findings.qmd) |

Example 1 below demonstrates the first two problems in the GLPX-001
raw data. Example 2 demonstrates the third. Neither example needs you
to know any CDISC yet — that is the point. You should feel the problem
before you are handed the solution.

### The shape of what is coming

The rest of this course is one pipeline, and it is worth seeing the
whole shape before you meet any of its parts:

- **Raw data** — what the study team receives: an EDC export, a lab
  vendor's file, each with its own conventions. This is where you are
  now.
- **SDTM** — the raw data, reorganised and renamed to the standard,
  with the original values preserved. It answers: *what happened in
  this trial?* (Part 2)
- **ADaM** — analysis-ready datasets derived from SDTM, carrying the
  variables a statistician needs to run the analysis and a reviewer
  needs to reproduce it. It answers: *what does the trial show?*
  (Part 3)
- **TLFs** — the tables, listings and figures that go in the study
  report. (Part 4)

Each stage has a different job, and a great deal of confusion in
clinical programming comes from asking one stage to do another's work.
[SDTM Concepts](sdtm-concepts.qmd) and [ADaM
Concepts](adam-concepts.qmd) draw that line properly.

::: {.callout-warning}
## VERIFY

Claim: Regulatory authorities require CDISC-conformant study data for
marketing applications — specifically, that FDA and PMDA mandate SDTM
and ADaM (and define.xml as the accompanying data definition document)
for applicable studies, with the requirement flowing from the FDA Data
Standards Catalog and the Study Data Technical Conformance Guide.

Source needed: FDA Data Standards Catalog (current version, and the
applicability/effective dates); FDA Study Data Technical Conformance
Guide; PMDA notification on electronic study data submission.

Author: confirm the requirement, the naming of each document, and the
dates before publish. The argument in this session deliberately does
not depend on these specifics — it is made from the data — but the
regulatory framing should be stated correctly or not at all.
:::

::: {.callout-note collapse="true"}
### Why not just write good documentation instead?

A reasonable objection: if the problem is that knowledge lives in
someone's head, why not write it down? A thorough data dictionary per
trial would solve it, wouldn't it?

It solves the *comprehension* problem and not the *automation* one. A
reviewer who can read your dictionary still cannot run last year's
program against your data, because your column is called `colldt` and
last year's was called `SAMPDT`. Documentation lets a human understand
$n$ conventions one at a time. A standard lets one program read all
$n$. The difference compounds: every validation rule, every review
tool, and every analysis macro is written once against the standard
rather than once per trial.

This is also why the standard is not negotiable per-study. A standard
that each sponsor adapts to taste is a dictionary with extra steps.
:::

## Example 1: Five Files, Five Conventions

*Estimated time: ~20 minutes (worked example)*

GLPX-001 is a simulated Phase III trial of a GLP-1-class agent in type
2 diabetes. [The next session](the-trial-glpx001.qmd) introduces it
properly. For now you only need to know that its raw data is what a
study team actually receives: an export from the EDC system where sites
enter data, plus a separate file from the central laboratory that
analyses the blood samples. Two organisations, two sets of habits.

Load all five raw files and look at nothing but their shape:

```{r read-raw}
#| eval: true

raw_paths <- list.files("data/raw", pattern = "\\.csv$", full.names = TRUE)

raw <- raw_paths |>
  set_names(\(p) basename(p)) |>
  map(\(p) read_csv(p, show_col_types = FALSE,
                    col_types = cols(.default = col_character())))

tibble(
  file    = names(raw),
  rows    = map_int(raw, nrow),
  columns = map_chr(raw, \(d) paste(names(d), collapse = ", "))
)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. You should see five files: `ae_raw.csv` (389
rows), `dm_raw.csv` (401), `ex_raw.csv` (400), `lb_raw.csv` (12,389)
and `vs_raw.csv` (12,788).

Before reading on, look only at the `columns` column and answer one
question: **if you had to write a single function that returned the
subject identifier from any of these five files, what would it look
like?**
:::

Everything here is read as character deliberately. Left to itself,
`read_csv()` guesses types, and a guess is exactly the kind of
trial-specific knowledge we are trying to expose rather than hide. A
date that is sometimes `2024-03-12` and sometimes `1969` is not a date
column to R; it is a mess, and we want to see the mess.

The answer to the question in the box is that your function needs a
lookup table, because the files do not agree:

```{r subject-col}
#| eval: true

tibble(
  file           = names(raw),
  subject_column = map_chr(raw, \(d) {
    paste(grep("subjid", names(d), ignore.case = TRUE, value = TRUE),
          collapse = ", ")
  }),
  date_columns   = map_chr(raw, \(d) {
    paste(grep("dt$", names(d), ignore.case = TRUE, value = TRUE),
          collapse = ", ")
  })
)
```

::: {.callout-note}
### Reading the Output Line by Line

| What you see | What it means |
|---|---|
| `SUBJID` in four files, `subjid` in `lb_raw.csv` | The lab vendor lower-cases its columns. R is case-sensitive, so `dm$SUBJID` and `lb$subjid` are different names for the same concept. Every join you write must know this. |
| `BIRTHDT`, `RANDDT`, `AESTDT`, `AEENDT`, `EXSTDT`, `EXENDT`, `VSDT`, `colldt` | Eight date columns, eight names, no shared pattern. `colldt` does not even end in the same convention as the others. |
| `ae_raw.csv` has 389 rows | Fewer rows than subjects. Adverse event data has one row per *event*, not per subject — and subjects with no events have no rows at all. |
| `lb_raw.csv` has 12,389 rows | One row per subject per visit per test. The same trial, at a completely different granularity. |

Notice what the last two rows of that table mean: **"one row" means
something different in every file.** Structure is not a presentational
detail you can fix with `rename()`.
:::

The `grep()` in that chunk is itself the lesson. To find the subject
column, we had to search for it with a case-insensitive pattern,
because we already knew — from having looked — that it might be
upper or lower case. A standard is what makes that search unnecessary.

### When two sources disagree about a fact

Both the EDC and the lab file record each subject's sex. They were
entered by different people, at different times, into different
systems. Nothing forces them to agree, so let us check whether they do:

```{r sex-conflict}
#| eval: true

edc_sex <- raw[["dm_raw.csv"]] |> distinct(SUBJID, edc_sex = SEX)
lab_sex <- raw[["lb_raw.csv"]] |> distinct(subjid, lab_sex = sex)

edc_sex |>
  inner_join(lab_sex, by = join_by(SUBJID == subjid)) |>
  filter(edc_sex != lab_sex)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. Exactly one subject comes back: `111-003`, whom
the EDC records as `F` and the lab records as `M`.

Now sit with the awkward part: **which one is right?** The data cannot
tell you. Nothing in either file is malformed. Both values are
perfectly valid. They simply disagree, and you have no way to resolve
it from the data alone.
:::

This is not a trick, and it is not rare. It is a defect deliberately
built into the GLPX-001 simulation because it is one of the most
ordinary things that happens in a real trial. You will resolve this
subject properly in [The DM Domain](sdtm-dm.qmd), where you will see
that the standard's contribution is not to *know* the answer, but to
force the question to be asked and the decision to be recorded.

Note the join itself: `join_by(SUBJID == subjid)`. Because the two
files name the same concept differently, the join cannot be written
without trial-specific knowledge. After standardisation it can.

### Counting subjects is not `nrow()`

One more, and it is the one that catches everybody:

```{r count-subjects}
#| eval: true

dm <- raw[["dm_raw.csv"]]

nrow(dm)
n_distinct(dm$SUBJID)
```

The demographics file has 401 rows and 400 subjects. If you reported
`nrow(dm)` as the number of enrolled subjects, you would be wrong, and
nothing in the file would warn you:

```{r count-subjects-dup}
#| eval: true

dm |>
  add_count(SUBJID, name = "rows_for_this_subject") |>
  filter(rows_for_this_subject > 1) |>
  select(SUBJID, SITEID, RANDDT, ARM)
```

::: {.callout-note}
### Reading the Output Line by Line

Subject `101-012` has two rows: one at site `101` randomised on
`2024-03-06`, one at site `108` randomised on `2024-03-27`. Read them
as a story rather than as a bug. A subject was screened at one site,
moved, and was enrolled again at another site under the same subject
number. Both rows are things that genuinely happened and both were
genuinely recorded.

The file is not corrupt. It simply has no rule that says *one row per
subject* — so nobody enforced one, and a plausible sequence of events
produced two. The standard's contribution here is the rule itself,
stated in advance and checkable by a machine. You will apply it in
[The DM Domain](sdtm-dm.qmd).
:::

This is the whole argument in miniature. Three ordinary questions —
what is the subject column, what sex is subject `111-003`, how many
subjects are there — and each one needed knowledge that is not in the
data.

## Example 2: One Question, Three Answers

*Estimated time: ~15 minutes (worked example)*

Naming and structure are the visible problems. Vocabulary is the one
that will actually hurt you, because it produces answers rather than
errors.

Here is a question a regulator genuinely asks, and one of the simplest
questions available: **how many subjects had nausea?**

Start by looking at what sites typed into the adverse event term
field:

```{r ae-terms}
#| eval: true

ae <- raw[["ae_raw.csv"]]

ae |> count(AETERM, sort = TRUE)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above and read the whole list — all 25 rows, not just
the top. Find every row that a clinician would call nausea. There are
four of them.
:::

::: {.callout-note}
### Reading the Output Line by Line

| What you see | What it means |
|---|---|
| `Nausea` (61), `NAUSEA` (2), `nausea` (2) | The same event, typed three ways. Case is not meaningful here — but R has no way to know that. |
| `Diarrhoea` (56), `DIARRHOEA` (2), `diarrhoea` (1), `Diarrea` (1) | Case again, plus a misspelling. `Diarrea` is a typo that no case-folding will rescue. |
| `nausea and vomitting` (1) | Two clinical events in one field, and a misspelling. This is not a formatting problem; it is a *data* problem. One AE record is describing two AEs. |
| 25 rows in total | Twenty-five distinct strings for what sites believed were ordinary, common events. |
:::

How many distinct things are actually in this column?

```{r ae-distinct}
#| eval: true

n_distinct(ae$AETERM)
n_distinct(toupper(ae$AETERM))
```

Twenty-five as recorded; twelve once you ignore case. But GLPX-001 only
ever produced **ten** clinical concepts. Upper-casing gets you from 25
to 12, and then stops:

```{r ae-upper-terms}
#| eval: true

sort(unique(toupper(ae$AETERM)))
```

The two extra entries are `DIARREA` and `NAUSEA AND VOMITTING`. No
string operation will fix them, because fixing them requires knowing
that "Diarrea" means diarrhoea and that "nausea and vomitting" is two
events. That knowledge is medical, not textual.

### Three defensible rules, three different answers

Now answer the regulator's question. Watch what happens:

```{r nausea-count}
#| eval: true

tibble(
  rule = c("AETERM == 'Nausea'",
           "toupper(AETERM) == 'NAUSEA'",
           "toupper(AETERM) contains 'NAUSEA'"),
  subjects = c(
    ae |> filter(AETERM == "Nausea")                    |> distinct(SUBJID) |> nrow(),
    ae |> filter(toupper(AETERM) == "NAUSEA")           |> distinct(SUBJID) |> nrow(),
    ae |> filter(str_detect(toupper(AETERM), "NAUSEA")) |> distinct(SUBJID) |> nrow()
  )
)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. You should get **61**, **65** and **66**.

Before reading on: which of those three numbers would you put in a
safety table? Notice that you cannot answer that question from the
data either.
:::

::: {.callout-note}
### Reading the Output Line by Line

| Rule | Subjects | What it does |
|---|---|---|
| `AETERM == 'Nausea'` | 61 | Exact match. Silently drops the four subjects whose term was typed in a different case. Wrong, and looks fine. |
| `toupper(AETERM) == 'NAUSEA'` | 65 | Case-insensitive exact match. Catches `NAUSEA` and `nausea`. Still misses `nausea and vomitting`. |
| `str_detect(toupper(AETERM), 'NAUSEA')` | 66 | Substring match. Catches all four spellings — but it counts `nausea and vomitting` as nausea, which is *arguably* right, while quietly failing to also count it as vomiting, which is *definitely* wrong. |

Three rules, all defensible, all written by a competent R programmer in
good faith. Three different numbers in a safety table.
:::

None of these is a bug. Every one of them runs, returns a plausible
integer, and produces no warning. That is precisely what makes this the
dangerous failure mode: **a naming problem throws an error, a
vocabulary problem returns a number.**

The standard's answer has two halves, and this course covers both:

- **Controlled terminology** — the column is not free text. It may hold
  only values from a published list, so `Nausea`, `NAUSEA` and `nausea`
  cannot coexist, because two of them are not legal values. You will
  check values against the published lists programmatically rather than
  from memory in [SDTM Concepts](sdtm-concepts.qmd).
- **Coding** — a human (or a human-supervised process) maps each
  verbatim term a site typed to a dictionary term, and *both* are kept:
  the original words and the coded concept. `Diarrea` and
  `nausea and vomitting` are resolved by that step, not by code.

Notice that the raw verbatim text is never destroyed. That is a
principle worth carrying through the whole course: standardisation adds
an interpretation alongside the original. It does not overwrite it.

## What Can Go Wrong

*Estimated time: ~10 minutes (reading)*

**Treating standardisation as renaming**

The most common misreading of this session is "so I should call the
column `USUBJID` instead of `subjid`." Renaming is the easiest third of
the job and the least important. Example 1 showed structure differing
(one row per event vs per subject per visit per test) and Example 2
showed vocabulary differing. `rename()` addresses neither. If your
mental model of SDTM is a lookup table of old name → new name, Part 2
will not make sense.

**"We'll standardise at the end"**

Standardising is a *decision-recording* activity, not a formatting one.
Look again at subject `111-003`: resolving that conflict requires
someone to find out which value is right. That is easy while the site
is still open and the data manager still remembers the query. It is
expensive a year later and impossible after the study closes. Every
deferred standardisation decision gets more costly, and some of them
become unanswerable. The dates in `dm_raw.csv` are the same story:
someone knew, at the time, why a subject's birth date was recorded as
`1969`.

**Assuming the vendor's file is the trustworthy one**

The lab file looks more authoritative — machine-generated, consistent
formatting, no free text. It is still the file with the wrong sex for
subject `111-003` (or the right one; you do not know yet). Tidiness is
not accuracy.

**Believing a validator's silence**

Conformance tools check that data conforms to the standard. They cannot
check that it is true. Every one of the three nausea counts would pass
a conformance check on the resulting dataset, because each produces a
structurally valid number. Passing validation means you have not made a
*standards* error; it says nothing about whether you have made a
*scientific* one.

::: {.callout-warning}
### Common misinterpretations

**"Standards are bureaucratic overhead imposed by regulators"**
The regulation exists because the problem does, not the other way
round. Everything demonstrated in this session — the disagreeing sex,
the duplicated subject, the 25 spellings — would still be a problem for
your own team six months from now if no agency existed.

**"My data is clean, so this doesn't apply to me"**
Nothing in the GLPX-001 raw data is dirty in the ordinary sense. There
are no impossible values, no corrupted rows, no encoding errors. Every
problem in this session came from *clean data with no agreed meaning*.

**"The CRO/vendor handles CDISC, so I don't need to know it"**
Someone still has to decide what sex subject `111-003` is, and that
someone needs to understand both the trial and the standard. Delegating
the mapping does not delegate the decisions — it just means you approve
decisions you did not make.

**"Once the data conforms, the analysis is objective"**
The three nausea counts differ by rule choice, not by data quality. The
standard removes the *arbitrary* differences (naming, case, structure)
so that the remaining differences are visible and arguable. It does not
remove judgement; it exposes it.
:::

## Exercises

This session has no exercise set. It makes an argument rather than
teaching a technique, and the argument is best tested by the
[Comprehension Check](#comprehension-check) below.

The first exercises arrive with [The DM Domain](sdtm-dm.qmd), by which
point you will have something to build. If you want to do something
now, re-run Example 2's three nausea rules against `Diarrhoea` instead,
and see how many different answers you can defend.

## Comprehension Check

*Estimated time: ~10 minutes*

1. In your own words, why can't a thorough data dictionary for each
   trial replace a standard?
2. `dm_raw.csv` has 401 rows and 400 subjects. Why is this not a data
   error, and what does the standard contribute?
3. Three defensible R rules gave 61, 65 and 66 subjects with nausea.
   Which is correct, and what does your answer tell you about where the
   fix has to happen?
4. A colleague says: "The lab file is machine-generated, so where the
   EDC and the lab disagree, trust the lab." What is wrong with this
   reasoning?
5. Name the three distinct things a standard fixes, and give an example
   of each from the GLPX-001 raw data.

::: {.callout-note collapse="true"}
### Answers

1. A dictionary solves comprehension, not automation. A reviewer can
   read your dictionary and understand your data, but they still cannot
   run an existing program against it, because your columns have your
   names. Documentation scales linearly with the number of trials (read
   each one); a standard is written against once and applies to all of
   them. Every validation rule and review tool in the industry depends
   on that second property.

2. Subject `101-012` was screened at one site and enrolled again at
   another after moving, producing two genuine, correctly recorded
   rows. The raw data has no rule saying one row per subject, so
   nothing prevented it. The standard contributes the rule itself —
   stated in advance, applying to every trial, and checkable by a
   machine rather than by someone noticing.

3. None of them is correct as a *code* problem, which is the point. 61
   is wrong (it drops case variants). 65 and 66 differ only in how they
   treat `nausea and vomitting`, and neither handles it properly,
   because that record describes two events and no filter can split it.
   The fix cannot happen in the analysis; it has to happen upstream, by
   constraining the column to a controlled vocabulary and coding the
   verbatim terms. If your rule choice changes the answer, the answer
   is not in the data yet.

4. It confuses consistency with accuracy. The lab file is
   machine-generated in its *formatting*, but the sex value in it
   originated from a human at some point, exactly like the EDC value.
   Neither file is more likely to be right *a priori*, and the
   disagreement is evidence that at least one human-entered value is
   wrong. Resolving it requires going back to the source, not picking
   the tidier file. (Note also that the lab file is where the duplicate
   record and the missing unit live.)

5. **Variable naming**: `SUBJID` in four files and `subjid` in the lab
   file — the same concept, two names, so no join can be written
   generically. **Dataset structure**: `ae_raw.csv` has one row per
   event (389 rows, fewer than the 400 subjects), while `lb_raw.csv`
   has one row per subject per visit per test (12,389 rows) — "one row"
   means something different in each. **Controlled vocabulary**:
   `AETERM` holds 25 distinct strings for 10 clinical concepts,
   including `Diarrea` and `nausea and vomitting`.
:::
